### 2022-04-29 EDA_Chang_Noordermeer_2021.Rmd
first commit

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(GenomicRanges)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
fileNameIn1 <- file.path(dir_data, "data/Chang_Noordermeer_2021.xlsx")
# Output files
fileNameOut1 <- file.path(dir_data, "")
```

# Load data

```{r data}
# Read in data
mtx <- read_xlsx(fileNameIn1, skip = 1)
```

## EDA

```{r}
summary(mtx$`-log10  p-value...7`)
summary(mtx$`-log10  q-value...8`)
summary(mtx$`-log10  p-value...13`)
summary(mtx$`-log10  q-value...14`)
table(mtx$`overlap with repeat(s)`)
sum
```

# GRanges

```{r}
gr <- GRanges(seqnames = mtx$coordinate, IRanges(start = mtx$...3, end = mtx$...4), strand = mtx$strand)
gr$ID <- mtx$number
gr$length <- mtx$length
gr$value <- mtx$value
gr$peak.log10pval <- mtx$`-log10  p-value...7`
gr$peak.log10qval <- mtx$`-log10  q-value...8`
gr$sequence <-mtx$sequence
gr$CTCF.log10pval <- mtx$`-log10  p-value...13`
gr$CTCF.log10qval <- mtx$`-log10  q-value...14`
gr$repeats <- mtx$`overlap with repeat(s)`

genome_id <- "mm10"
# Get chromosome info and match it to the chromosome order in ctcfBED
chrom_data <- GenomeInfoDb::getChromInfoFromUCSC(genome = genome_id)
chrom_data <- chrom_data[chrom_data$chrom %in% seqlevels(gr), ]
chrom_data <- chrom_data[match(seqlevels(gr), chrom_data$chrom), ]
# Check if chromosome order is the same
if (!all.equal(seqlevels(gr), chrom_data$chrom)) {
  print(paste("Chromosome order does not match for", genome_id, "genome."))
  break
}
# Assign seqinfo data
seqlengths(gr) <- chrom_data$size
isCircular(gr) <- chrom_data$circular
genome(gr)     <- genome_id

gr

# Save as RData object. subdir is the character name of the CTCF GRanges variable
fileNameOut <- "CTCF_mm10_Chang_Noordermeer_2021.RData"
save(list = gr, file = fileNameOut)
# load(file = paste0(subdir, ".RData"))

```

"ID" - numerical ID of a region with multiple (overlapping) CTCF ChIP-seq peaks. "length" - length of the peak. "value" - strength of the peak. "peak.log10pval", "peak.log10qval" - significance of the peak. "sequence" - sequence of the CTCF motif. "CTCF.log10pval", "CTCF.log10qval" - significance of the motif. "repeats" - whether the motif overlaps a repeat.

### 2022-04-29 manuscript/manuscript.Rmd
first commit

---
# csl: /Volumes/GoogleDrive/My Drive/manuscript_template/styles.ref/cell-reports.csl
csl: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.ref/cell-reports.csl
output: 
  word_document:
#    reference_docx: /Volumes/GoogleDrive/My Drive/manuscript_template/styles.doc/NIH_grant_style.docx
    reference_docx: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.doc/NIH_grant_style.docx
# https://github.com/mdozmorov/manuscript_template/tree/master/references
# bibliography: /Volumes/GoogleDrive/My Drive/manuscript_template/references/references_MD.bib
bibliography: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/references/references_MD.bib
---

Authors

# Abstract

### Keywords

# Introduction

# Methods

# Results

Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif.

How to insert single [@Dozmorov:2016aa][@Glass:2016aa] or multiple references [@Dozmorov:2016aa; @Glass:2016aa].

+-------+
| ![](figures/chicken.png)      |
| **Figure 1. How to insert a figure.** Chicken chicken chicken.     |
+-------+
# Discussion

# Abbreviations

# Acknowledgements

_Conflict of Interest._ None.

# Funding

# Tables

# Figure legends

# Supplementary material

# References

### 2022-04-30 EDA_CTCFBSDB.Rmd
Add CTCFBSDB processing

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(GenomicRanges)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
# Predicted data
fileNameIn1 <- file.path(dir_data, "data/allcomp.txt.gz")
# download.file("https://insulatordb.uthsc.edu/download/allcomp.txt.gz", fileNameIn1)
# Experimental data
fileNameIn2 <- file.path(dir_data, "data/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz")
# download.file("https://insulatordb.uthsc.edu/download/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz", fileNameIn2)
```

# Load PREDICTED data

```{r data}
# Read in data
mtx <- read_tsv(fileNameIn1)
```

## EDA

```{r}
table(mtx$Species)
mtx_human <- mtx[mtx$Species == "Human", ]
mtx_mouse <- mtx[mtx$Species == "Mouse", ]
```

## GRanges

```{r}
mtx_to_gr <- function(mtx = mtx_human, genome_id = "hg19") {
  gr <- GRanges(seqnames = sapply(mtx$`Chromosome Location`, function(x) strsplit(x, ":|-")[[1]][1]), 
              IRanges(start = sapply(mtx$`Chromosome Location`, function(x) strsplit(x, ":|-")[[1]][2]) %>% as.numeric(), 
                      end = sapply(mtx$`Chromosome Location`, function(x) strsplit(x, ":|-")[[1]][3]) %>% as.numeric()))

  gr$`5PrimeGene` <- mtx$`5' Flanking Gene`
  gr$`3PrimeGene` <- mtx$`3' Flanking Gene`

  # Get chromosome info and match it to the chromosome order in ctcfBED
  chrom_data <- GenomeInfoDb::getChromInfoFromUCSC(genome = genome_id)
  # Subset to common chromosomes
  common_chromosomes <- intersect(chrom_data$chrom, seqlevels(gr))
  chrom_data <- chrom_data[chrom_data$chrom %in% common_chromosomes, ]
  gr <- keepSeqlevels(gr, common_chromosomes, pruning.mode = "tidy")
  # Match order
  chrom_data <- chrom_data[match(seqlevels(gr), chrom_data$chrom), ]
  # Check if chromosome order is the same
  if (!all.equal(seqlevels(gr), chrom_data$chrom)) {
    print(paste("Chromosome order does not match for", genome_id, "genome."))
    break
  }
  # Assign seqinfo data
  seqlengths(gr) <- chrom_data$size
  isCircular(gr) <- chrom_data$circular
  genome(gr)     <- genome_id

  return(gr)
}

# Save as RData object. subdir is the character name of the CTCF GRanges variable
save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm9"), file = fileNameOut2)
# load(file = paste0(subdir, ".RData"))
```

# Load EXPERIMENTAL data

```{r data}
# Read in data
mtx <- read_tsv(fileNameIn2)
```

## EDA

```{r}
# head(mtx, n = 100) %>% View()
colnames(mtx)
table(mtx$Species)
table(mtx$Build)
mtx$Build[mtx$Build == "galgal3"] <- "galGal3"
table(mtx$ValidationStatus)
table(mtx$CellType)
table(mtx$Occupancy)
summary(mtx$CTCF_Rank %>% as.numeric())
summary(mtx$Readcount %>% as.numeric())
table(mtx$ENCODE)
summary(mtx$ENCODE_Peak)
table(mtx$M1M2_class)
```

## GRanges

```{r}
mtx_to_gr <- function(mtx = mtx, org = "Rhesus") {
  mtx_selected <- mtx[mtx$Species == org, ]
  gr <- GRanges(seqnames = mtx_selected$Chr, IRanges(start = mtx_selected$GenomeBeginLoc, end = mtx_selected$GenomeEndLoc))

  gr$Reference        <- mtx_selected$Reference
  gr$ValidationStatus <- mtx_selected$ValidationStatus
  gr$UpstreamGene     <- mtx_selected$UpstreamGene
  gr$DownstreamGene   <- mtx_selected$DownstreamGene
  gr$CellType         <- mtx_selected$CellType
  gr$Occupancy        <- mtx_selected$Occupancy
  gr$CTCF_Rank        <- mtx_selected$CTCF_Rank
  gr$Readcount        <- mtx_selected$Readcount
  gr$ENCODE           <- mtx_selected$ENCODE
  gr$ENCODE_Peak      <- mtx_selected$ENCODE_Peak
  gr$M1M2_class       <- mtx_selected$M1M2_class

  genome_id <- unique(mtx_selected$Build)
  # Get chromosome info and match it to the chromosome order in ctcfBED
  chrom_data <- GenomeInfoDb::getChromInfoFromUCSC(genome = genome_id)
  # Subset to common chromosomes
  common_chromosomes <- intersect(chrom_data$chrom, seqlevels(gr))
  chrom_data <- chrom_data[chrom_data$chrom %in% common_chromosomes, ]
  gr <- keepSeqlevels(gr, common_chromosomes, pruning.mode = "tidy")
  # Match order
  chrom_data <- chrom_data[match(seqlevels(gr), chrom_data$chrom), ]
  # Check if chromosome order is the same
  if (!all.equal(seqlevels(gr), chrom_data$chrom)) {
    print(paste("Chromosome order does not match for", genome_id, "genome."))
    break
  }
  # Assign seqinfo data
  seqlengths(gr) <- chrom_data$size
  isCircular(gr) <- chrom_data$circular
  genome(gr)     <- genome_id

  return(gr)
}

organisms <- unique(mtx$Species)
organisms <- organisms[organisms != c("Rhesus")]
for(i in organisms) {
  print(i)
  gr <- mtx_to_gr(mtx, org = i)
  # Output files
  fileNameOut1 <- file.path(dir_data, paste0("results/CTCF_", i, "_CTCFBSDB_experimental.RData"))
  # Save as RData object. 
  # save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
}

# load(file = paste0(subdir, ".RData"))
```

### 2022-04-30 EDA_Chang_Noordermeer_2021.Rmd
Add CTCFBSDB processing

@@
 fileNameIn1 <- file.path(dir_data, "data/Chang_Noordermeer_2021.xlsx")
+# download.file("https://www.biorxiv.org/content/biorxiv/early/2021/04/15/2021.04.15.440007/DC1/embed/media-1.xlsx", fileNameIn1)
+
 # Output files
-fileNameOut1 <- file.path(dir_data, "")
+fileNameOut1 <- "results/CTCF_mm10_Chang_Noordermeer_2021.RData"
 ```
@@
 # Load data
+
+"ID" - numerical ID of a region with multiple (overlapping) CTCF ChIP-seq peaks.
+"length" - length of the peak.
+"value" - strength of the peak.
+"peak.log10pval", "peak.log10qval" - significance of the peak.
+"sequence" - sequence of the CTCF motif.
+"CTCF.log10pval", "CTCF.log10qval" - significance of the motif.
+"repeats" - whether the motif overlaps a repeat.
 
@@
 table(mtx$`overlap with repeat(s)`)
-sum
 ```
@@
 genome(gr)     <- genome_id
-
 gr
 
-# Save as RData object. subdir is the character name of the CTCF GRanges variable
-fileNameOut <- "CTCF_mm10_Chang_Noordermeer_2021.RData"
-save(list = gr, file = fileNameOut)
+# Save as RData object.
+save(list = gr, file = fileNameOut1)
 # load(file = paste0(subdir, ".RData"))
-
 ```
-
-"ID" - numerical ID of a region with multiple (overlapping) CTCF ChIP-seq peaks.
-"length" - length of the peak.
-"value" - strength of the peak.
-"peak.log10pval", "peak.log10qval" - significance of the peak.
-"sequence" - sequence of the CTCF motif.
-"CTCF.log10pval", "CTCF.log10qval" - significance of the motif.
-"repeats" - whether the motif overlaps a repeat.

### 2022-04-30 manuscript/manuscript.Rmd
Add CTCFBSDB processing

@@
 
+CTCF (CCCTC-binding factor) is a versatile transctiprion regulators that is evolutionary conserved from fruit fly to human.
+CTCF binds to different DNA sequences through combinatorial use of 11-zinc fingers, and shows distingc functions (transcription activation/repression and chromatin insulation).
+
+Topologically Associating Domains (TADs) are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa].
+CTCF binding, especially in convergent orientation, has long been known to be critical for loop formation [@Dixon:2012aa].
+
+ CTCF binding site database (CTCFBSDB), a comprehensive collection of experimentally determined and computationally predicted CTCF binding sites (CTCFBS) from the literature.
+
 # Methods
@@
 Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif.
+
+CTCFBSDB assembled predictred and experimentally detected CTCF binding information. Human predictions containes 13,356 sites (hg19) and mouse predictions contained 5,504 sites (mm9).
 

### 2022-04-30 manuscript/manuscript.Rmd
More writeup

--- 2022-04-29 manuscript/manuscript.Rmd
@@
 
+CTCF (CCCTC-binding factor) is a versatile transctiprion regulators that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu].
+<!--It was originally defined as a transcriptional repressor of the MYC oncogene [@Lobanenkov:1990vm][@Filippova:1996tv].--> Originally defined as a repressor/insulator [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
+
+CTCF binds to different DNA sequences through combinatorial use of 11-zinc fingers.
+While many functional CTCF-binding sites do not match a consensus motif [@Ohlsson:2010uy], a multi-part CTCF-binding motif has been defined [@Schmidt:2012uk].
+For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif.
+Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif.
+
+Topologically Associating Domains (TADs) are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa].
+CTCF binding, especially in convergent orientation, has long been known to be critical for loop formation [@Dixon:2012aa].
+
+The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj].
+Unfortunately, the database has not been updated since 2013 and its data are limited to old genome assemblies (e.g., hg19, mm9).
+
 # Methods
@@
 Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif.
+
+CTCFBSDB assembled predictred and experimentally detected CTCF binding information [@Bao:2008vq]. Human predictions containes 13,356 sites (hg19) and mouse predictions contained 5,504 sites (mm9).
 

### 2022-05-01 manuscript/manuscript.Rmd
Add Encode Screen data

@@
 
+SCREEN database provides CTCF-bound cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies.
+The human set of regions contains 450,641 records.
+They were classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%).
+The mouse set contained 82,777 records similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites.
+
 How to insert single [@Dozmorov:2016aa][@Glass:2016aa] or multiple references [@Dozmorov:2016aa; @Glass:2016aa].

### 2022-05-02 EDA_SCREEN.Rmd
More writeup about SCREEN

--- 2022-04-30 EDA_CTCFBSDB.Rmd
@@
 # Input files
-# Predicted data
-fileNameIn1 <- file.path(dir_data, "data/allcomp.txt.gz")
-# download.file("https://insulatordb.uthsc.edu/download/allcomp.txt.gz", fileNameIn1)
-# Experimental data
-fileNameIn2 <- file.path(dir_data, "data/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz")
-# download.file("https://insulatordb.uthsc.edu/download/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz", fileNameIn2)
+# Human CTCF-bound cCRE (hg38)
+fileNameIn1 <- file.path(dir_data, "data/GRCh38-CTCF.bed")
+# download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/GRCh38-CTCF.bed", fileNameIn1)
+# Mouse CTCF-bound cCRE (mm10)
+fileNameIn2 <- file.path(dir_data, "data/mm10-CTCF.bed")
+# download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/mm10-CTCF.bed", fileNameIn2)
+# Output files
+fileNameOut1 <- file.path(dir_data, "results/CTCF_hg38_SCREEN.RData")
+fileNameOut2 <- file.path(dir_data, "results/CTCF_mm10_SCREEN.RData")
 ```
 
-# Load PREDICTED data
+PLS/pELS/dELS - Promoter-Like Signatures, proximal/distal Enhancer-Like Signatures
+
+# Load hg38 data
 
@@
 # Read in data
-mtx <- read_tsv(fileNameIn1)
+mtx_human <- read_tsv(fileNameIn1, col_names = FALSE)
+mtx_mouse <- read_tsv(fileNameIn2, col_names = FALSE)
 ```
@@
 ```{r}
-table(mtx$Species)
-mtx_human <- mtx[mtx$Species == "Human", ]
-mtx_mouse <- mtx[mtx$Species == "Mouse", ]
+table(mtx_human$X6) / length(mtx_human$X6)
+table(mtx_mouse$X6) / length(mtx_mouse$X6)
 ```
@@
 ```{r}
-mtx_to_gr <- function(mtx = mtx_human, genome_id = "hg19") {
-  gr <- GRanges(seqnames = sapply(mtx$`Chromosome Location`, function(x) strsplit(x, ":|-")[[1]][1]), 
-              IRanges(start = sapply(mtx$`Chromosome Location`, function(x) strsplit(x, ":|-")[[1]][2]) %>% as.numeric(), 
-                      end = sapply(mtx$`Chromosome Location`, function(x) strsplit(x, ":|-")[[1]][3]) %>% as.numeric()))
+mtx_to_gr <- function(mtx = mtx_human, genome_id = "hg38") {
+  gr <- GRanges(seqnames = mtx$X1, 
+              IRanges(start = mtx$X2, 
+                      end = mtx$X3))
 
-  gr$`5PrimeGene` <- mtx$`5' Flanking Gene`
-  gr$`3PrimeGene` <- mtx$`3' Flanking Gene`
+  gr$ID1 <- mtx$X4
+  gr$ID2 <- mtx$X5
+  gr$Type <- mtx$X6
 
@@
 # Save as RData object. subdir is the character name of the CTCF GRanges variable
-save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
-save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm9"), file = fileNameOut2)
+save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg38"), file = fileNameOut1)
+save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10"), file = fileNameOut2)
 # load(file = paste0(subdir, ".RData"))
 ```
-
-# Load EXPERIMENTAL data
-
-```{r data}
-# Read in data
-mtx <- read_tsv(fileNameIn2)
-```
-
-## EDA
-
-```{r}
-# head(mtx, n = 100) %>% View()
-colnames(mtx)
-table(mtx$Species)
-table(mtx$Build)
-mtx$Build[mtx$Build == "galgal3"] <- "galGal3"
-table(mtx$ValidationStatus)
-table(mtx$CellType)
-table(mtx$Occupancy)
-summary(mtx$CTCF_Rank %>% as.numeric())
-summary(mtx$Readcount %>% as.numeric())
-table(mtx$ENCODE)
-summary(mtx$ENCODE_Peak)
-table(mtx$M1M2_class)
-```
-
-## GRanges
-
-```{r}
-mtx_to_gr <- function(mtx = mtx, org = "Rhesus") {
-  mtx_selected <- mtx[mtx$Species == org, ]
-  gr <- GRanges(seqnames = mtx_selected$Chr, IRanges(start = mtx_selected$GenomeBeginLoc, end = mtx_selected$GenomeEndLoc))
-
-  gr$Reference        <- mtx_selected$Reference
-  gr$ValidationStatus <- mtx_selected$ValidationStatus
-  gr$UpstreamGene     <- mtx_selected$UpstreamGene
-  gr$DownstreamGene   <- mtx_selected$DownstreamGene
-  gr$CellType         <- mtx_selected$CellType
-  gr$Occupancy        <- mtx_selected$Occupancy
-  gr$CTCF_Rank        <- mtx_selected$CTCF_Rank
-  gr$Readcount        <- mtx_selected$Readcount
-  gr$ENCODE           <- mtx_selected$ENCODE
-  gr$ENCODE_Peak      <- mtx_selected$ENCODE_Peak
-  gr$M1M2_class       <- mtx_selected$M1M2_class
-
-  genome_id <- unique(mtx_selected$Build)
-  # Get chromosome info and match it to the chromosome order in ctcfBED
-  chrom_data <- GenomeInfoDb::getChromInfoFromUCSC(genome = genome_id)
-  # Subset to common chromosomes
-  common_chromosomes <- intersect(chrom_data$chrom, seqlevels(gr))
-  chrom_data <- chrom_data[chrom_data$chrom %in% common_chromosomes, ]
-  gr <- keepSeqlevels(gr, common_chromosomes, pruning.mode = "tidy")
-  # Match order
-  chrom_data <- chrom_data[match(seqlevels(gr), chrom_data$chrom), ]
-  # Check if chromosome order is the same
-  if (!all.equal(seqlevels(gr), chrom_data$chrom)) {
-    print(paste("Chromosome order does not match for", genome_id, "genome."))
-    break
-  }
-  # Assign seqinfo data
-  seqlengths(gr) <- chrom_data$size
-  isCircular(gr) <- chrom_data$circular
-  genome(gr)     <- genome_id
-
-  return(gr)
-}
-
-organisms <- unique(mtx$Species)
-organisms <- organisms[organisms != c("Rhesus")]
-for(i in organisms) {
-  print(i)
-  gr <- mtx_to_gr(mtx, org = i)
-  # Output files
-  fileNameOut1 <- file.path(dir_data, paste0("results/CTCF_", i, "_CTCFBSDB_experimental.RData"))
-  # Save as RData object. 
-  # save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
-}
-
-# load(file = paste0(subdir, ".RData"))
-```

### 2022-05-02 manuscript/manuscript.Rmd
More writeup about SCREEN

@@
 
-SCREEN database provides CTCF-bound cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies.
-The human set of regions contains 450,641 records.
+The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq].
+The human set of CTCF-bound cCREs contains 450,641 records.
 They were classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%).

### 2022-05-09 manuscript/manuscript.Rmd
More misc stuff

@@
 Topologically Associating Domains (TADs) are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa].
-CTCF binding, especially in convergent orientation, has long been known to be critical for loop formation [@Dixon:2012aa].
+CTCF binding, especially in convergent orientation, has long been known to be critical for loop formation [@Dixon:2012aa][@Oti:2016aa].
 
@@
 
-Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif.
+Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif (mm10 genome assembly).
 
-CTCFBSDB assembled predictred and experimentally detected CTCF binding information [@Bao:2008vq]. Human predictions containes 13,356 sites (hg19) and mouse predictions contained 5,504 sites (mm9).
+CTCFBSDB assembled predicted and experimentally detected CTCF binding information [@Bao:2008vq].
+Human predictions containes 13,356 sites and mouse predictions contained 5,504 sites (hg19 and mm9 genome assemblies, respectively).
+Unfortunately, the database has not been updated since 2012; therefore, its data is of limited use.
 

### 2022-05-15 manuscript/manuscript.Rmd
More writeup

@@
 
-CTCF (CCCTC-binding factor) is a versatile transctiprion regulators that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu].
-<!--It was originally defined as a transcriptional repressor of the MYC oncogene [@Lobanenkov:1990vm][@Filippova:1996tv].--> Originally defined as a repressor/insulator [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
+The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiprion regulator that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux].
+Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
 
-CTCF binds to different DNA sequences through combinatorial use of 11-zinc fingers.
-While many functional CTCF-binding sites do not match a consensus motif [@Ohlsson:2010uy], a multi-part CTCF-binding motif has been defined [@Schmidt:2012uk].
+Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].
+
+CTCF localization is largely invariant across different cell types [@Kim:2007ux].
+The goal of our package is to provide such lists of genome-wide CTCF binding sites generally applicable across cell types.
+
+CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to $\sim50bp$ target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Ohlsson:2001wu].
+A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1.
+It was subsequently refined to include an additional 9bp motif (M2) [@Schmidt:2012uk].
 For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif.
 Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif.
+The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al.
+[@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding.
 
-Topologically Associating Domains (TADs) are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa].
-CTCF binding, especially in convergent orientation, has long been known to be critical for loop formation [@Dixon:2012aa][@Oti:2016aa].
+Position Weight Matrices (PWMs) [@Stormo:2000uw] for CTCF are available from multiple databases.
+JASPAR 2022 [@Castro-Mondragon:2022uv] and HOCOMOCO [@Kulakovskiy:2018vw] databases list one 19bp-long CTCF PWM ([MA0139.1](https://jaspar.genereg.net/matrix/MA0139.1/) and [CTCF_HUMAN.H11MO.0.A](https://hocomoco11.autosome.org/motif/CTCF_HUMAN.H11MO.0.A), respectively).
+Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific [CTCF](http://cisbp.ccbr.utoronto.ca/TFreport.php?searchTF=T094831_2.00) binding motifs, likely reflecting the variability of CTCF motifs without considering the M1+M2 combinatorial structure.
+The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al.
+[@Schmidt:2012uk], the Ren_20 motif identified by Kim et al.
+[@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al.
+[@Xie:2007wm].
 
-The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj].
 Unfortunately, the database has not been updated since 2013 and its data are limited to old genome assemblies (e.g., hg19, mm9).
+
+CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa].
+Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Rao:2014aa][@Vietri-Rudan:2015tu], and the enrichment of CTCF and cohesin is observed at TAD boundaries  [@Beagan:2020wz][@Stilianoudakis:2020aa].
+To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops
+
+We utilized the PWMscan web tool [@Ambrosini:2018wv] because it has a large collection of PWMs (JASPAR, HOCOMOCO, SwissRegulon, UniPROBE, CIS-BP, data from Jomla, Isakova publications) and genomes (49 genome assemblies); its results show over 99% agreement with FIMO, HOMER, and other motif scanning tools.
 
@@
 
-Chang et al. provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl]. They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif (mm10 genome assembly).
+Chang et al.
+provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl].
+They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif (mm10 genome assembly).
+It should be noted that mouse embryonic stem cells (ESCs) have thousands of CTCF-binding events that cannot be conserved in the human genome, as they are found in rodent- specific B2 repeat elements [@Bourque:2008ws].
 

### 2022-05-15 EDA_PWMScan.Rmd
EDA_PWMScan

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(rtracklayer)
library(stringr)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
library(plyranges)
library(patchwork)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev/data/PWMscan/"
# Genome assembly
genome_assembly <- "hg38"

# Input files
fileNameIn1 <- file.path(dir_data, "")
# Output files
fileNameOut1 <- file.path(dir_data, "")
```

# Load data

```{r data}
# Read in data
files <- list.files(path = dir_data, pattern = genome_assembly, full.names = TRUE)
# Object to store GRanges 
bed_list_all <- list() # All chromosomes
bed_list_standard <- list() # Standard chromosomes
# Additional columns
extraCols_names <- c("character", "character")
names(extraCols_names) <- c("motif", "pval")
# Process each file
for(i in 1:length(files)) {
  print(basename(files[i]))
  # Read data
  bed_file <- import(files[i], format = "bed", genome = genome_assembly, colnames = c("chrom", "start", "end", "name", "score", "strand", "motif", "pval"), extraCols = extraCols_names)
  # Save all data
  bed_list_all <- c(bed_list_all, list(bed_file))
  # Keep standard chromosomes
  bed_file_standard <- keepStandardChromosomes(bed_file)
  # Save filtered data
  bed_list_standard <- c(bed_list_standard, list(bed_file_standard))
}
# Add names
names(bed_list_all) <- basename(files) %>% str_replace(., ".bed", "") %>% str_replace(., paste0(genome_assembly, ".PWMScan."), "")
names(bed_list_standard) <- basename(files) %>% str_replace(., ".bed", "") %>% str_replace(., paste0(genome_assembly, ".PWMScan."), "")
```

# Summaries

## Genomewide Number

```{r fig.height=3, fig.width=10}
stats_number_pos <- c()
stats_number_neg <- c()
for(i in 1:length(bed_list_standard)) {
  stats_number_pos <- c(stats_number_pos, length(bed_list_standard[[i]][strand(bed_list_standard[[i]]) == "+" ]))
  stats_number_neg <- c(stats_number_neg, length(bed_list_standard[[i]][strand(bed_list_standard[[i]]) == "-" ]))
}
mtx_to_plot <- data.frame(Number = c(stats_number_pos, stats_number_neg), 
                          List = c(names(bed_list_standard), names(bed_list_standard)), 
                          Strand = c(rep("+", length(stats_number_pos)), rep("-", length(stats_number_neg))))

stats_number_total <- aggregate(mtx_to_plot$Number, list(mtx_to_plot$List), sum)
colnames(stats_number_total) <- c("List", "Total")
mtx_to_plot$List <- factor(mtx_to_plot$List, levels = stats_number_total$List[order(stats_number_total$Total)])
ggplot(mtx_to_plot, aes(x = List, y = Number, fill = Strand)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw(base_size = 15) +
  # get rid of the grid 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        text = element_text(size = 15),
        legend.position = c(0.8, 0.30)) +
  scale_fill_manual(values = mycols[1:2]) + # change colors 
  xlab("Database") +
  ylab("Number of motifs") +
  ggtitle(paste(genome_assembly, "genome assembly"))
```

## Chromosome-specific Number

```{r fig.height=16, fig.width=10}
bed_to_number_per_chr_plot <- function(i = 1) {
  bed_list_standard_selected = bed_list_standard[[i]]
  # Summarize counts per chromosome and strand
  mtx_to_plot <- bed_list_standard_selected %>% group_by(seqnames, strand) %>% summarise(Number = n()) %>% as.data.frame()
  # Reorder chromosomes
  mtx_to_plot$seqnames <- factor(mtx_to_plot$seqnames, levels = (unique(mtx_to_plot$seqnames)))
  p <- ggplot(mtx_to_plot, aes(x = seqnames, y = Number, fill = strand)) +
    geom_bar(stat = "identity") +
    # coord_flip() +
    theme_bw(base_size = 15) +
    # get rid of the grid 
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(), 
          text = element_text(size = 15),
          legend.position = c(0.8, 0.80)) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) +
    scale_fill_manual(values = mycols[1:2]) + # change colors 
    xlab("Chromosome") +
    ylab("Number of motifs") +
    ggtitle(paste(genome_assembly, names(bed_list_standard)[i]))
  return(p)
}

p_list <- list()
for(i in 1: length(bed_list_standard)) {
  print(names(bed_list_standard)[i])
  p_list <- c(p_list, list(bed_to_number_per_chr_plot(i = i)))
}
wrap_plots(p_list, nrow = length(p_list))
```

### 2022-05-15 EDA_PWMScan.Rmd
mm10 PWMScan

@@
 library(plyranges)
+library(pheatmap)
 library(patchwork)
@@
 # Genome assembly
-genome_assembly <- "hg38"
+# genome_assembly <- "hg38"
+genome_assembly <- "mm10"
 
@@
 ```
+
+## Jaccard
+
+```{r jaccard, warning=FALSE, fig.width=6.5, fig.height=6}
+save_pheatmap_svg <- function(x, filename, width=4.5, height=3, units = "in", res = 300) {
+  stopifnot(!missing(x))
+  stopifnot(!missing(filename))
+  svg(filename, width=width, height=height)
+  grid::grid.newpage()
+  grid::grid.draw(x$gtable)
+  dev.off()
+}
+# Jaccard calculations
+jaccard <- function(gr_a, gr_b) {
+  intersects <- GenomicRanges::intersect(gr_a, gr_b, ignore.strand = TRUE)
+  intersection <- sum(width(intersects))
+  union <- sum(width(GenomicRanges::union(gr_a, gr_b, ignore.strand = TRUE)))
+  DataFrame(intersection, union, 
+            jaccard = intersection/union,
+             n_intersections = length(intersects))
+}
+# Correlation matrix, empty
+mtx_to_plot <- matrix(data = 0, nrow = length(bed_list_standard), ncol = length(bed_list_standard))
+# Fill it in
+for (i in 1:length(bed_list_standard)) {
+  for (j in 1:length(bed_list_standard)) {
+    # If diagonal, set to zero
+    if (i == j) mtx_to_plot[i, j] <- 0
+    # Process only one half, the other is symmetric
+    if (i > j) {
+      mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- jaccard(bed_list_standard[[i]], bed_list_standard[[j]])[["jaccard"]]
+    }
+  }
+}
+# Trim row/colnames
+rownames(mtx_to_plot) <- colnames(mtx_to_plot) <- str_trunc(names(bed_list_standard), width = 25) 
+# Save the plot
+# png("man/figures/excluderanges_hg38_jaccard.png", width = 1000, height = 900, res = 200)
+if(genome_assembly == "hg38") {
+  clust_method <- "euclidean"
+}
+if(genome_assembly == "mm10") {
+  clust_method <- "correlation"
+}
+pheatmap(data.matrix(mtx_to_plot), cluster_cols = T, cluster_rows = T,
+         clustering_method = "ward.D", 
+         clustering_distance_rows = clust_method, 
+         clustering_distance_cols = clust_method, 
+         # annotation_row = mydf, annotation_colors = list(Group = c(PR = mycols[1], CR = mycols[2])),
+         treeheight_row = 40,
+         treeheight_col = 0, 
+         display_numbers = TRUE)
+# dev.off()
+```

### 2022-05-17 manuscript/manuscript.Rmd
More writeup

@@
 Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
+Point mutation and loss of heterozygosity of CTCF is associated with human cancer, identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].
 
@@
 
-CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to $\sim50bp$ target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Ohlsson:2001wu].
+CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Filippova:1996tv][@Ohlsson:2001wu].
 A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1.
@@
 [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding.
+Studies have reported 20,000-50,000 CTCF binding sites within the human genome [@Kim:2007ux][@Nakahashi:2013uw].
 
@@
 CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa].
-Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Rao:2014aa][@Vietri-Rudan:2015tu], and the enrichment of CTCF and cohesin is observed at TAD boundaries  [@Beagan:2020wz][@Stilianoudakis:2020aa].
+Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa].
 To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops
+
+The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr].
+Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains.
+BORIS became progressively specialized to testis-specific expression throughout amniote evolution.
+CTCFL is normally expressed only in germ cells but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv].
+CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv].
+CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].
 

### 2022-05-17 EDA_PWMScan.Rmd
Playing with GenometriCorr

--- 2022-05-15 EDA_PWMScan.Rmd
@@
 library(plyranges)
+library(pheatmap)
 library(patchwork)
+library(GenometriCorr)
 ```
@@
 ```
+
+## Jaccard
+
+```{r jaccard, warning=FALSE, fig.width=6.5, fig.height=6}
+save_pheatmap_svg <- function(x, filename, width=4.5, height=3, units = "in", res = 300) {
+  stopifnot(!missing(x))
+  stopifnot(!missing(filename))
+  svg(filename, width=width, height=height)
+  grid::grid.newpage()
+  grid::grid.draw(x$gtable)
+  dev.off()
+}
+# Jaccard calculations
+jaccard <- function(gr_a, gr_b) {
+  intersects <- GenomicRanges::intersect(gr_a, gr_b, ignore.strand = TRUE)
+  intersection <- sum(width(intersects))
+  union <- sum(width(GenomicRanges::union(gr_a, gr_b, ignore.strand = TRUE)))
+  DataFrame(intersection, union, 
+            jaccard = intersection/union,
+             n_intersections = length(intersects))
+}
+# Correlation matrix, empty
+mtx_to_plot <- matrix(data = 0, nrow = length(bed_list_standard), ncol = length(bed_list_standard))
+# Fill it in
+for (i in 1:length(bed_list_standard)) {
+  for (j in 1:length(bed_list_standard)) {
+    # If diagonal, set to zero
+    if (i == j) mtx_to_plot[i, j] <- 0
+    # Process only one half, the other is symmetric
+    if (i > j) {
+      # mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- jaccard(bed_list_standard[[i]], bed_list_standard[[j]])[["jaccard"]]
+      mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- GCorr(bed_list_standard[[i]], bed_list_standard[[j]])
+    }
+  }
+}
+# Trim row/colnames
+rownames(mtx_to_plot) <- colnames(mtx_to_plot) <- str_trunc(names(bed_list_standard), width = 25) 
+# Save the plot
+# png("man/figures/excluderanges_hg38_jaccard.png", width = 1000, height = 900, res = 200)
+if(genome_assembly == "hg38") {
+  clust_method <- "euclidean"
+}
+if(genome_assembly == "mm10") {
+  clust_method <- "correlation"
+}
+pheatmap(data.matrix(mtx_to_plot), cluster_cols = T, cluster_rows = T,
+         clustering_method = "ward.D", 
+         clustering_distance_rows = clust_method, 
+         clustering_distance_cols = clust_method, 
+         # annotation_row = mydf, annotation_colors = list(Group = c(PR = mycols[1], CR = mycols[2])),
+         treeheight_row = 40,
+         treeheight_col = 0, 
+         display_numbers = TRUE)
+# dev.off()
+```
+
+## GenometriCorr
+
+```{r}
+# CIS-BP
+gr1 <- bed_list_standard[[1]][seqnames(bed_list_standard[[1]]) == "chr1"]
+gr2 <- bed_list_standard[[5]][seqnames(bed_list_standard[[5]]) == "chr1"]
+
+# GenometriCorr of smaller query vs. larger reference
+# Returns either Jaccard or ECDF area correlation
+GCorr <- function(gr1, gr2, return_value = "relative.distances.ecdf.area.correlation") {
+  # Select smaller query and larger reference
+  if (length(gr1) >= length(gr2)) {
+    reference <- gr1
+    query <- gr2
+  } else {
+    reference <- gr2
+    query <- gr1
+  }
+  # GenometriCorr itselv
+  gr1_vs_gr2 <- GenometriCorrelation(query, reference, awhole.only = TRUE, showProgressBar = FALSE, permut.number = 0)
+  # What to return
+  if (return_value == "relative.distances.ecdf.area.correlation") {
+    return(gr1_vs_gr2$awhole$relative.distances.ecdf.area.correlation)
+  }
+  if (return_value == "jaccard.measure") {
+    return(gr1_vs_gr2$awhole$jaccard.measure)
+  }
+}
+
+```

### 2022-05-25 EDA_SCREEN.Rmd
Adjust SCREEN data processing

@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "results/CTCF_hg38_SCREEN.RData")
-fileNameOut2 <- file.path(dir_data, "results/CTCF_mm10_SCREEN.RData")
+fileNameOut1 <- file.path(dir_data, "results/hg38.SCREEN.GRCh38_CTCF.RData")
+fileNameOut2 <- file.path(dir_data, "results/mm10.SCREEN.mm10_CTCF.RData")
 ```
@@
 
+hg38.SCREEN.GRCh38_CTCF <- mtx_to_gr(mtx = mtx_human, genome_id = "hg38")
+length(hg38.SCREEN.GRCh38_CTCF)
+summary(width(hg38.SCREEN.GRCh38_CTCF))
+mm10.SCREEN.mm10_CTCF <- mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10")
+length(mm10.SCREEN.mm10_CTCF)
+summary(width(mm10.SCREEN.mm10_CTCF))
+
 # Save as RData object. subdir is the character name of the CTCF GRanges variable
-save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg38"), file = fileNameOut1)
-save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10"), file = fileNameOut2)
+save(list = hg38.SCREEN.GRCh38_CTCF, file = fileNameOut1)
+save(list = mm10.SCREEN.mm10_CTCF, file = fileNameOut2)
 # load(file = paste0(subdir, ".RData"))

### 2022-05-25 EDA_SCREEN.Rmd
save to BED

@@
 library(GenomicRanges)
+library(rtracklayer)
 ```
@@
 fileNameOut1 <- file.path(dir_data, "results/hg38.SCREEN.GRCh38_CTCF.RData")
+fileNameOut1.1 <- file.path(dir_data, "results/hg38.SCREEN.GRCh38_CTCF.bed")
 fileNameOut2 <- file.path(dir_data, "results/mm10.SCREEN.mm10_CTCF.RData")
+fileNameOut2.1 <- file.path(dir_data, "results/mm10.SCREEN.mm10_CTCF.bed")
 ```
@@
 # load(file = paste0(subdir, ".RData"))
+# Save BED files
+export.bed(hg38.SCREEN.GRCh38_CTCF, fileNameOut1.1)
+export.bed(mm10.SCREEN.mm10_CTCF, fileNameOut2.1)
 ```

### 2022-05-26 EDA_CTCFBSDB.Rmd
work on EDA_CTCFBSDB.Rmd

@@
 # download.file("https://insulatordb.uthsc.edu/download/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz", fileNameIn2)
+# Output files
+fileNameOut1 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_predicted_human.RData")
+fileNameOut1.1 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_predicted_human.bed")
+fileNameOut2 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_predicted_mouse.RData")
+fileNameOut2.1 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_predicted_mouse.bed")
+# fileNameOut3 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_experimental_human.RData")
+# fileNameOut3.1 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_experimental_human.bed")
+# fileNameOut4 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.RData")
+# fileNameOut4.1 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.bed")
 ```
@@
 
+hg19.CTCFBSDB.CTCF_predicted_human <- mtx_to_gr(mtx = mtx_human, genome_id = "hg38")
+length(hg19.CTCFBSDB.CTCF_predicted_human)
+summary(width(hg19.CTCFBSDB.CTCF_predicted_human))
+mm9.CTCFBSDB.CTCF_predicted_mouse <- mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10")
+length(mm9.CTCFBSDB.CTCF_predicted_mouse)
+summary(width(mm9.CTCFBSDB.CTCF_predicted_mouse))
+
 # Save as RData object. subdir is the character name of the CTCF GRanges variable
-save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
-save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm9"), file = fileNameOut2)
+save(list = hg19.CTCFBSDB.CTCF_predicted_human, file = fileNameOut1)
+save(list = mm9.CTCFBSDB.CTCF_predicted_mouse, file = fileNameOut2)
 # load(file = paste0(subdir, ".RData"))
+# Save BED files
+export.bed(hg19.CTCFBSDB.CTCF_predicted_human, fileNameOut1.1)
+export.bed(mm9.CTCFBSDB.CTCF_predicted_mouse, fileNameOut2.1)
 ```
@@
 
-```{r data}
+```{r eval=FALSE}
 # Read in data
@@
 
-```{r}
+```{r eval=FALSE}
 # head(mtx, n = 100) %>% View()
@@
 
-```{r}
+```{r eval=FALSE}
 mtx_to_gr <- function(mtx = mtx, org = "Rhesus") {
@@
 organisms <- unique(mtx$Species)
-organisms <- organisms[organisms != c("Rhesus")]
-for(i in organisms) {
-  print(i)
-  gr <- mtx_to_gr(mtx, org = i)
-  # Output files
-  fileNameOut1 <- file.path(dir_data, paste0("results/CTCF_", i, "_CTCFBSDB_experimental.RData"))
-  # Save as RData object. 
-  # save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
-}
+# organisms <- organisms[organisms != c("Rhesus")]
+# for(i in organisms) {
+#   print(i)
+#   gr <- mtx_to_gr(mtx, org = i)
+#   # Output files
+#   fileNameOut1 <- file.path(dir_data, paste0("results/CTCF_", i, "_CTCFBSDB_experimental.RData"))
+#   # Save as RData object. 
+#   # save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg19"), file = fileNameOut1)
+# }
 
 # load(file = paste0(subdir, ".RData"))
+
+hg18.CTCFBSDB.CTCF_experimental_human <- mtx_to_gr(mtx = mtx, org = "Human")
+length(hg18.CTCFBSDB.CTCF_experimental_human)
+summary(width(hg18.CTCFBSDB.CTCF_experimental_human))
 ```

### 2022-05-26 EDA_CTCFBSDB.Rmd
Change assemblies to hg18 and mm8

@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_predicted_human.RData")
-fileNameOut1.1 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_predicted_human.bed")
-fileNameOut2 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_predicted_mouse.RData")
-fileNameOut2.1 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_predicted_mouse.bed")
+fileNameOut1 <- file.path(dir_data, "results/hg18.CTCFBSDB.CTCF_predicted_human.RData")
+fileNameOut1.1 <- file.path(dir_data, "results/hg18.CTCFBSDB.CTCF_predicted_human.bed")
+fileNameOut2 <- file.path(dir_data, "results/mm8.CTCFBSDB.CTCF_predicted_mouse.RData")
+fileNameOut2.1 <- file.path(dir_data, "results/mm8.CTCFBSDB.CTCF_predicted_mouse.bed")
 # fileNameOut3 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_experimental_human.RData")
@@
 
-hg19.CTCFBSDB.CTCF_predicted_human <- mtx_to_gr(mtx = mtx_human, genome_id = "hg38")
-length(hg19.CTCFBSDB.CTCF_predicted_human)
-summary(width(hg19.CTCFBSDB.CTCF_predicted_human))
-mm9.CTCFBSDB.CTCF_predicted_mouse <- mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10")
-length(mm9.CTCFBSDB.CTCF_predicted_mouse)
-summary(width(mm9.CTCFBSDB.CTCF_predicted_mouse))
+hg18.CTCFBSDB.CTCF_predicted_human <- mtx_to_gr(mtx = mtx_human, genome_id = "hg18")
+length(hg18.CTCFBSDB.CTCF_predicted_human)
+summary(width(hg18.CTCFBSDB.CTCF_predicted_human))
+mm8.CTCFBSDB.CTCF_predicted_mouse <- mtx_to_gr(mtx = mtx_mouse, genome_id = "mm8")
+length(mm8.CTCFBSDB.CTCF_predicted_mouse)
+summary(width(mm8.CTCFBSDB.CTCF_predicted_mouse))
 
 # Save as RData object. subdir is the character name of the CTCF GRanges variable
-save(list = hg19.CTCFBSDB.CTCF_predicted_human, file = fileNameOut1)
-save(list = mm9.CTCFBSDB.CTCF_predicted_mouse, file = fileNameOut2)
+save(list = hg18.CTCFBSDB.CTCF_predicted_human, file = fileNameOut1)
+save(list = mm8.CTCFBSDB.CTCF_predicted_mouse, file = fileNameOut2)
 # load(file = paste0(subdir, ".RData"))
 # Save BED files
-export.bed(hg19.CTCFBSDB.CTCF_predicted_human, fileNameOut1.1)
-export.bed(mm9.CTCFBSDB.CTCF_predicted_mouse, fileNameOut2.1)
+export.bed(hg18.CTCFBSDB.CTCF_predicted_human, fileNameOut1.1)
+export.bed(mm8.CTCFBSDB.CTCF_predicted_mouse, fileNameOut2.1)
 ```

### 2022-05-29 EDA_PWMScan.Rmd
PWMscan analysis

--- 2022-05-15 EDA_PWMScan.Rmd
@@
 library(knitr)
-opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700,
+opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700,
 library(pander)
@@
 # Project folder path
-dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev/data/PWMscan/"
+dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev/"
 # Genome assembly
 genome_assembly <- "hg38"
+genome_assembly <- "mm10"
 
@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "")
+fileNameOut1 <- file.path(dir_data, paste0("results/Table_stats_", genome_assembly, ".csv"))
+fileNameOut2 <- file.path(dir_data, paste0("results/Figure_stats_", genome_assembly, ".svg"))
+fileNameOut3 <- file.path(dir_data, paste0("results/Figure_jaccard_", genome_assembly, ".svg"))
 ```
@@
 # Read in data
-files <- list.files(path = dir_data, pattern = genome_assembly, full.names = TRUE)
+files <- list.files(path = file.path(dir_data, "data/PWMscan/"), pattern = genome_assembly, full.names = TRUE)
 # Object to store GRanges 
@@
                           Strand = c(rep("+", length(stats_number_pos)), rep("-", length(stats_number_neg))))
+# Wide format for saving
+mtx_to_save <- pivot_wider(mtx_to_plot, names_from = "Strand", values_from = "Number")
+mtx_to_save <- mtx_to_save %>% mutate(Total = `+` + `-`, .after = List) %>% arrange(desc(Total))
+colnames(mtx_to_save) <- c("Database", "Total CTCF motifs", "Positive strand", "Negative strand")
+write_csv(mtx_to_save, fileNameOut1)
 
@@
   ggtitle(paste(genome_assembly, "genome assembly"))
+ggsave(fileNameOut2, width = 10, height = 3)
 ```
@@
 ```
+
+## Jaccard / GenometriCorr
+
+```{r jaccard, warning=FALSE, fig.width=6.5, fig.height=6}
+save_pheatmap_svg <- function(x, filename, width=4.5, height=3, units = "in", res = 300) {
+  stopifnot(!missing(x))
+  stopifnot(!missing(filename))
+  svg(filename, width=width, height=height)
+  grid::grid.newpage()
+  grid::grid.draw(x$gtable)
+  dev.off()
+}
+# Jaccard calculations
+jaccard <- function(gr_a, gr_b) {
+  intersects <- GenomicRanges::intersect(gr_a, gr_b, ignore.strand = TRUE)
+  intersection <- sum(width(intersects))
+  union <- sum(width(GenomicRanges::union(gr_a, gr_b, ignore.strand = TRUE)))
+  DataFrame(intersection, union, 
+            jaccard = intersection/union,
+             n_intersections = length(intersects))
+}
+# GenometriCorr of smaller query vs. larger reference
+# Returns either Jaccard or ECDF area correlation
+GCorr <- function(gr1, gr2, return_value = "relative.distances.ecdf.area.correlation") {
+  # Select smaller query and larger reference
+  # GenometriCorr itselv
+  gr1_vs_gr2 <- GenometriCorrelation(query, reference, awhole.only = TRUE, showProgressBar = FALSE, permut.number = 0)
+  # What to return
+  if (return_value == "relative.distances.ecdf.area.correlation") {
+    return(gr1_vs_gr2$awhole$relative.distances.ecdf.area.correlation)
+  }
+  if (return_value == "jaccard.measure") {
+    return(gr1_vs_gr2$awhole$jaccard.measure)
+  }
+}
+
+# Correlation matrix, empty
+mtx_to_plot <- matrix(data = 0, nrow = length(bed_list_standard), ncol = length(bed_list_standard))
+# Fill it in
+for (i in 1:length(bed_list_standard)) {
+  for (j in 1:length(bed_list_standard)) {
+    # If diagonal, set to zero
+    if (i == j) mtx_to_plot[i, j] <- 0
+    # Process only one half, the other is symmetric
+    if (i > j) {
+      gr1 <- bed_list_standard[[i]]
+      gr2 <- bed_list_standard[[j]]
+      # Jaccard
+      query <- gr1
+      reference <- gr2
+      mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- jaccard(query, reference)[["jaccard"]]
+      # GenometriCorr
+      # if (length(gr1) >= length(gr2)) {
+      #   reference <- gr1
+      #   query <- gr2
+      # } else {
+      #   reference <- gr2
+      #   query <- gr1
+      # }
+      # mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- GCorr(query, reference)
+    }
+  }
+}
+
+# Trim row/colnames
+rownames(mtx_to_plot) <- colnames(mtx_to_plot) <- str_trunc(names(bed_list_standard), width = 25) 
+# Adjust clustering method
+# if(genome_assembly == "hg38") {
+#   clust_method <- "euclidean"
+# }
+# if(genome_assembly == "mm10") {
+#   clust_method <- "correlation"
+# }
+clust_method <- "euclidean"
+# Save the plot
+# png("man/figures/excluderanges_hg38_jaccard.png", width = 1000, height = 900, res = 200)
+p <- pheatmap(data.matrix(mtx_to_plot), cluster_cols = T, cluster_rows = T,
+         clustering_method = "ward.D2",# "ward.D", 
+         clustering_distance_rows = clust_method, 
+         clustering_distance_cols = clust_method, 
+         # annotation_row = mydf, annotation_colors = list(Group = c(PR = mycols[1], CR = mycols[2])),
+         treeheight_row = 40,
+         treeheight_col = 0, 
+         display_numbers = TRUE)
+save_pheatmap_svg(p, filename = fileNameOut3, width = 5.5, height = 4.5, units = "in", res = 300)
+# dev.off()
+```
+
+```{r eval=FALSE}
+# MDS
+mds <- cmdscale(d = (1 - data.matrix(mtx_to_plot))) %>% as_tibble()
+colnames(mds) <- c("Dim.1", "Dim.2")
+mds <- mds %>% mutate(Groups = sample_annotation$Tumor, Samples = sample_annotation$Sample.Name)
+mds$Groups <- factor(mds$Groups, levels = c("PR", "CR"))
+ggplot(mds, aes(x = Dim.1, y = Dim.2)) + # , color = Groups, label = Samples
+  geom_point(size = 2) +
+  # scale_color_lancet() +
+  scale_color_manual(values = mycols[1:2]) +
+  # scale_color_brewer(palette = "Spectral") +
+  theme_bw() +
+  xlab("Coordinate 1") + ylab("Coordinate 2") 
+#  geom_label_repel(show.legend = FALSE) #, force=1, box.padding=0.5, label.padding = 0.2, direction = c("y"), segment.color = 'grey50', max.overlaps = 15)
+ggsave(file.path(dir_project, "manuscript/figures/figure_replicates/FigureS1A.svg"), width = 4, height = 3)
+```

### 2022-05-29 Logos.Rmd
Add Logos.Rmd

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
library(motifStack)
library(universalmotif)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/"
```

```{r}
# Object to collect motifs
motifs_all <- list()
```

# JASPAR 2022

```{r fig.width = 6, fig.height = 4}
# Input files
fileNameIn1 <- file.path(dir_data, "motif_databases/JASPAR2022_CORE_vertebrates_non-redundant_v2.meme")
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/JASPAR2022_CORE_vertebrates_non-redundant_v2.svg")
# Read in data
motifs <- importMatrix(fileNameIn1)
# plot(motifs[[1]])
motifStack(motifs)
dev.off()
svg(fileNameOut1, width = 6, height = 4)
motifStack(motifs)
dev.off()
# Standardize names
names(motifs) <- paste("JASPAR2022", "CORE", "vertebrates", "CTCF", names(motifs) %>% str_remove(., ".CTCF"), sep = "_")
# Append to the whole list
motifs_all <- c(motifs_all, motifs)
```

# HOCOMOCO v11

```{r fig.width = 6, fig.height = 2}
# Input files
fileNameIn1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_HUMAN_mono_meme_format.meme")
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_HUMAN_mono_meme_format.svg")
# Read in data
motifs <- importMatrix(fileNameIn1)
plot(motifs[[1]])
dev.off()
svg(fileNameOut1, width = 6, height = 2)
plot(motifs[[1]])
dev.off()
# Standardize names
names(motifs) <- paste("HOCOMOCO_v11", "core", "HUMAN", "CTCF", names(motifs) %>% str_remove(., "CTCF_HUMAN."), sep = "_")
# Append to the whole list
motifs_all <- c(motifs_all, motifs)
```

```{r fig.width = 6, fig.height = 2}
# Input files
fileNameIn1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_MOUSE_mono_meme_format.meme")
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_MOUSE_mono_meme_format.svg")
# Read in data
motifs <- importMatrix(fileNameIn1)
plot(motifs[[1]])
dev.off()
svg(fileNameOut1, width = 6, height = 2)
plot(motifs[[1]])
dev.off()
# Standardize names
names(motifs) <- paste("HOCOMOCO_v11", "core", "MOUSE", "CTCF", names(motifs) %>% str_remove(., "CTCF_MOUSE."), sep = "_")
# Append to the whole list
motifs_all <- c(motifs_all, motifs)
```

# SwissRegulon

```{r fig.width = 6, fig.height = 2}
# Input files
fileNameIn1 <- file.path(dir_data, "motif_databases/SwissRegulon_human_and_mouse.meme")
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/SwissRegulon_human_and_mouse.svg")
# Read in data
motifs <- importMatrix(fileNameIn1)
plot(motifs[[1]])
dev.off()
svg(fileNameOut1, width = 6, height = 2)
plot(motifs[[1]])
dev.off()
# Standardize names
names(motifs) <- paste("SwissRegulon", "HUMAN_MOUSE", "CTCF", names(motifs) %>% str_remove(., "CTCF.") %>% str_remove(., fixed(".")), sep = "_")
# Append to the whole list
motifs_all <- c(motifs_all, motifs)
```

# Jomla 2013

Needs to be reverse complement

```{r fig.width = 6, fig.height = 2}
# Input files
fileNameIn1 <- file.path(dir_data, "motif_databases/jolma2013.meme")
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/jolma2013.svg")
# Reverse complement
fileNameOut2 <- file.path(dir_data, "motif_databases/jolma2013_RC.meme")
# Read in data
motifs <- importMatrix(fileNameIn1)
# Reverse complement
motifs_RC <- motif_rc(motifs[[1]])
plot(motifs_RC)
dev.off()
svg(fileNameOut1, width = 6, height = 2)
plot(motifs[[1]])
dev.off()
# Save Reverse complement
write_meme(motifs_RC, fileNameOut2)
# Standardize names
motifs_RC@name <- names(motifs) %>% str_remove(., fixed(".")) # paste("Jomla2013", "HUMAN_MOUSE", "HT_SELEX", "CTCF", names(motifs) %>% str_remove(., "CTCF_") %>% str_remove(., fixed(".")), sep = "_")
# Append to the whole list
motifs_all <- c(motifs_all, motifs_RC)
```

# Clustering

```{r}
## plot stacks with hierarchical tree
motifStack(motifs_all, layout="tree") 
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/Figure_clustered_main_PWMs.svg")
dev.off()
svg(fileNameOut1, width = 8, height = 8)
motifStack(motifs_all, layout="tree") 
dev.off()
```

# CIS_BP Homo Sapiens

```{r}
# Modify input dir
dir_data1 <- file.path(dir_data, "Homo_sapiens_2022_05_20_3_16_pm/pwms_CTCF_motifs")
files <- list.files(dir_data1, pattern = "*.txt", full.names = TRUE)
# Output file
fileNameOut1 <- file.path(dir_data1, "CIS-BP_2.00_Homo_sapiens.meme")
fileNameOut2 <- file.path(dir_data1, "CIS-BP_2.00_Homo_sapiens.svg")
# Standard header
write_lines("MEME version 5.4.1 (Tue Mar 1 19:18:48 2022 -0800)", file = fileNameOut1)
write_lines("ALPHABET= ACGT", file = fileNameOut1, append = TRUE)
write_lines("strands: + -", file = fileNameOut1, append = TRUE)
write_lines("Background letter frequencies (from uniform background):", file = fileNameOut1, append = TRUE)
write_lines("A 0.25000 C 0.25000 G 0.25000 T 0.25000 ", file = fileNameOut1, append = TRUE)
# Store number of bases
nsites <- c()
for(i in 1:length(files)) {
  file=files[i]
  # Make motif name
  motif_name <- basename(file) %>% str_remove(., ".txt")
  mtx <- read_tsv(file, col_types = c("ccccc"))
  # If non-empty, proceed
  if (nrow(mtx) > 0) {
    # Collect number of bases
    nsites <- c(nsites, nrow(mtx))
    # Motif name
    write_lines(paste0("MOTIF ", motif_name), file = fileNameOut1, append = TRUE)
    # Header line
    header_line <- paste0("letter-probability matrix: alength= 4 w= ", nrow(mtx), " nsites= 20 E= 0")
    write_lines(header_line, file = fileNameOut1, append = TRUE)
    # Make first numerical column empty
    mtx$Pos <- ""
    # Write actual data
    write_tsv(mtx, file = fileNameOut1, col_names = FALSE, append = TRUE)
    # Separate from the next motif
    write_lines("", file = fileNameOut1, append = TRUE)
  }
}
print("Number of motifs")
print(length(nsites))
print("Summary of motif widths")
summary(nsites)
# Test if can be imported
motifs_all <- importMatrix(fileNameOut1)
if (!file.exists(fileNameOut2)) {
  dev.off()
  svg(fileNameOut2, width = 8, height = 78)
  motifStack(motifs_all, layout="tree") 
  dev.off()
}
```

# CIS_BP Mus Musculus

```{r}
# Modify input dir
dir_data1 <- file.path(dir_data, "Mus_musculus_2022_05_20_4_01_pm/pwms_CTCF_motifs")
files <- list.files(dir_data1, pattern = "*.txt", full.names = TRUE)
# Output file
fileNameOut1 <- file.path(dir_data1, "CIS-BP_2.00_Mus_musculus.meme")
fileNameOut2 <- file.path(dir_data1, "CIS-BP_2.00_Mus_musculus.svg")
# Standard header
write_lines("MEME version 5.4.1 (Tue Mar 1 19:18:48 2022 -0800)", file = fileNameOut1)
write_lines("ALPHABET= ACGT", file = fileNameOut1, append = TRUE)
write_lines("strands: + -", file = fileNameOut1, append = TRUE)
write_lines("Background letter frequencies (from uniform background):", file = fileNameOut1, append = TRUE)
write_lines("A 0.25000 C 0.25000 G 0.25000 T 0.25000 ", file = fileNameOut1, append = TRUE)
# Store number of bases
nsites <- c()
for(i in 1:length(files)) {
  file=files[i]
  # Make motif name
  motif_name <- basename(file) %>% str_remove(., ".txt")
  mtx <- read_tsv(file, col_types = c("ccccc"))
  # If non-empty, proceed
  if (nrow(mtx) > 0) {
    # Collect number of bases
    nsites <- c(nsites, nrow(mtx))
    # Motif name
    write_lines(paste0("MOTIF ", motif_name), file = fileNameOut1, append = TRUE)
    # Header line
    header_line <- paste0("letter-probability matrix: alength= 4 w= ", nrow(mtx), " nsites= 20 E= 0")
    write_lines(header_line, file = fileNameOut1, append = TRUE)
    # Make first numerical column empty
    mtx$Pos <- ""
    # Write actual data
    write_tsv(mtx, file = fileNameOut1, col_names = FALSE, append = TRUE)
    # Separate from the next motif
    write_lines("", file = fileNameOut1, append = TRUE)
  }
}
print("Number of motifs")
print(length(nsites))
print("Summary of motif widths")
summary(nsites)
# Test if can be imported
motifs_all <- importMatrix(fileNameOut1)
if (!file.exists(fileNameOut2)) {
  dev.off()
  svg(fileNameOut2, width = 8, height = 2.5)
  motifStack(motifs_all, layout="tree") 
  dev.off()
}
```

# CTCFBSDB

EMBL_M1 needs to be reverse complement

```{r}
fileNameIn1 <- file.path(dir_data, "CTCFBSDB_PWM.meme")
# Reverse complement
fileNameOut1 <- file.path(dir_data, "CTCFBSDB_PWM_corrected.meme")
# Read in data
motifs <- importMatrix(fileNameIn1)
# Correct EMBL_M1
motifs[[1]] <- motif_rc(motifs[[1]])
# Save the corrected matrix
write_meme(motifs, fileNameOut1)
# Plot
motifStack(motifs, layout="tree")
# Output files
fileNameOut1 <- file.path(dir_data, "motif_databases/Figure_clustered_CTCFBSDB_PWMs.svg")
dev.off()
svg(fileNameOut1, width = 8, height = 8)
motifStack(motifs, layout="tree") 
dev.off()
```

### 2022-05-29 FIMO_processing.Rmd
Add FIMO_processing

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
library(GenomicRanges)
library(rtracklayer)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot"
# Motifs
motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM_corrected", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013_corrected", "SwissRegulon_human_and_mouse")
motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM_corrected", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013_corrected", "SwissRegulon_human_and_mouse")
# Genomes
genome_human <- c("T2T", "hg38", "hg19")
genome_mouse <- c("mm10", "mm9")
# Chromosomes
chromosome_human <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19", "chr20", "chr21", "chr22", "chrX", "chrY")
chromosome_mouse <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", "chr16", "chr17", "chr18", "chr19", "chrX", "chrY")

# Input files
fileNameIn1 <- file.path(dir_data, "")
# Output files
fileNameOut1 <- file.path(dir_data, "")
```

# Process data

```{r data}
combine_fimo <- function(motif = "SwissRegulon_human_and_mouse", genome_id = "T2T", chromosomes = chromosome_human) {
  ctcfGR <- c()
  for(chrom in chromosomes) {
    # print(chrom)
    fileNameIn <- file.path(dir_data, paste(motif, genome_id, chrom, sep = "_"), "fimo.txt.gz")
    # Read "fimo.bed" created by "fimo.qsub"
    ctcfBED <- read_tsv(fileNameIn, col_types = c("ccddcdddc"))
    # Convert to GRanges object
    ctcf_chr_gr <- GRanges(seqnames = ctcfBED$`sequence name`, 
                      IRanges(start = ctcfBED$start, end = ctcfBED$stop),
                      strand = ctcfBED$strand)
    # Add metadata
    ctcf_chr_gr$score = ctcfBED$score
    ctcf_chr_gr$pvalue = ctcfBED$`p-value`
    ctcf_chr_gr$qvalue = ctcfBED$`q-value`
    ctcf_chr_gr$sequence = ctcfBED$`matched sequence`
    ctcfGR <- c(ctcfGR, ctcf_chr_gr)
  }
  # Combine all chromosomes
  ctcfGR <- do.call(c, as(ctcfGR, "GRangesList"))
  # Sort
  ctcfGR <- ctcfGR %>% sort()

  # Add seqinfo
  # Parse out genome ID from the folder name, to get hg19, hg38, mm9, or mm10
  if (genome_id == "T2T") {
    # Seqinfor for T2T genome
    chrom_data <- GenomeInfoDb::getChromInfoFromNCBI(assembly = "GCA_009914755.4")
    chrom_data$AssignedMolecule <- as.character(paste0("chr", chrom_data$AssignedMolecule))
    chrom_data <- chrom_data[chrom_data$AssignedMolecule %in% seqlevels(ctcfGR), ]
    chrom_data <- chrom_data[match(seqlevels(ctcfGR), chrom_data$AssignedMolecule), ]
    # Check if chromosome order is the same
    if (!all.equal(seqlevels(ctcfGR), chrom_data$AssignedMolecule)) {
      print(paste("Chromosome order does not match for", genome_id, "genome."))
      break
    }
    # Assign seqinfo data
    seqlengths(ctcfGR) <- chrom_data$SequenceLength
    isCircular(ctcfGR) <- ifelse(is.na(chrom_data$circular), FALSE, TRUE)
    genome(ctcfGR)     <- "GCA_009914755.4"
  } else {
    # Get chromosome info and match it to the chromosome order in ctcfBED
    chrom_data <- GenomeInfoDb::getChromInfoFromUCSC(genome = genome_id)
    chrom_data <- chrom_data[chrom_data$chrom %in% seqlevels(ctcfGR), ]
    chrom_data <- chrom_data[match(seqlevels(ctcfGR), chrom_data$chrom), ]
    # Check if chromosome order is the same
    if (!all.equal(seqlevels(ctcfGR), chrom_data$chrom)) {
      print(paste("Chromosome order does not match for", genome_id, "genome."))
      break
    }
    # Assign seqinfo data
    seqlengths(ctcfGR) <- chrom_data$size
    isCircular(ctcfGR) <- chrom_data$circular
    genome(ctcfGR)     <- genome_id
  }

  ctcfGR_to_save <- paste(genome_id, motif, "FIMO", sep = ".")
  fileNameOut1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".RData"))
  fileNameOut1.1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".bed"))
  # Assign this object to the subfolder-specific variable name
  assign(ctcfGR_to_save, ctcfGR)
  # Save as RData object. ctcfGR_to_save is the character name of the CTCF GRanges variable
  save(list = ctcfGR_to_save, file = fileNameOut1)
  # load(file = fileNameOut1)
  export.bed(eval(parse(text = ctcfGR_to_save)), fileNameOut1.1)
  # Summary
  # Width
  summary_gr <- summary(width(eval(parse(text = ctcfGR_to_save))))
  # All summary
  summary_to_save <- data.frame(File = ctcfGR_to_save,
                                N = length(width(eval(parse(text = ctcfGR_to_save)))), 
                                Min = as.numeric(summary_gr["Min."]), 
                                Median = as.numeric(summary_gr["Median"]), 
                                Mean = as.numeric(summary_gr["Mean"]), 
                                Max = as.numeric(summary_gr["Max."]))
  fileNameLog <- file.path(dir_data, "log_PWMs.csv")
  write_csv(summary_to_save, fileNameLog, append = TRUE)
}
```

# Human data hg18

```{r hg18}
for (gen in c("hg18", genome_human)) {
  for (mot in "MA0139.1") {
    print(paste(gen, mot, sep = "."))
    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_human)
  }
}
```

# Human data all

```{r hgall}
# motif_human  <- c("CTCFBSDB_PWM_corrected", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non-redundant_v2", "jolma2013_corrected", "SwissRegulon_human_and_mouse")
for (gen in genome_human) {
  for (mot in motif_human) {
    print(paste(gen, mot, sep = "."))
    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_human)
  }
}
```

# Mouse data all

```{r mmall}
for (gen in genome_mouse) {
  for (mot in motif_mouse) {
    print(paste(gen, mot, sep = "."))
    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_mouse)
  }
}
```

### 2022-05-29 EDA_liftOver.Rmd
Add EDA_liftOver

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
library(pheatmap)
# Color palette for the heatmap, https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
col3 <- colorRampPalette(c('blue', 'white', 'red'))(20)
# col3 <- colorRampPalette(c('blue', 'gray', 'yellow'))(20)
# col3 <- colorRampPalette(c('green', 'black', 'red'))(20)
# col3 <- colorRamps::green2red(n = 20)
library(rtracklayer)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/CTCF_liftover/"

# Input files
fileNameIn1 <- file.path(dir_data, "sample_annotation.csv")
# Output files
fileNameOut1 <- file.path(dir_data, "Figure_liftOverJaccard.svg")
```

# Load data

```{r data}
sample_annotation <- read_csv(fileNameIn1)
# Read in data
files <- sample_annotation$File
genome_assembly <- sample_annotation$Assembly
# Object to store GRanges 
bed_list_all <- list() # All chromosomes
bed_list_standard <- list() # Standard chromosomes
# Additional columns
# extraCols_names <- c("character", "character", "character")
# names(extraCols_names) <- c("motif", "pval", "qval")
# Process each file
for(i in 1:length(files)) {
  print(basename(files[i]))
  # Read data
  bed_file <- import(file.path(dir_data, "data", files[i]), format = "bed") # , genome = genome_assembly[i], colnames = c("chrom", "start", "end", "name", "score", "strand", "motif", "pval", "qval"), extraCols = extraCols_names
  # Save all data
  bed_list_all <- c(bed_list_all, list(bed_file))
  # Keep standard chromosomes
  bed_file_standard <- keepStandardChromosomes(bed_file, pruning.mode="coarse")
  # Save filtered data
  bed_list_standard <- c(bed_list_standard, list(bed_file_standard))
}
# Add names
names(bed_list_all) <- sample_annotation$Name
names(bed_list_standard) <- sample_annotation$Name
```

## Jaccard / GenometriCorr

```{r jaccard, warning=FALSE, fig.width=6.5, fig.height=6}
save_pheatmap_svg <- function(x, filename, width=4.5, height=3, units = "in", res = 300) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  svg(filename, width=width, height=height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}
# Jaccard calculations
jaccard <- function(gr_a, gr_b) {
  intersects <- GenomicRanges::intersect(gr_a, gr_b, ignore.strand = TRUE)
  intersection <- sum(width(intersects))
  union <- sum(width(GenomicRanges::union(gr_a, gr_b, ignore.strand = TRUE)))
  DataFrame(intersection, union, 
            jaccard = intersection/union,
             n_intersections = length(intersects))
}
# GenometriCorr of smaller query vs. larger reference
# Returns either Jaccard or ECDF area correlation
GCorr <- function(gr1, gr2, return_value = "relative.distances.ecdf.area.correlation") {
  # Select smaller query and larger reference
  # GenometriCorr itselv
  gr1_vs_gr2 <- GenometriCorrelation(query, reference, awhole.only = TRUE, showProgressBar = FALSE, permut.number = 0)
  # What to return
  if (return_value == "relative.distances.ecdf.area.correlation") {
    return(gr1_vs_gr2$awhole$relative.distances.ecdf.area.correlation)
  }
  if (return_value == "jaccard.measure") {
    return(gr1_vs_gr2$awhole$jaccard.measure)
  }
}

# Correlation matrix, empty
mtx_to_plot <- matrix(data = 0, nrow = length(bed_list_standard), ncol = length(bed_list_standard))
# Fill it in
for (i in 1:length(bed_list_standard)) {
  for (j in 1:length(bed_list_standard)) {
    # If diagonal, set to zero
    if (i == j) mtx_to_plot[i, j] <- 1
    # Process only one half, the other is symmetric
    if (i > j) {
      gr1 <- bed_list_standard[[i]]
      gr2 <- bed_list_standard[[j]]
      # Jaccard
      query <- gr1
      reference <- gr2
      mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- jaccard(query, reference)[["jaccard"]]
      # GenometriCorr
      # if (length(gr1) >= length(gr2)) {
      #   reference <- gr1
      #   query <- gr2
      # } else {
      #   reference <- gr2
      #   query <- gr1
      # }
      # mtx_to_plot[i, j] <- mtx_to_plot[j, i] <- GCorr(query, reference)
    }
  }
}

# Trim row/colnames
rownames(mtx_to_plot) <- colnames(mtx_to_plot) <- str_trunc(names(bed_list_standard), width = 25) 
```

# Pheatmap

```{r}
# Adjust clustering method
# if(genome_assembly == "hg38") {
#   clust_method <- "euclidean"
# }
# if(genome_assembly == "mm10") {
#   clust_method <- "correlation"
# }
clust_method <- "euclidean"
# Save the plot
col3 <- colorRampPalette(c('white', 'white','white','white','white','white','white','orangered'))(20)
# all.equal(colnames(mtx_to_plot), sample_annotation$Name)
mydf <- data.frame(Assembly = sample_annotation$Assembly)
rownames(mydf) <- sample_annotation$Name
# png("man/figures/excluderanges_hg38_jaccard.png", width = 1000, height = 900, res = 200)
p <- pheatmap(data.matrix(mtx_to_plot), cluster_cols = T, cluster_rows = T, color = col3,
         clustering_method = "ward.D",# "ward.D", 
         clustering_distance_rows = clust_method, 
         clustering_distance_cols = clust_method, 
         annotation_row = mydf, annotation_colors = list(Assembly = c(hg18 = mycols[2], hg19 = mycols[4], hg38 = mycols[5], T2T = mycols[1])),
         treeheight_row = 40,
         treeheight_col = 0, 
         display_numbers = TRUE)

# mtx_to_barplot <- data.frame(hg19 = mtx_to_plot["hg19", "hg19 (from hg38)"],
#                              hg38 = mtx_to_plot["hg38", "hg38 (from hg19)"],
#                              T2T = mtx_to_plot["T2T (from hg19)", "T2T (from hg38)"])
# mtx_to_barplot <- mtx_to_barplot %>% pivot_longer(., cols = everything(), names_to = "Assembly", values_to = "Jaccard")
# mtx_to_barplot$Assembly <- c("hg19 vs.\nhg19 from hg38", "hg38 vs.\nhg38 from hg19", "T2T from hg19 vs.\nT2T from hg38")
# mtx_to_barplot$Assembly <- factor(mtx_to_barplot$Assembly, levels = rev(mtx_to_barplot$Assembly))
# ggplot(mtx_to_barplot, aes(x = Assembly, y = Jaccard, fill = Assembly)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   theme_bw(base_size = 15) +
#   # get rid of the grid 
#   theme(panel.grid.major = element_blank(), 
#         panel.grid.minor = element_blank(), 
#         text = element_text(size = 15),
#         legend.position = "none") + # c(0.8, 0.30)
#   scale_fill_manual(values = mycols[1:3]) + # change colors 
#   xlab("Assembly") +
#   ylab("Jaccard")

save_pheatmap_svg(p, filename = fileNameOut1, width = 7.5, height = 5.5, units = "in", res = 300)
# dev.off()
```

### 2022-05-29 manuscript/manuscript.Rmd
More writeup

@@
 Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
+CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs].
 Point mutation and loss of heterozygosity of CTCF is associated with human cancer, identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].
@@
 
-CTCF localization is largely invariant across different cell types [@Kim:2007ux].
-The goal of our package is to provide such lists of genome-wide CTCF binding sites generally applicable across cell types.
+ with binding motif length varying between 20 and 40bp
 
-CTCF binds to different DNA sequences through combinatorial use of 11-zinc finger domains to target sites that have remarkable sequence variation with binding motif length varying between 20 and 40bp [@Filippova:1996tv][@Ohlsson:2001wu].
+CTCF binds to different DNA sequences through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Filippova:1996tv][@Ohlsson:2001wu][@Jolma:2013vd].
 A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1.
-It was subsequently refined to include an additional 9bp motif (M2) [@Schmidt:2012uk].
-For the majority of CTCF–DNA- binding events, the N-terminal zinc fingers interact with a 14-bp long M1 motif.
-Additionally, the C-terminal fingers interact with a shorter M2 sequence, creating a 34-bp-long M1+M2 motif.
+This motif interacts with the N-terminal zinc fingers for the majority of CTCF–DNA-binding events.
+Subsequently, a shorter motif (M2) [@Schmidt:2012uk], interacting with the C-terminal fingers.
 The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al.
 [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding.
-Studies have reported 20,000-50,000 CTCF binding sites within the human genome [@Kim:2007ux][@Nakahashi:2013uw].
-
-Position Weight Matrices (PWMs) [@Stormo:2000uw] for CTCF are available from multiple databases.
-JASPAR 2022 [@Castro-Mondragon:2022uv] and HOCOMOCO [@Kulakovskiy:2018vw] databases list one 19bp-long CTCF PWM ([MA0139.1](https://jaspar.genereg.net/matrix/MA0139.1/) and [CTCF_HUMAN.H11MO.0.A](https://hocomoco11.autosome.org/motif/CTCF_HUMAN.H11MO.0.A), respectively).
-Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific [CTCF](http://cisbp.ccbr.utoronto.ca/TFreport.php?searchTF=T094831_2.00) binding motifs, likely reflecting the variability of CTCF motifs without considering the M1+M2 combinatorial structure.
-The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al.
-[@Schmidt:2012uk], the Ren_20 motif identified by Kim et al.
-[@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al.
-[@Xie:2007wm].
-
-Unfortunately, the database has not been updated since 2013 and its data are limited to old genome assemblies (e.g., hg19, mm9).
+Subsequently, $\sim80%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences and ZFs 9-11 associate with a second conserved motif [@Nakahashi:2013uw][@Pugacheva:2020wz].
+Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux][@Nakahashi:2013uw][@Wang:2012ty].
 
 CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa].
-Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa].
+Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa][@Pugacheva:2020wz].
 To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops
@@
 
-We utilized the PWMscan web tool [@Ambrosini:2018wv] because it has a large collection of PWMs (JASPAR, HOCOMOCO, SwissRegulon, UniPROBE, CIS-BP, data from Jomla, Isakova publications) and genomes (49 genome assemblies); its results show over 99% agreement with FIMO, HOMER, and other motif scanning tools.
+We aimed to collect general-purpose CTCF binding sites for human and mouse genomes, the top two species by the number of experiments in GEO.
+CTCF binding profiles between different tissues exhibit substantial concordance [@Kim:2007ux].
+For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].
 
@@
 
+MEME motif databases (updated 20 Mar 2022) containing HOCOMOCO, CIS-BP, JASPAR2022, Jolma 2013, and SwissRegulon motifs in MEME format were downloaded from https://meme-suite.org/meme/meme-software/Databases/motifs/motif_databases.12.23.tgz.
+Motifs from the CTCFBSDB database were downloaded from https://insulatordb.uthsc.edu/download/CTCFBSDB_PWM.mat and converted to MEME format using `transfac2meme -use_acc CTCFBSDB_PWM.mat > CTCFBSDB_PWM.meme`.
+Human and mouse motifs from CIS-BP were downloaded via the bulk download interface, combined and converted to the MEME format using a custom script.
+Motifs were visualized and clustered using the `motifStack` v.1.38.0 R package.
+Jolma 2013 and CTCFBSDB EMBL_M1 motifs were converted to their reverse complements to match orientation of other motifs (`universalmotif` v.1.24.4).
+
+CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] was used for motif scanning using default parameters.
+Softmasked genome sequences were downloaded from the USCS Genome Browser database with the exception of T2T genome (chm13v2.0) downloaded from the https://github.com/marbl/CHM13 repository.
+Motifs were scanned separately for each chromosome (autosomal and sex chromosomes only), and the results were combined.
+We used human- and mouse-specific matrices for the corresponding genomes (HOCOMOCO, CIS-BP).
+As JASPAR, CTCFBSDB, and CIS-BP databases provide several CTCF motifs, all database-specific motifs were used to scan the genomes; consequently, the detected CTCF sites may overlap due to motif similarity.
+Predefined CTCF binding sites were obtained from the SCREEN ENCODE and CTCFBSDB databases.
+We did not include CTCFBSDB data compiled from public experimental data because 1) mouse data was lacking, 2) hg18 genome assembly limits usability, and 3) region size was highly variable, ranging from 2bp to >350Kb.
+
+To investigate the effect of coordinate conversion, we used `liftOver` v.377 on CTCF binding sites (JASPAR 2022 MA0139.1) aligned to hg18, hg19, hg38, and T2T (NCBI GCA_009914755.4 assembly) genomes.
+The conversion chains were obtained from the UCSC Genome Browser database except for the T2T genome providing chains at the https://github.com/marbl/CHM13 repository.
+We performed all conversions except hg38->hg18 and hg18->T2T, T2T->hg18 due to lack of the corresponding chains.
+Jaccard overlap matrix between the original and liftOver CTCF binding sites was clustered using Euclidean distance and ward.D clustering method and visualized using the `pheatmap` v.1.0.12 R package.
+
 # Results
+
+CTCF binding specificities in a form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases.
+The JASPAR 2022 [@Castro-Mondragon:2022uv] database list one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table`).
+The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure_clustered_main_PWMs`).
+Although human and mouse CTCF motifs are considered highly conserved [@Xie:2007wm], the HOCOMOCO database provides mouse-specific CTCF PWM which is expectedly similar to human (`Figure_clustered_main_PWMs`).
+SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs.
+The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the 17bp PWM identified in Jolma et al.
+study [@Jolma:2013vd] (`Figure_clustered_main_PWMs`).
+The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al.
+[@Schmidt:2012uk], the Ren_20 motif identified by Kim et al.
+[@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al.
+[@Xie:2007wm] (`Figure_clustered_CTCFBSDB_PWMs.svg`).
+The Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), likely reflecting the variability of CTCF motifs without considering the M1+M2 combinatorial structure (`Table`).
+These PWMs were used to scan human and mouse genomes for CTCF binding motifs.
+
+Several databases and papers provide predicted and experimental CTCF binding information.
+The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq].
+The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%).
+The mouse set contained 82,777 records (median size 284bp) similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites.
+A limitation of this data is that it lacks strand information.
+We added human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.
+
+The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes.
+The database has not been updated from 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table`).
+Coordinate conversion between genome assemblies is a well-established practice; consequently, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover versions.
+Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure_liftOverJaccard.svg`).
+liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table_liftOver_stats.xlsx`).
+These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly.
+We provide hg19 and hg38 versions of predicted CTCF sites from the CTCFBSDB database.
 
@@
 
-CTCFBSDB assembled predicted and experimentally detected CTCF binding information [@Bao:2008vq].
-Human predictions containes 13,356 sites and mouse predictions contained 5,504 sites (hg19 and mm9 genome assemblies, respectively).
-Unfortunately, the database has not been updated since 2012; therefore, its data is of limited use.
-
-The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq].
-The human set of CTCF-bound cCREs contains 450,641 records.
-They were classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%).
-The mouse set contained 82,777 records similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites.
+Jolma et al.
+[@Jolma:2013vd] analyzed the sequence-specific binding of human TFs using high-throughput SELEX and ChIP sequencing.
+This study defined 830 binding profiles were obtained, describing 239 distinctly different binding specificities.
+They defined a 17bp long CTCF motif dubbed CTCF_full.
 
@@
 +-------+
+
 # Discussion
+
+Our goal was to provide easy programmatic access to CTCF binding data that is generally applicable for any cell or tissue type.
+For those interested in cell-type-specific CTCF binding data, many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, provide download and/or REST API access to it.
+Bioconductor's AnnotationHub itself contains human and mouse CTCF binding data from ENCODE.
+Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data.
+Given the use of cell-type-specific data is specialized, we aimed to provide broadly applicable CTCF binding data.
+
+Besides FIMO, other tools for motif scanning can be utilized.
+They have their own algorithms and defaults which may lead to different results.
+The popular HOMER motif screening tool [@Heinz:2010aa] has several options, including its unique autonormalization of sequencing bias algorithm.
+The PWMscan genomewide motif scanning web tool [@Ambrosini:2018wv] uses its own matrix_scan C program.
+PWMscan has the default background nucleotide composition A=0.29, C=0.21, G=0.21, T=0.29 instead of the FIMO's default A=0.25, C=0.25, G=0.25, T=0.25, and uses a 1E-5 p-value threshold instead of 1E-4 for FIMO.
+PWMscan performed extensive benchmarking of its scanning tool and showed over 99% agreement with FIMO, HOMER, and other motif scanning tools.
+Our data is includes results using FIMO tool and we expect these results to be generally consistent when using other tools.
 

### 2022-05-30 FIMO_processing.Rmd
Make Summary a separate chunk

@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "")
+fileNameOut1 <- file.path(dir_data, "log_PWMs.csv")
 ```
@@
   export.bed(eval(parse(text = ctcfGR_to_save)), fileNameOut1.1)
-  # Summary
-  # Width
-  summary_gr <- summary(width(eval(parse(text = ctcfGR_to_save))))
-  # All summary
-  summary_to_save <- data.frame(File = ctcfGR_to_save,
-                                N = length(width(eval(parse(text = ctcfGR_to_save)))), 
-                                Min = as.numeric(summary_gr["Min."]), 
-                                Median = as.numeric(summary_gr["Median"]), 
-                                Mean = as.numeric(summary_gr["Mean"]), 
-                                Max = as.numeric(summary_gr["Max."]))
-  fileNameLog <- file.path(dir_data, "log_PWMs.csv")
-  write_csv(summary_to_save, fileNameLog, append = TRUE)
 }
@@
 ```
+
+# Summary
+
+```{r}
+files <- list.files(dir_data, pattern = "*.bed", full.names = TRUE)
+for (file in files) {
+  print(basename(file) %>% str_remove(., ".bed"))
+  gr <- import.bed(file)
+  # Summary
+  summary_to_save <- data.frame(File = basename(file) %>% str_remove(., ".bed"),
+                                Assembly = strsplit(basename(file), ".", fixed = TRUE)[[1]][1],
+                                All = length(gr),
+                                Reduced = length(reduce(gr)))
+  write_csv(summary_to_save, fileNameOut1, append = TRUE)
+}
+
+```

### 2022-06-03 BED_to_BEDPE.Rmd
Add BED_to_BEDPE and FIMO_EDA

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
# Color palette for the heatmap, https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
col3 <- colorRampPalette(c('blue', 'white', 'red'))(20)
# col3 <- colorRampPalette(c('blue', 'gray', 'yellow'))(20)
# col3 <- colorRampPalette(c('green', 'black', 'red'))(20)
# col3 <- colorRamps::green2red(n = 20)

```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/Avocado_preciseTAD/Maggie/GM12878/PTBR_Peakachu_outputs/"

# Input files
fileNameIn1 <- file.path(dir_data, "")
# Output files
fileNameOut1 <- file.path(dir_data, "")
```

# Convert BED to BEDPE

```{r data}
# Read in data
files <- list.files(dir_data, pattern = "*.bed$")
for (fil in files) {
  print(fil)
  # BED data
  mtx <- read_tsv(file.path(dir_data, fil), col_names = FALSE)
  # Convert to pairs
  bedpe <- rbind()
  for (i in 1:(nrow(mtx) - 1)) {
    bedpe <-rbind(bedpe, c(mtx[i, 1:3], mtx[i + 1, 1:3]))
  }
  # BEDPE file name
  newfile <- file.path(dir_data, paste0(fil, "pe"))
  write.table(bedpe, newfile, quote = FALSE, sep = "\t", col.names = FALSE, row.names = FALSE)
  # write_tsv(as.data.frame(bedpe), newfile, col_names = FALSE)
}
```

### 2022-06-03 FIMO_EDA.Rmd
Add BED_to_BEDPE and FIMO_EDA

---
title: "Analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
# Color palette for the heatmap, https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
col3 <- colorRampPalette(c('blue', 'white', 'red'))(20)
# col3 <- colorRampPalette(c('blue', 'gray', 'yellow'))(20)
# col3 <- colorRampPalette(c('green', 'black', 'red'))(20)
# col3 <- colorRamps::green2red(n = 20)

```

# Settings

```{r settings}
dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot"
# Motifs
motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
# Genomes
genome_human <- c("T2T", "hg38", "hg19")
genome_mouse <- c("mm10", "mm9")
```

```{r hg18, eval=FALSE}
# Human data hg18
pvals <- rbind()
pval_threshold <- 1e-6
for (gen in c("hg18", genome_human)) {
  for (mot in "JASPAR2022_MA0139.1") {
    print(paste(gen, mot, sep = "."))
    mtx <- read_tsv(fileNameIn, col_names = FALSE, col_types = c("ciicdciddc"))
    fileNameIn <- file.path(dir_data, paste(gen, mot, "bed", sep = "."))
    print(paste("Number of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold), digits = 5) ))
    print(paste("Proportion of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold) / nrow(mtx), digits = 5) ))
    pvals <- rbind(pvals, cbind(-log10(mtx$X8), rep(gen, nrow(mtx))))
    # pvals$pvals <- c(pvals$pvals, -log10(mtx$X8))
    # pvals$assembly <- c(pvals$assembly, rep(gen, nrow(mtx)))
  }
}
pvals <- as.data.frame(pvals)
pvals[, 1] <- as.numeric(pvals[, 1])
colnames(pvals) <- c("-log10(p-value)", "Assembly")

ggplot(pvals, aes(x = `-log10(p-value)`, color = Assembly)) +
  geom_density()
```

# Human data all

```{r hgall}
pvals <- rbind()
pval_threshold <- 1e-6
for (gen in "hg38") {
  for (mot in motif_human) {
    print(paste(gen, mot, sep = "."))
    fileNameIn <- file.path(dir_data, paste(gen, mot, "bed", sep = "."))
    mtx <- read_tsv(fileNameIn, col_names = FALSE, col_types = c("ciicdciddc"))
    print(paste("Number of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold), digits = 5) ))
    print(paste("Proportion of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold) / nrow(mtx), digits = 5) ))
    pvals <- rbind(pvals, cbind(-log10(mtx$X8), rep(mot, nrow(mtx))))
  }
}
pvals <- as.data.frame(pvals)
pvals[, 1] <- as.numeric(pvals[, 1])
colnames(pvals) <- c("-log10(p-value)", "MotifDB")

ggplot(pvals, aes(x = `-log10(p-value)`, color = MotifDB)) +
  geom_density() +
  geom_vline(xintercept = -log10(pval_threshold), linetype = "dashed", color = "red") +
  scale_color_manual(values = mycols[1:6]) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5)) +
  ylab("Density")

fileNameOut1 <- file.path(dir_data, "Figure_human_pvalues.svg")
ggsave(filename = fileNameOut1, width = 7, height = 3)
```

# Mouse data all

```{r mmall}
pvals <- rbind()
pval_threshold <- 1e-6
for (gen in "mm10") {
  for (mot in motif_mouse) {
    print(paste(gen, mot, sep = "."))
    fileNameIn <- file.path(dir_data, paste(gen, mot, "bed", sep = "."))
    mtx <- read_tsv(fileNameIn, col_names = FALSE, col_types = c("ciicdciddc"))
    print(paste("Number of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold), digits = 5) ))
    print(paste("Proportion of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold) / nrow(mtx), digits = 5) ))
    pvals <- rbind(pvals, cbind(-log10(mtx$X8), rep(mot, nrow(mtx))))
  }
}
pvals <- as.data.frame(pvals)
pvals[, 1] <- as.numeric(pvals[, 1])
colnames(pvals) <- c("-log10(p-value)", "MotifDB")

ggplot(pvals, aes(x = `-log10(p-value)`, color = MotifDB)) +
  geom_density() +
  geom_vline(xintercept = -log10(pval_threshold), linetype = "dashed", color = "red") +
  scale_color_manual(values = mycols[1:6]) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5)) +
  ylab("Density")

fileNameOut1 <- file.path(dir_data, "Figure_mouse_pvalues.svg")
ggsave(filename = fileNameOut1, width = 7, height = 3)
```

### 2022-06-03 FIMO_processing.Rmd
Add BED_to_BEDPE and FIMO_EDA

@@
 # Motifs
-motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM_corrected", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013_corrected", "SwissRegulon_human_and_mouse")
-motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM_corrected", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013_corrected", "SwissRegulon_human_and_mouse")
+motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
+motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
 # Genomes
@@
     # Add metadata
+    ctcf_chr_gr$name = ctcfBED$`#pattern name`
     ctcf_chr_gr$score = ctcfBED$score
@@
   # load(file = fileNameOut1)
-  export.bed(eval(parse(text = ctcfGR_to_save)), fileNameOut1.1)
+  # export.bed(eval(parse(text = ctcfGR_to_save)), fileNameOut1.1)
+  ctcfDF_to_save <- as.data.frame(eval(parse(text = ctcfGR_to_save)))
+  ctcfDF_to_save <- ctcfDF_to_save[, c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")]
+  write_tsv(ctcfDF_to_save, fileNameOut1.1, col_names = FALSE)
 }
@@
 for (gen in c("hg18", genome_human)) {
-  for (mot in "MA0139.1") {
+  for (mot in "JASPAR2022_MA0139.1") {
     print(paste(gen, mot, sep = "."))
@@
 ```{r}
+# for file in `find . -type f -name "*MA0139*"`; do
+#   echo $file
+#   newfile=`echo $file | sed 's/MA0139/JASPAR2022_MA0139/'`
+#   echo $newfile
+#   mv $file $newfile
+# done
+# 
 files <- list.files(dir_data, pattern = "*.bed", full.names = TRUE)
@@
   print(basename(file) %>% str_remove(., ".bed"))
-  gr <- import.bed(file)
+  mtx <- read_tsv(file, col_names = FALSE)
+  colnames(mtx) <- c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")
+  gr <- makeGRangesFromDataFrame(mtx)
   # Summary
@@
 }
-
 ```

### 2022-06-03 FIMO_processing.Rmd
Savind extended summary and tweaks

@@
 fileNameIn1 <- file.path(dir_data, "")
-# Output files
-fileNameOut1 <- file.path(dir_data, "log_PWMs.csv")
 ```
@@
 
-  ctcfGR_to_save <- paste(genome_id, motif, "FIMO", sep = ".")
+  ctcfGR_to_save <- paste(genome_id, motif, sep = ".")
   fileNameOut1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".RData"))
@@
 # 
+# Output files
+fileNameOut1 <- file.path(dir_data, "log_PWMs.csv")
+if (file.exists(fileNameOut1)) { file.remove(fileNameOut1) }
+# Defined from FIMO_EDA
+pval_threshold <- 1e-6
 files <- list.files(dir_data, pattern = "*.bed", full.names = TRUE)
-for (file in files) {
-  print(basename(file) %>% str_remove(., ".bed"))
-  mtx <- read_tsv(file, col_names = FALSE)
+for (fil in files) {
+  print(basename(fil) %>% str_remove(., ".bed"))
+  mtx <- read_tsv(fil, col_names = FALSE, col_types = c("ciicdciddc"))
   colnames(mtx) <- c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")
-  gr <- makeGRangesFromDataFrame(mtx)
+  gr <- makeGRangesFromDataFrame(mtx, keep.extra.columns = TRUE)
   # Summary
-  summary_to_save <- data.frame(File = basename(file) %>% str_remove(., ".bed"),
-                                Assembly = strsplit(basename(file), ".", fixed = TRUE)[[1]][1],
-                                All = length(gr),
-                                Reduced = length(reduce(gr)))
+  summary_to_save <- data.frame(File = basename(fil) %>% str_remove(., ".bed"),
+                                Assembly = strsplit(basename(fil), ".", fixed = TRUE)[[1]][1],
+                                `All (p-value threshold 1e-4)` = length(gr),
+                                `Reduced (p-value threshold 1e-4)` = length(reduce(gr)),
+                                `All (p-value threshold 1e-6)` = length(gr[gr$pvalue < pval_threshold]),
+                                `Reduced (p-value threshold 1e-6)` = length(reduce(gr[gr$pvalue < pval_threshold])) )
   write_csv(summary_to_save, fileNameOut1, append = TRUE)
+  gc()
 }

### 2022-06-03 Logos.Rmd
Trying processing ENCODE-provided PWMs, not use

@@
 ```
+
+# ENCODE
+
+```{r eval=FALSE}
+# Modify input dir
+dir_data1 <- file.path(dir_data, "ENCODE_motifs_CTCF")
+files <- list.files(dir_data1, pattern = "*.txt", full.names = TRUE)
+# Output file
+fileNameOut1 <- file.path(dir_data1, "motifs_CTCF.meme")
+fileNameOut2 <- file.path(dir_data1, "motifs_CTCF.svg")
+# Standard header
+write_lines("MEME version 5.4.1 (Tue Mar 1 19:18:48 2022 -0800)", file = fileNameOut1)
+write_lines("ALPHABET= ACGT", file = fileNameOut1, append = TRUE)
+write_lines("strands: + -", file = fileNameOut1, append = TRUE)
+write_lines("Background letter frequencies (from uniform background):", file = fileNameOut1, append = TRUE)
+write_lines("A 0.25000 C 0.25000 G 0.25000 T 0.25000 ", file = fileNameOut1, append = TRUE)
+# Store number of bases
+nsites <- c()
+for(i in 1:length(files)) {
+  fileNameIn=files[i]
+  mtx <- read.table(fileNameIn, sep = " ", fill = TRUE)
+  # Make motif name
+  motif_name <- strsplit(mtx[1, 1], "\t")[[1]]
+  motif_id <- motif_name[1] %>% str_remove(">")
+  motif_alt <- motif_name[length(motif_name)]
+  # Drop header row
+  mtx <- mtx[2:nrow(mtx), ]
+  # If non-empty, proceed
+  if (nrow(mtx) > 0) {
+    # Collect number of bases
+    nsites <- c(nsites, nrow(mtx))
+    # Motif name
+    write_lines(paste0("MOTIF ", paste(motif_id, motif_alt)), file = fileNameOut1, append = TRUE)
+    # Header line
+    header_line <- paste0("letter-probability matrix: alength= 4 w= ", nrow(mtx), " nsites= 20 E= 0")
+    write_lines(header_line, file = fileNameOut1, append = TRUE)
+    # Make first numerical column empty
+    mtx[, 1] <- ""
+    # Write actual data
+    write_tsv(mtx, file = fileNameOut1, col_names = FALSE, append = TRUE)
+    # Separate from the next motif
+    write_lines("", file = fileNameOut1, append = TRUE)
+  }
+}
+print("Number of motifs")
+print(length(nsites))
+print("Summary of motif widths")
+summary(nsites)
+# Test if can be imported
+motifs_all <- importMatrix(fileNameOut1)
+if (!file.exists(fileNameOut2)) {
+  dev.off()
+  svg(fileNameOut2, width = 10, height = 14)
+  motifStack(motifs_all, layout="tree") 
+  dev.off()
+  }
+```
+
+```{r}
+# Modify input file
+fileNameIn1 <- file.path(dir_data, "motifs_CTCF.txt")
+# Output file
+fileNameOut1 <- file.path(dir_data1, "motifs_CTCF.meme")
+fileNameOut2 <- file.path(dir_data1, "motifs_CTCF.svg")
+# Standard header
+write_lines("MEME version 5.4.1 (Tue Mar 1 19:18:48 2022 -0800)", file = fileNameOut1)
+write_lines("ALPHABET= ACGT", file = fileNameOut1, append = TRUE)
+write_lines("strands: + -", file = fileNameOut1, append = TRUE)
+write_lines("Background letter frequencies (from uniform background):", file = fileNameOut1, append = TRUE)
+write_lines("A 0.25000 C 0.25000 G 0.25000 T 0.25000 ", file = fileNameOut1, append = TRUE)
+# Store number of bases
+nsites <- c()
+for(i in 1:length(files)) {
+  file=files[i]
+  mtx <- read.table(fileNameIn1, sep = " ", fill = TRUE)
+  # Make motif name
+  motif_name <- strsplit(mtx[1, 1], "\t")[[1]]
+  motif_id <- motif_name[1] %>% str_remove(">")
+  motif_alt <- motif_name[length(motif_name)]
+  # If non-empty, proceed
+  if (nrow(mtx) > 0) {
+    # Collect number of bases
+    nsites <- c(nsites, nrow(mtx))
+    # Motif name
+    write_lines(paste0("MOTIF ", motif_name), file = fileNameOut1, append = TRUE)
+    # Header line
+    header_line <- paste0("letter-probability matrix: alength= 4 w= ", nrow(mtx), " nsites= 20 E= 0")
+    write_lines(header_line, file = fileNameOut1, append = TRUE)
+    # Make first numerical column empty
+    mtx$Pos <- ""
+    # Write actual data
+    write_tsv(mtx, file = fileNameOut1, col_names = FALSE, append = TRUE)
+    # Separate from the next motif
+    write_lines("", file = fileNameOut1, append = TRUE)
+  }
+}
+print("Number of motifs")
+print(length(nsites))
+print("Summary of motif widths")
+summary(nsites)
+# Test if can be imported
+motifs_all <- importMatrix(fileNameOut1)
+if (!file.exists(fileNameOut2)) {
+  dev.off()
+  svg(fileNameOut2, width = 8, height = 2.5)
+  motifStack(motifs_all, layout="tree") 
+  dev.off()
+}
+```

### 2022-06-03 manuscript/manuscript.Rmd
More writeup

@@
 
-The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiprion regulator that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux].
+The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiption regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux].
 Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
 CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs].
-Point mutation and loss of heterozygosity of CTCF is associated with human cancer, identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].
+Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg][@Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].
 
-Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].
-
- with binding motif length varying between 20 and 40bp
+<!-- Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].
+binding motif length varying between 20 and 40bp -->
 
@@
 This motif interacts with the N-terminal zinc fingers for the majority of CTCF–DNA-binding events.
-Subsequently, a shorter motif (M2) [@Schmidt:2012uk], interacting with the C-terminal fingers.
+Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered, that, together with M1, form CTCF binding sites spanning 34-35bp [@Schmidt:2012uk].
 The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al.
@@
 Subsequently, $\sim80%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences and ZFs 9-11 associate with a second conserved motif [@Nakahashi:2013uw][@Pugacheva:2020wz].
-Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux][@Nakahashi:2013uw][@Wang:2012ty].
 
-CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa].
+<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa].
 Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa][@Pugacheva:2020wz].
-To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops
+To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops -->
 
-The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr].
-Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains.
-BORIS became progressively specialized to testis-specific expression throughout amniote evolution.
-CTCFL is normally expressed only in germ cells but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv].
-CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv].
-CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].
-
-We aimed to collect general-purpose CTCF binding sites for human and mouse genomes, the top two species by the number of experiments in GEO.
 CTCF binding profiles between different tissues exhibit substantial concordance [@Kim:2007ux].
 For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].
+Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux][@Nakahashi:2013uw][@Wang:2012ty].
+We aimed to provide uniformly detected CTCF binding sites for human and mouse genomes, the top two species by the number of experiments in GEO.
+Give the differences in genome assemblies, we defined CTCF binding sites directly for each assembly, including the recently created Telomere to Telomere human genome assembly (T2T) [@Nurk:2022up].
+Given the importance of CTCF binding in convergent orientation for forming chromatin loops, we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding along with a Python script that quantifies the proportion of loops with various patterns of CTCF binding.
+We additionally include selected experimentally obtained CTCF binding sites with the goal to provide a comprehensive general-purpose collection of CTCF binding profiles.
 
@@
 
-CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] was used for motif scanning using default parameters.
+CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] using default parameters.
 Softmasked genome sequences were downloaded from the USCS Genome Browser database with the exception of T2T genome (chm13v2.0) downloaded from the https://github.com/marbl/CHM13 repository.
@@
 As JASPAR, CTCFBSDB, and CIS-BP databases provide several CTCF motifs, all database-specific motifs were used to scan the genomes; consequently, the detected CTCF sites may overlap due to motif similarity.
+We provide data as GenomicRanges objects and BED files with "seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence" columns.
+
 Predefined CTCF binding sites were obtained from the SCREEN ENCODE and CTCFBSDB databases.
@@
 CTCF binding specificities in a form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases.
-The JASPAR 2022 [@Castro-Mondragon:2022uv] database list one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table`).
+The JASPAR 2022 [@Castro-Mondragon:2022uv] database lists one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table 1`).
 The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure_clustered_main_PWMs`).
-Although human and mouse CTCF motifs are considered highly conserved [@Xie:2007wm], the HOCOMOCO database provides mouse-specific CTCF PWM which is expectedly similar to human (`Figure_clustered_main_PWMs`).
+Although human and mouse CTCF motifs are considered to be highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM which is expectedly similar to human.
 SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs.
 The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the 17bp PWM identified in Jolma et al.
-study [@Jolma:2013vd] (`Figure_clustered_main_PWMs`).
+study [@Jolma:2013vd].
 The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al.
@@
 [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al.
-[@Xie:2007wm] (`Figure_clustered_CTCFBSDB_PWMs.svg`).
-The Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), likely reflecting the variability of CTCF motifs without considering the M1+M2 combinatorial structure (`Table`).
+[@Xie:2007wm].
+The Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Table 1`, `CIS-BP_2.00_Homo_sapiens`).
+The number of human-specific CTCF PWMs likely reflect the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure.
 These PWMs were used to scan human and mouse genomes for CTCF binding motifs.
 
+We used the FIMO tool from the MEME suite [@Bailey:2009aa] as one of the oldest and most well-known tool for motif scanning.
+The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj].
+The default p-value cutoff used to generate the package's data is 1E-4.
+However, we observed that up to 98% of motifs have detection p-value more than 1E-6 (`Figure 2A`), suggesting the majority of CTCF motifs have relatively poor match.
+Filtering results at 1E-6 allows to focus on more reliably detected CTCF sites.
+Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S1`).
+Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering reduces the number of CTCF sites making them comparable with previously reported 20,000-50,000 number of CTCF binding sites.
+Although we provide FIMO results as-is, we recommend filtering and reducing them.
+
 Several databases and papers provide predicted and experimental CTCF binding information.
-The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq].
+The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatory element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq] (`Table 2`).
 The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%).
 The mouse set contained 82,777 records (median size 284bp) similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites.
-A limitation of this data is that it lacks strand information.
+A limitation of this data is that it lacks strand information limiting the opportunity to use this data for CTCF orientation-specific analysis.
 We added human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.
@@
 The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes.
-The database has not been updated from 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table`).
-Coordinate conversion between genome assemblies is a well-established practice; consequently, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover versions.
+The database has not been updated from 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`).
+As coordinate conversion between genome assemblies (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts.
 Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure_liftOverJaccard.svg`).
@@
 Our goal was to provide easy programmatic access to CTCF binding data that is generally applicable for any cell or tissue type.
-For those interested in cell-type-specific CTCF binding data, many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, provide download and/or REST API access to it.
-Bioconductor's AnnotationHub itself contains human and mouse CTCF binding data from ENCODE.
-Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data.
-Given the use of cell-type-specific data is specialized, we aimed to provide broadly applicable CTCF binding data.
+However, despite high conservation of CTCF binding across cell types, cell-type-specific CTCF binding may be of interest.
+Many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, and individual publications [@Oti:2016wq][@Chang:2021wl][@Keilwagen:2019uv] provide download and/or REST API access to cell-type-specific data.<!--Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data.--> Within R ecosystem, Bioconductor's AnnotationHub contains human and mouse CTCF binding data from ENCODE.
+Given the use of cell-type-specific data is specialized and the data are frequently genome assembly-specific, we did not include cell-type-specific data in our package.
 
@@
 The PWMscan genomewide motif scanning web tool [@Ambrosini:2018wv] uses its own matrix_scan C program.
-PWMscan has the default background nucleotide composition A=0.29, C=0.21, G=0.21, T=0.29 instead of the FIMO's default A=0.25, C=0.25, G=0.25, T=0.25, and uses a 1E-5 p-value threshold instead of 1E-4 for FIMO.
+Furthermore, PWMscan has the default background nucleotide composition A=0.29, C=0.21, G=0.21, T=0.29 instead of the FIMO's default A=0.25, C=0.25, G=0.25, T=0.25, and uses a 1E-5 p-value threshold instead of 1E-4 for FIMO.
 PWMscan performed extensive benchmarking of its scanning tool and showed over 99% agreement with FIMO, HOMER, and other motif scanning tools.
 Our data is includes results using FIMO tool and we expect these results to be generally consistent when using other tools.
+
+<!-- The ENCODE project provides CTCF motif models derived from human ChIP-seq data using curated motif models from literature (JASPAR MA0139.1 and jolma_full_M2 PWMs) and de novo motif models discovered from five motif discovery tools (10 PWMs) [@Kheradpour:2014we] (http://compbio.mit.edu/encode-motifs/.
+We did not use these PWMs because of differences in tool-specific format, cell-type-specificity of de novo PWMs, and inconsistencies of curated motif models to current databases.
+-->
+
+The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr].
+CTCFL became progressively specialized to testis-specific expression throughout amniote evolution and is considered to normally being expressed only in germ cells.<!--but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv]--> Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains.
+CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv].
+Because of CTCFL's tissue specific expression and distinct functionality, we did not include its binding information.
+<!--CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].-->
 

### 2022-06-04 Logos.Rmd
Logo cloud for CIS-BP human

@@
   # Make motif name
-  motif_name <- basename(file) %>% str_remove(., ".txt")
+  motif_name <- basename(file) %>% str_remove(., ".txt") %>% str_replace(., ".", "_")
   mtx <- read_tsv(file, col_types = c("ccccc"))
@@
   dev.off()
-  svg(fileNameOut2, width = 8, height = 78)
-  motifStack(motifs_all, layout="tree") 
+  svg(fileNameOut2, width = 8, height = 4)
+  # motifStack(motifs_all, layout="tree") 
+  ## use matalign to calculate the distances of motifs
+  hc <- clusterMotifs(motifs_all)
+  ## convert the hclust to phylog object
+  library(ade4)
+  phylog1 <- ade4::hclust2phylog(hc)
+  ## reorder the motifs_all by the order of hclust
+  leaves <- names(phylog1$leaves)
+  motifs_all <- motifs_all[leaves]
+  ## extract the motif signatures
+  motifSig <- motifSignature(motifs_all, phylog1, cutoffPval=0.0001, min.freq=1)
+  ## draw the motifs with a tag-cloud style.
+  motifCloud(motifSig, scale=c(1, 0.55), rot.per = 0,
+             layout="rectangles", 
+             freq.col = "black")
   dev.off()
@@
 
-```{r}
+```{r eval=FALSE}
 # Modify input file

### 2022-06-04 manuscript/manuscript.Rmd
--max-stored-scores 1000000

@@
 
-CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] using default parameters.
+CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] using default parameters and the `--max-stored-scores 1000000` setting to avoid automatic drop of less significant motifs in case of memory limits.
 Softmasked genome sequences were downloaded from the USCS Genome Browser database with the exception of T2T genome (chm13v2.0) downloaded from the https://github.com/marbl/CHM13 repository.
@@
 Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S1`).
-Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering reduces the number of CTCF sites making them comparable with previously reported 20,000-50,000 number of CTCF binding sites.
+Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable with previously reported 20,000-50,000 sites.
 Although we provide FIMO results as-is, we recommend filtering and reducing them.

### 2022-06-06 EDA_CTCFBSDB.Rmd
Quote objects to save

@@
 # Save as RData object. subdir is the character name of the CTCF GRanges variable
-save(list = hg18.CTCFBSDB.CTCF_predicted_human, file = fileNameOut1)
-save(list = mm8.CTCFBSDB.CTCF_predicted_mouse, file = fileNameOut2)
+save(list = "hg18.CTCFBSDB.CTCF_predicted_human", file = fileNameOut1)
+save(list = "mm8.CTCFBSDB.CTCF_predicted_mouse", file = fileNameOut2)
 # load(file = paste0(subdir, ".RData"))

### 2022-06-06 EDA_liftOver_mm.Rmd
mm liftOver

--- 2022-05-29 EDA_liftOver.Rmd
@@
 # Input files
-fileNameIn1 <- file.path(dir_data, "sample_annotation.csv")
+fileNameIn1 <- file.path(dir_data, "sample_annotation_mm.csv")
 # Output files
-fileNameOut1 <- file.path(dir_data, "Figure_liftOverJaccard.svg")
+fileNameOut1 <- file.path(dir_data, "Figure_liftOverJaccard_mm.svg")
 ```
@@
          clustering_distance_cols = clust_method, 
-         annotation_row = mydf, annotation_colors = list(Assembly = c(hg18 = mycols[2], hg19 = mycols[4], hg38 = mycols[5], T2T = mycols[1])),
+         annotation_row = mydf, annotation_colors = list(Assembly = c(mm9 = mycols[2], mm10 = mycols[4])),
          treeheight_row = 40,

### 2022-06-06 FIMO_EDA.Rmd
SVG save code

@@
       # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
-      plot.title = element_text(hjust = 0.5)) +
+      plot.title = element_text(hjust = 0.5),
+      legend.text = element_text(size = 6)) +
+      # legend.key.height = unit(0.2, "mm"),
+      # legend.key.width = unit(5, "mm")) +
   ylab("Density")
@@
 fileNameOut1 <- file.path(dir_data, "Figure_human_pvalues.svg")
-ggsave(filename = fileNameOut1, width = 7, height = 3)
+
+dev.off()
+svg(filename = fileNameOut1, width = 7, height = 3)
+print(p)
+dev.off()
 ```
@@
 
-ggplot(pvals, aes(x = `-log10(p-value)`, color = MotifDB)) +
+p <- ggplot(pvals, aes(x = `-log10(p-value)`, color = MotifDB)) +
   geom_density() +
@@
 fileNameOut1 <- file.path(dir_data, "Figure_mouse_pvalues.svg")
-ggsave(filename = fileNameOut1, width = 7, height = 3)
+# ggsave(filename = fileNameOut1, width = 7, height = 3)
+dev.off()
+svg(filename = fileNameOut1, width = 7, height = 3)
+print(p)
+dev.off()
 ```

### 2022-06-06 FIMO_processing.Rmd
Process mouse liftOver data
Additional summary of mm MA0139 data

@@
 
-# Human data hg18
+# Human data liftOver
 
@@
     combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_human)
+  }
+}
+```
+
+# Mouse data liftOver
+
+```{r hg18}
+for (gen in c(genome_mouse)) {
+  for (mot in "MA0139.1") {
+    print(paste(gen, paste0("JASPAR2022_", mot), sep = "."))
+    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_mouse)
   }
@@
 pval_threshold <- 1e-6
-files <- list.files(dir_data, pattern = "*.bed", full.names = TRUE)
+files <- list.files(dir_data, pattern = "*MA0139.1.bed", full.names = TRUE)
+files <- files[grepl("mm", files)]
 for (fil in files) {

### 2022-06-06 manuscript/manuscript.Rmd
First full draft

--- 2022-06-03 manuscript/manuscript.Rmd
@@
 bibliography: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/references/references_MD.bib
+always_allow_html: true
 ---
 
-Authors
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+# Set up the environment
+library(knitr)
+opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700,
+library(pander)
+panderOptions('table.split.table', Inf)
+set.seed(1)
+# library(kableExtra)
+library(readxl)
+dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
+```
+
+# CTCF: A Bioconductor data package providing genomic coordinates of human and mouse CTCF binding sites
+
+# Authors (alphabetical order)
+
+Douglas Phanstiel  
+Eric Davis  
+Michael Love  
+Mikhail Dozmorov  
+Stuart Lee  
+Tim Triche  
+Wancen Mu
 
 # Abstract
+
+CTCF (CCCTC-binding factor) is considered the master regulator of the genome's structure and function.
+Numerous studies investigated the anatomy of CTCF binding, revealing the diversity of CTCF motifs.
+This diversity leads to the lack of a unified source of CTCF binding data.
+We collected position weight matrices of CTCF binding motifs and defined strand-oriented CTCF binding sites in human and mouse genome assemblies, including the Telomere to Telomere human genome.
+We also included selected experimentally determined and predicted CTCF binding sites, such as CTCF-bound cis-regulatory element from SCREEN ENCODE.
+We recommend filtering less significant CTCF binding motifs and demonstrate that liftOver is a viable option to convert CTCF coordinates between genome assemblies.
+Our data is assembled in the CTCF Bioconductor package (https://bioconductor.org/packages/CTCF/) and is available on AnnotationHub.
 
@@
 
-The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiption regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux].
-Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy].
+The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux].
+Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy].
 CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs].
-Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg][@Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].
+Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg; @Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].
 
-<!-- Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].
+<!-- Most CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].
 binding motif length varying between 20 and 40bp -->
 
-CTCF binds to different DNA sequences through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Filippova:1996tv][@Ohlsson:2001wu][@Jolma:2013vd].
-A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1.
-This motif interacts with the N-terminal zinc fingers for the majority of CTCF–DNA-binding events.
-Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered, that, together with M1, form CTCF binding sites spanning 34-35bp [@Schmidt:2012uk].
+CTCF binds to different DNA sequences through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Filippova:1996tv; @Ohlsson:2001wu; @Jolma:2013vd].
+A 15-20bp CTCF consensus motif from the ChIP-seq data analysis has been defined [@Kim:2007ux], referred to hereafter as M1.
+This motif interacts with the N-terminal zinc fingers for most CTCF–DNA-binding events.
+Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered that, together with M1, forms CTCF binding sites spanning 34-35bp [@Schmidt:2012uk].
 The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al.
-[@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding.
-Subsequently, $\sim80%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences and ZFs 9-11 associate with a second conserved motif [@Nakahashi:2013uw][@Pugacheva:2020wz].
+[@Xie:2007wm], detecting LM3, LM7, and LM23 motifs as parts of CTCF binding.
+Subsequently, $\sim80\%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences, and ZFs 9-11 were associated with a second conserved motif [@Nakahashi:2013uw; @Pugacheva:2020wz].
+Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux; @Nakahashi:2013uw; @Wang:2012ty].
 
-<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa].
-Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa][@Pugacheva:2020wz].
+<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa; @Oti:2016aa; @Vietri-Rudan:2015aa].
+Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa; @Fudenberg:2016aa; @Goloborodko:2016vz; @Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv; @Rao:2014aa; @Vietri-Rudan:2015tu; @Beagan:2020wz; @Stilianoudakis:2020aa; @Pugacheva:2020wz].
 To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops -->
@@
 For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].
-Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux][@Nakahashi:2013uw][@Wang:2012ty].
 We aimed to provide uniformly detected CTCF binding sites for human and mouse genomes, the top two species by the number of experiments in GEO.
-Give the differences in genome assemblies, we defined CTCF binding sites directly for each assembly, including the recently created Telomere to Telomere human genome assembly (T2T) [@Nurk:2022up].
-Given the importance of CTCF binding in convergent orientation for forming chromatin loops, we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding along with a Python script that quantifies the proportion of loops with various patterns of CTCF binding.
-We additionally include selected experimentally obtained CTCF binding sites with the goal to provide a comprehensive general-purpose collection of CTCF binding profiles.
-
-# Methods
-
-MEME motif databases (updated 20 Mar 2022) containing HOCOMOCO, CIS-BP, JASPAR2022, Jolma 2013, and SwissRegulon motifs in MEME format were downloaded from https://meme-suite.org/meme/meme-software/Databases/motifs/motif_databases.12.23.tgz.
-Motifs from the CTCFBSDB database were downloaded from https://insulatordb.uthsc.edu/download/CTCFBSDB_PWM.mat and converted to MEME format using `transfac2meme -use_acc CTCFBSDB_PWM.mat > CTCFBSDB_PWM.meme`.
-Human and mouse motifs from CIS-BP were downloaded via the bulk download interface, combined and converted to the MEME format using a custom script.
-Motifs were visualized and clustered using the `motifStack` v.1.38.0 R package.
-Jolma 2013 and CTCFBSDB EMBL_M1 motifs were converted to their reverse complements to match orientation of other motifs (`universalmotif` v.1.24.4).
-
-CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] using default parameters.
-Softmasked genome sequences were downloaded from the USCS Genome Browser database with the exception of T2T genome (chm13v2.0) downloaded from the https://github.com/marbl/CHM13 repository.
-Motifs were scanned separately for each chromosome (autosomal and sex chromosomes only), and the results were combined.
-We used human- and mouse-specific matrices for the corresponding genomes (HOCOMOCO, CIS-BP).
-As JASPAR, CTCFBSDB, and CIS-BP databases provide several CTCF motifs, all database-specific motifs were used to scan the genomes; consequently, the detected CTCF sites may overlap due to motif similarity.
-We provide data as GenomicRanges objects and BED files with "seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence" columns.
-
-Predefined CTCF binding sites were obtained from the SCREEN ENCODE and CTCFBSDB databases.
-We did not include CTCFBSDB data compiled from public experimental data because 1) mouse data was lacking, 2) hg18 genome assembly limits usability, and 3) region size was highly variable, ranging from 2bp to >350Kb.
-
-To investigate the effect of coordinate conversion, we used `liftOver` v.377 on CTCF binding sites (JASPAR 2022 MA0139.1) aligned to hg18, hg19, hg38, and T2T (NCBI GCA_009914755.4 assembly) genomes.
-The conversion chains were obtained from the UCSC Genome Browser database except for the T2T genome providing chains at the https://github.com/marbl/CHM13 repository.
-We performed all conversions except hg38->hg18 and hg18->T2T, T2T->hg18 due to lack of the corresponding chains.
-Jaccard overlap matrix between the original and liftOver CTCF binding sites was clustered using Euclidean distance and ward.D clustering method and visualized using the `pheatmap` v.1.0.12 R package.
+Given the differences in genome assemblies, we defined CTCF binding sites directly for each assembly, including the recently created Telomere to Telomere human genome assembly (T2T) [@Nurk:2022up].
+We demonstrate that coordinate conversion of CTCF binding sites between genome assemblies (liftOver) is a viable way to obtain assembly-specific CTCF binding sites.
+We further demonstrate the need for filtering less significant CTCF binding sites.
+Given the importance of CTCF binding in convergent orientation for forming chromatin loops [@Rao:2014aa; @Nanni:2020aa; @Vietri-Rudan:2015tu], we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding.
+We also include selected experimental CTCF binding sites to provide a comprehensive general-purpose collection of CTCF binding profiles.
+The CTCF R package assembles human and mouse CTCF binding data accessible in the R/Bioconductor ecosystem via AnnotationHub.
 
@@
 
-CTCF binding specificities in a form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases.
+CTCF binding specificities in the form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases.
 The JASPAR 2022 [@Castro-Mondragon:2022uv] database lists one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table 1`).
-The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure_clustered_main_PWMs`).
-Although human and mouse CTCF motifs are considered to be highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM which is expectedly similar to human.
+The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure 1A`).
+Although human and mouse CTCF motifs are considered highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM that is expectedly similar to human.
 SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs.
-The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the 17bp PWM identified in Jolma et al.
+The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the CTCF PWM identified in Jolma et al.
 study [@Jolma:2013vd].
+The Jolma et al.
+study analyzed the sequence-specific binding of human TFs using high-throughput SELEX and ChIP-sequencing and described 239 distinctly different binding specificities, including a 17bp long CTCF motif dubbed CTCF_full [@Jolma:2013vd].
 The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al.
@@
 [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al.
-[@Xie:2007wm].
-The Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Table 1`, `CIS-BP_2.00_Homo_sapiens`).
-The number of human-specific CTCF PWMs likely reflect the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure.
-These PWMs were used to scan human and mouse genomes for CTCF binding motifs.
+[@Xie:2007wm] (`Figure 1B`).
+The CIS-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Table 1`, `Supplementary Figure 1`).
+The number of human-specific CTCF PWMs likely reflects the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure.
+These PWMs were used to scan the human and mouse genomes for CTCF binding motifs.
+
+```{r}
+fileNameIn1 <- file.path(dir_data, "Figures", "Figure_1.png")
+knitr::include_graphics(fileNameIn1)
+```
+**Figure 1. CTCF motif logos of PWMs used to obtain CTCF binding sites.** PWMs from (A) MEME and (B) CTCFBSDB databases. Motifs were clustered using the `motifStack` R package.
+
+**Table 1.
+CTCF PWM information.** "Motif" - individual motif IDs or the total number of motifs per database; "Length" - motif length of the range of lengths; "URL" - direct links to motif pages; "Ref" - literature reference.
+```{r eval=FALSE}
+fileNameIn1 <- file.path(dir_data, "Tables", "Table_1_PWMs.xlsx")
+mtx <- read_xlsx(fileNameIn1)
+mtx[is.na(mtx)] <- ""
+kable(mtx)
+```
 
@@
 The default p-value cutoff used to generate the package's data is 1E-4.
-However, we observed that up to 98% of motifs have detection p-value more than 1E-6 (`Figure 2A`), suggesting the majority of CTCF motifs have relatively poor match.
-Filtering results at 1E-6 allows to focus on more reliably detected CTCF sites.
-Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S1`).
-Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering reduces the number of CTCF sites making them comparable with previously reported 20,000-50,000 number of CTCF binding sites.
+However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 2A`, `Supplementary Figure 2A`), suggesting most CTCF motifs have a relatively poor match.
+Filtering results at 1E-6 allows focusing on more reliably detected CTCF sites.
+Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Supplementary Table S1`).
+Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported 20,000-50,000 sites.
 Although we provide FIMO results as-is, we recommend filtering and reducing them.
+
+```{r}
+fileNameIn1 <- file.path(dir_data, "Figures", "Figure_2.png")
+knitr::include_graphics(fileNameIn1)
+```
+**Figure 2.
+Properties of CTCF motifs detected by FIMO.** (A) Density plot of p-value distributions of motifs detected using different motif databases in the hg38 genome assembly.
+Red dashed line - 1E-6 p-value cutoff.
+(B) Jaccard overlaps among CTCF binding sites detected in the original and liftOver genome assemblies.
+CTCF sites were detected using JASPAR 2022 MA0139.1 PWM.
+The correlogram was clustered using Euclidean distance and Ward.D clustering method.
+White-red gradient indicate low-to-high Jaccard overlaps.
+Jaccard values are shown in the corresponding cells.
+
+**Supplementary Table S1.
+Summary of CTCF binding data provided in the package.** CTCF sites for each genome assembly and PWM combination were detected using FIMO.
+"ID" - object names formatted as `<assembly>.<database name>`; "Assembly" - genome assembly, T2T - telomere to telomere (GCA_009914755.4) genome assembly; "All (p-value threshold Xe-Y)" - the total number of CTCF binding sites in the corresponding BED file at the 1E-4 or 1E-6 threshold; "Non-overlapping (p-value 1E-4 or 1E-6 threshold)" - number of non-overlapping CTCF binding sites (overlapping regions are merged) at the 1E-4 or 1E-6 threshold.
 
@@
 
+**Table 2.
+Predefined CTCF binding data.** "Database" - source of data; "Number" - number of binding sites; "Assembly" - genome assembly; "URL" - direct link to data download; "Ref" - literature reference.
+
 The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes.
-The database has not been updated from 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`).
+The database has not been updated since 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`).
 As coordinate conversion between genome assemblies (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts.
-Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure_liftOverJaccard.svg`).
-liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table_liftOver_stats.xlsx`).
+Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure 2B`).
+liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Supplementary Table S2`).
+The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Supplementary Figure 2B`).
 These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly.
-We provide hg19 and hg38 versions of predicted CTCF sites from the CTCFBSDB database.
+Using liftOver, we provide hg19/hg38 and mm9/mm10 versions of predicted CTCF sites from the CTCFBSDB database.
 
-Chang et al.
-provides CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl].
+**Supplementary Table S2.
+liftOver statistics.** "File name" - file names prefixed with genome assembly.
+Files with sites directly detected from a genome assembly have suffix ".MA0139.1.FIMO".
+Sites that can't be mapped are outputted into the "Unmapped." files.
+"Number of CTCF sites" - CTCF binding site counts.
+
+<!-- Chang et al.
+provide CTCF ChIP-seq peaks from mouse ESCs [@Chang:2021wl].
 They report 174,268 peaks (83,522 unique regions) having at least one significant CTCF binding motif (mm10 genome assembly).
-It should be noted that mouse embryonic stem cells (ESCs) have thousands of CTCF-binding events that cannot be conserved in the human genome, as they are found in rodent- specific B2 repeat elements [@Bourque:2008ws].
-
-Jolma et al.
-[@Jolma:2013vd] analyzed the sequence-specific binding of human TFs using high-throughput SELEX and ChIP sequencing.
-This study defined 830 binding profiles were obtained, describing 239 distinctly different binding specificities.
-They defined a 17bp long CTCF motif dubbed CTCF_full.
-
-How to insert single [@Dozmorov:2016aa][@Glass:2016aa] or multiple references [@Dozmorov:2016aa; @Glass:2016aa].
-
-+-------+
-| ![](figures/chicken.png)      |
-| **Figure 1. How to insert a figure.** Chicken chicken chicken.     |
-+-------+
+It should be noted that mouse embryonic stem cells (ESCs) have thousands of CTCF-binding events that cannot be conserved in the human genome, as they are found in rodent-specific B2 repeat elements [@Bourque:2008ws].
+-->
 
@@
 
-Our goal was to provide easy programmatic access to CTCF binding data that is generally applicable for any cell or tissue type.
-However, despite high conservation of CTCF binding across cell types, cell-type-specific CTCF binding may be of interest.
-Many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, and individual publications [@Oti:2016wq][@Chang:2021wl][@Keilwagen:2019uv] provide download and/or REST API access to cell-type-specific data.<!--Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data.--> Within R ecosystem, Bioconductor's AnnotationHub contains human and mouse CTCF binding data from ENCODE.
+Our goal was to provide easy programmatic access to CTCF binding data applicable to any cell or tissue type.
+However, despite the high conservation of CTCF binding across cell types, cell-type-specific CTCF binding may be of interest.
+Many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, and individual publications [@Oti:2016wq; @Chang:2021wl; @Keilwagen:2019uv] provide download and/or REST API access to cell-type-specific data.<!--Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data.--> Within the R ecosystem, Bioconductor's AnnotationHub cell type-specific human and mouse CTCF binding data from ENCODE.
 Given the use of cell-type-specific data is specialized and the data are frequently genome assembly-specific, we did not include cell-type-specific data in our package.
@@
 They have their own algorithms and defaults which may lead to different results.
-The popular HOMER motif screening tool [@Heinz:2010aa] has several options, including its unique autonormalization of sequencing bias algorithm.
-The PWMscan genomewide motif scanning web tool [@Ambrosini:2018wv] uses its own matrix_scan C program.
+The popular HOMER motif screening tool [@Heinz:2010aa] has several options, including its unique autonormalization of the sequencing bias algorithm.
+The PWMscan genomewide motif scanning web tool [@Ambrosini:2018wv] uses its own matrix_scan program.
 Furthermore, PWMscan has the default background nucleotide composition A=0.29, C=0.21, G=0.21, T=0.29 instead of the FIMO's default A=0.25, C=0.25, G=0.25, T=0.25, and uses a 1E-5 p-value threshold instead of 1E-4 for FIMO.
-PWMscan performed extensive benchmarking of its scanning tool and showed over 99% agreement with FIMO, HOMER, and other motif scanning tools.
-Our data is includes results using FIMO tool and we expect these results to be generally consistent when using other tools.
+PWMscan benchmarked its tool and showed over 99% agreement with FIMO, HOMER, and other motif scanning tools.
+Our data includes results using the FIMO tool and we expect these results to be generally consistent when using other tools.
 
@@
 The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr].
-CTCFL became progressively specialized to testis-specific expression throughout amniote evolution and is considered to normally being expressed only in germ cells.<!--but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv]--> Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains.
+CTCFL became progressively specialized to testis-specific expression throughout amniote evolution and is considered to normally be expressed only in germ cells.<!--but also aberrantly activated in numerous cancers [@Pugacheva:2010tl; @Nishana:2020tv]--> Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains.
 CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv].
-Because of CTCFL's tissue specific expression and distinct functionality, we did not include its binding information.
-<!--CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].-->
+Because of CTCFL's tissue-specific expression and distinct functionality, we did not include its binding information.
+<!--CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv; @Sleutels:2012vd].-->
+
+# Methods
+
+MEME motif databases (updated 20 Mar 2022) containing HOCOMOCO, CIS-BP, JASPAR2022, Jolma 2013, and SwissRegulon motifs in MEME format were downloaded from https://meme-suite.org/meme/meme-software/Databases/motifs/motif_databases.12.23.tgz.
+Motifs from the CTCFBSDB database were downloaded from https://insulatordb.uthsc.edu/download/CTCFBSDB_PWM.mat and converted to MEME format using `transfac2meme -use_acc CTCFBSDB_PWM.mat > CTCFBSDB_PWM.meme`.
+Human and mouse motifs from CIS-BP were downloaded via the bulk download interface, combined and converted to the MEME format using a custom script.
+Motifs were visualized and clustered using the `motifStack` v.1.38.0 R package.
+Jolma 2013 and CTCFBSDB EMBL_M1 motifs were converted to their reverse complements to match the orientation of other motifs (`universalmotif` v.1.24.4).
+
+CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] using default parameters and the `--max-stored-scores 1000000` setting to avoid the automatic drop of less significant motifs in case of memory limits.
+Softmasked genome sequences were downloaded from the USCS Genome Browser database except for the T2T genome (chm13v2.0) from the https://github.com/marbl/CHM13 repository.
+Motifs were scanned separately for each chromosome (autosomal and sex chromosomes only), and the results were combined.
+We used human- and mouse-specific matrices for the corresponding genomes (HOCOMOCO, CIS-BP).
+As JASPAR, CTCFBSDB, and CIS-BP databases provide several CTCF motifs, all database-specific motifs were used to scan the genomes; consequently, the detected CTCF sites may overlap due to motif similarity.
+We provide data as GenomicRanges objects and BED files with "seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence" columns.
+
+To investigate the effect of coordinate conversion, we used `liftOver` v.377 on CTCF binding sites (JASPAR 2022 MA0139.1) aligned to hg18, hg19, hg38, and T2T (NCBI GCA_009914755.4 assembly) genomes.
+The conversion chains were obtained from the UCSC Genome Browser database except for the T2T genome providing chains at the https://github.com/marbl/CHM13 repository.
+We performed all conversions except hg38->hg18 and hg18->T2T, T2T->hg18 due to the lack of the corresponding chains.
+Jaccard overlap matrix between the original and liftOver CTCF binding sites was clustered using Euclidean distance and ward.D clustering method and visualized using the `pheatmap` v.1.0.12 R package.
+
+Predefined CTCF binding sites were obtained from the SCREEN ENCODE and CTCFBSDB databases.
+The hg18 and mm8 CTCF binding sites were converted to hg19/hg38 and mm9/mm10 versions using `liftOver`.
+We did not include CTCFBSDB data compiled from public experimental data because 1) mouse data was lacking, 2) hg18 genome assembly limits usability, and 3) region size was highly variable, ranging from 2bp to >350Kb.
 
@@
 
+This work was supported in part by the George and Lavinia Blick Research Fund scholarship to MD.
+
 # Tables

### 2022-07-01 CTCF_Threshold.Rmd
Add CTCF threshold EDA

---
title: "CTCF threshold analysis"
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
```

# Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
# Color palette for the heatmap, https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
col3 <- colorRampPalette(c('blue', 'white', 'red'))(20)
# col3 <- colorRampPalette(c('blue', 'gray', 'yellow'))(20)
# col3 <- colorRampPalette(c('green', 'black', 'red'))(20)
# col3 <- colorRamps::green2red(n = 20)
library(GenomicRanges)
```

# Settings

```{r settings}
# Project folder path
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"

# Input files
fileNameIn1 <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot/hg38.JASPAR2022_MA0139.1.bed"
fileNameIn2 <- file.path(dir_data, "data/GRCh38-CTCF.bed")
# Output files
fileNameOut1 <- file.path(dir_data, "")
```

# Load data

```{r data}
# Read in data
mtx <- read_tsv(fileNameIn1, col_names = FALSE, col_types = c("cddcdcdd"))
gr_meme <- GRanges(seqnames = mtx$X1, ranges = IRanges(start = mtx$X2, end = mtx$X3), strand = mtx$X6)
gr_meme$pval <- mtx$X8
gr_meme$qval <- mtx$X9

mtx1 <- read_tsv(fileNameIn2, col_names = FALSE, col_types = c("cddccc"))
gr_encode <- GRanges(seqnames = mtx1$X1, ranges = IRanges(start = mtx1$X2, end = mtx1$X3))
```

## Summary statistics

```{r}
print("Summary of full MEME p-value distribution")
summary(gr_meme$pval)

print(paste("Length of full MEME CTCF set:", length(gr_meme)))
print(paste("Length of ENCODE SCREEN CTCF set:", length(gr_encode)))
```

## Scanning the threshold

```{r}
# thresholds <- seq(from = 1E-8, to = 1E-4, length.out = 10) %>% rev()
thresholds <- c(1.00E-4, 5.00E-4, 1.00E-5, 5.00E-5, 1.00E-6, 5.00E-6, 1.00E-7, 5.00E-7, 1.00E-8)
# formatC(thresholds, format = "e")

gr_meme_pos_list <- vector(mode = "numeric", length = length(thresholds))
gr_meme_neg_list <- vector(mode = "numeric", length = length(thresholds))
for (i in 1:length(thresholds)) {
  pval_cutoff <- thresholds[i]
  gr_meme_cutoff <- gr_meme[gr_meme$pval < pval_cutoff]
  gr_meme_overlaps <- countOverlaps(gr_meme_cutoff, gr_encode)
  gr_meme_pos_list[i] <- sum(gr_meme_overlaps > 0) / length(gr_meme_overlaps)
  gr_meme_neg_list[i] <- sum(gr_meme_overlaps == 0) / length(gr_meme_overlaps)
}

gr_summary <- data.frame(Threshold = thresholds, TP = gr_meme_pos_list, FP = gr_meme_neg_list)

mtx_to_plot <- pivot_longer(gr_summary, cols = c("TP", "FP"), names_to = "Type")
ggplot(mtx_to_plot, aes(x = Threshold, y = value, color = Type)) +
  geom_line() +
  scale_x_continuous(breaks = thresholds, labels = scientific) +
  scale_y_continuous(breaks = seq(0, 1, 0.1))   + 
  scale_color_manual(values = mycols[1:2]) +
  coord_trans(x ="log10") +
  geom_vline(xintercept = 1E-6, linetype = "dashed", color = "black") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      # axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5),
      legend.text = element_text(size = 6)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  xlab("FIMO p-value threshold") +
  ylab("FP/TP rate")

fileNameOut1 <- file.path(dir_data, "results", "Figure_human_pvalues_threshold.svg")
ggsave(filename = fileNameOut1, width = 4, height = 2)
# dev.off()
# svg(filename = fileNameOut1, width = 7, height = 3)
# print(p)
# dev.off()

```

### 2022-07-01 manuscript/manuscript.Rmd
Reorder authors

@@
 
-Douglas Phanstiel  
+Mikhail Dozmorov *  
+Wancen Mu  
 Eric Davis  
-Michael Love  
-Mikhail Dozmorov  
 Stuart Lee  
 Tim Triche  
-Wancen Mu
+Douglas Phanstiel  
+Michael Love
 
@@
 CTCF PWM information.** "Motif" - individual motif IDs or the total number of motifs per database; "Length" - motif length of the range of lengths; "URL" - direct links to motif pages; "Ref" - literature reference.
-```{r eval=FALSE}
+```{r eval=TRUE}
 fileNameIn1 <- file.path(dir_data, "Tables", "Table_1_PWMs.xlsx")

### 2022-07-02 CTCF_Threshold.Rmd
Figure tweaks

@@
   scale_x_continuous(breaks = thresholds, labels = scientific) +
-  scale_y_continuous(breaks = seq(0, 1, 0.1))   + 
+  scale_y_continuous(breaks = seq(0, 1, 0.2))   + 
   scale_color_manual(values = mycols[1:2]) +
@@
 fileNameOut1 <- file.path(dir_data, "results", "Figure_human_pvalues_threshold.svg")
-ggsave(filename = fileNameOut1, width = 4, height = 2)
+ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
 # dev.off()

### 2022-07-02 FIMO_EDA.Rmd
Figure tweaks

@@
   geom_density() +
-  geom_vline(xintercept = -log10(pval_threshold), linetype = "dashed", color = "red") +
+  geom_vline(xintercept = -log10(pval_threshold), linetype = "dashed", color = "black") +
   scale_color_manual(values = mycols[1:6]) +
@@
 fileNameOut1 <- file.path(dir_data, "Figure_human_pvalues.svg")
+ggsave(filename = fileNameOut1, width = 6.4, height = 1.6)
 
-dev.off()
-svg(filename = fileNameOut1, width = 7, height = 3)
-print(p)
-dev.off()
+# dev.off()
+# svg(filename = fileNameOut1, width = 7, height = 2)
+# print(p)
+# dev.off()
+
 ```

### 2022-07-02 manuscript/manuscript.Rmd
Final rewriting incorporating CTCF threshold

@@
 
-CTCF binding specificities in the form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases.
+CTCF sequence motifs in the form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases.
 The JASPAR 2022 [@Castro-Mondragon:2022uv] database lists one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table 1`).
-The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure 1A`).
-Although human and mouse CTCF motifs are considered highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM that is expectedly similar to human.
+The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure S1A`).
+Although human and mouse CTCF motifs are considered to be highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM.
 SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs.
@@
 [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al.
-[@Xie:2007wm] (`Figure 1B`).
-The CIS-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Table 1`, `Supplementary Figure 1`).
+[@Xie:2007wm] (`Figure S1B`).
+The CIS-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Figure S1CD`).
 The number of human-specific CTCF PWMs likely reflects the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure.
-These PWMs were used to scan the human and mouse genomes for CTCF binding motifs.
-
-```{r}
-fileNameIn1 <- file.path(dir_data, "Figures", "Figure_1.png")
-knitr::include_graphics(fileNameIn1)
-```
-**Figure 1. CTCF motif logos of PWMs used to obtain CTCF binding sites.** PWMs from (A) MEME and (B) CTCFBSDB databases. Motifs were clustered using the `motifStack` R package.
+These PWMs were used to scan the human and mouse genomes for CTCF binding motifs using the FIMO tool from the MEME suite [@Bailey:2009aa] (`Table S1`).
 
@@
 
-We used the FIMO tool from the MEME suite [@Bailey:2009aa] as one of the oldest and most well-known tool for motif scanning.
-The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj].
-The default p-value cutoff used to generate the package's data is 1E-4.
-However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 2A`, `Supplementary Figure 2A`), suggesting most CTCF motifs have a relatively poor match.
-Filtering results at 1E-6 allows focusing on more reliably detected CTCF sites.
-Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Supplementary Table S1`).
-Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported 20,000-50,000 sites.
-Although we provide FIMO results as-is, we recommend filtering and reducing them.
-
-```{r}
-fileNameIn1 <- file.path(dir_data, "Figures", "Figure_2.png")
-knitr::include_graphics(fileNameIn1)
-```
-**Figure 2.
-Properties of CTCF motifs detected by FIMO.** (A) Density plot of p-value distributions of motifs detected using different motif databases in the hg38 genome assembly.
-Red dashed line - 1E-6 p-value cutoff.
-(B) Jaccard overlaps among CTCF binding sites detected in the original and liftOver genome assemblies.
-CTCF sites were detected using JASPAR 2022 MA0139.1 PWM.
-The correlogram was clustered using Euclidean distance and Ward.D clustering method.
-White-red gradient indicate low-to-high Jaccard overlaps.
-Jaccard values are shown in the corresponding cells.
-
 **Supplementary Table S1.
@@
 
-Several databases and papers provide predicted and experimental CTCF binding information.
-The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatory element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq] (`Table 2`).
+In addition to sequence-derived motifs, several databases provide predicted and experimental CTCF binding sites.
+The ENCODE Phase III has released SCREEN, a database of cis-regulatory elements (cCREs) for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq] (`Table 2`).
 The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%).
@@
 A limitation of this data is that it lacks strand information limiting the opportunity to use this data for CTCF orientation-specific analysis.
-We added human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.
+We included human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.
 
@@
 
-The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes.
+The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding sites for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes.
 The database has not been updated since 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`).
 As coordinate conversion between genome assemblies (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts.
-Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure 2B`).
-liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Supplementary Table S2`).
-The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Supplementary Figure 2B`).
+Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure 1A`).
+liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table S2`).
+The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Figure S2A`).
 These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly.
 Using liftOver, we provide hg19/hg38 and mm9/mm10 versions of predicted CTCF sites from the CTCFBSDB database.
+
+```{r}
+fileNameIn1 <- file.path(dir_data, "Figures", "Figure_1.png")
+knitr::include_graphics(fileNameIn1)
+```
+**Figure 1. CTCF motif logos of PWMs used to obtain CTCF binding sites.** PWMs from (A) MEME and (B) CTCFBSDB databases. Motifs were clustered using the `motifStack` R package.
 
@@
 "Number of CTCF sites" - CTCF binding site counts.
+
+We used the FIMO tool from the MEME suite [@Bailey:2009aa] as one of the oldest and most well-known tool for motif scanning.
+The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj].
+The default p-value cutoff is 1E-4.
+However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 1B`, `Figure S2B`), suggesting most CTCF motifs have a relatively poor match.
+Using the experimental SCREEN CTCF sites as ground truth, we quantified the proportion of true and false positives depending on the FIMO p-value cutoff.
+We found 1E-6 threshold as the most optimal, providing $\sim80\%$ of true positive CTCF motifs.
+Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S1`).
+Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported 20,000-50,000 sites.
+Although we provide FIMO results using the default settings, for downstream analyses we recommend filtering and reducing them.
 

### 2022-07-03 manuscript/manuscript_bioinformatics.Rmd
Version for Bioinformatics

---
# csl: /Volumes/GoogleDrive/My Drive/manuscript_template/styles.ref/cell-reports.csl
csl: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.ref/cell-reports.csl
output: 
  word_document:
#    reference_docx: /Volumes/GoogleDrive/My Drive/manuscript_template/styles.doc/NIH_grant_style.docx
    reference_docx: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.doc/NIH_grant_style.docx
# https://github.com/mdozmorov/manuscript_template/tree/master/references
# bibliography: /Volumes/GoogleDrive/My Drive/manuscript_template/references/references_MD.bib
bibliography: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/references/references_MD.bib
always_allow_html: true
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
# library(kableExtra)
library(readxl)
dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
```

# CTCF: A Bioconductor data package of human and mouse CTCF binding sites

Mikhail Dozmorov*, Wancen Mu, Eric Davis, Stuart Lee, Tim Triche, Douglas Phanstiel, Michael Love

# Abstract

**Summary:** CTCF (CCCTC-binding factor) is considered the master regulator of the genome's structure and function. The diversity of CTCF binding motifs leads to the lack of a unified source of CTCF binding data. We collected position weight matrices of CTCF binding motifs and defined strand-oriented CTCF binding sites in the human and mouse genome assemblies, including the Telomere to Telomere human and mm39 mouse genomes. We also included selected experimentally determined and predicted CTCF binding sites, such as CTCF-bound cis-regulatory element from SCREEN ENCODE. We recommend filtering less significant CTCF binding motifs and demonstrate that liftOver is a viable option to convert CTCF coordinates between genome assemblies. Our comprehensive data resource and usage recommendations will strengthen reproducibility of genomic studies utilizing CTCF binding data.

**Availability:** https://bioconductor.org/packages/CTCF  
**Contact:** mdozmorov@vcu.edu  
**Supplementary information:** Supplementary data are available at Bioinformatics online.

### Keywords

# Introduction

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator <!--evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux]. Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including -->involved in transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, <!--intra- and inter-chromosomal contacts -->chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy]. <!--CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs]. Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg; @Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].-->
<!-- Most CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux]. binding motif length varying between 20 and 40bp -->
CTCF binds DNA through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Ohlsson:2001wu; @Jolma:2013vd]<!--@Filippova:1996tv-->. A 15-20bp CTCF consensus motif from the ChIP-seq data analysis has been defined [@Kim:2007ux], referred to hereafter as M1. This motif interacts with the N-terminal zinc fingers for most CTCF–DNA-binding events. Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered that, together with M1, forms CTCF binding sites spanning 34-35bp [@Schmidt:2012uk]. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, and LM23 motifs as parts of CTCF binding.<!--Subsequently, $\sim80\%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences, and ZFs 9-11 were associated with a second conserved motif [@Nakahashi:2013uw; @Pugacheva:2020wz].--> Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes<!--[@Kim:2007ux; @Nakahashi:2013uw; @Wang:2012ty]-->.
<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa; @Oti:2016aa; @Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa; @Fudenberg:2016aa; @Goloborodko:2016vz; @Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv; @Rao:2014aa; @Vietri-Rudan:2015tu; @Beagan:2020wz; @Stilianoudakis:2020aa; @Pugacheva:2020wz]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops -->

CTCF binding profiles between different tissues exhibit substantial concordance [@Kim:2007ux].<!--For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].--> We aimed to provide uniformly detected strand-specific CTCF binding sites for the most studied human and mouse genomes. Given the differences in genome assemblies, we defined CTCF binding sites directly for each assembly, including the Telomere to Telomere (T2T) human genome assembly [@Nurk:2022up] and the GRCm39/mm39 mouse genome assembly. We demonstrate that coordinate conversion<!--of CTCF binding sites between genome assemblies--> (liftOver) is a viable way to obtain assembly-specific CTCF binding sites. We further demonstrate the need for filtering less significant CTCF binding sites.<!--Given the importance of CTCF binding in convergent orientation for forming chromatin loops [@Rao:2014aa; @Nanni:2020aa; @Vietri-Rudan:2015tu], we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding.--> We also include selected experimental CTCF binding sites<!--to provide a comprehensive general-purpose collection of CTCF binding profiles.--> and assemble them in the CTCF R/Bioconductor package.

# Implementation

Position Weight Matrices (PWMs) [@Stormo:2000uw] of CTCF sequence motifs from Jaspar 2022 [@Castro-Mondragon:2022uv], HOCOMOCO v11 [@Kulakovskiy:2018vw], CIS-BP [@Weirauch:2014ul], Jolma 2013 [@Jolma:2013vd], and SwissRegulon [@Pachkov:2013uk] databases were downloaded from the MEME Motif database (`Table S1`). They were largely similar with the exception of Jolma 2013 and two long (34bp and 35bp) Jaspar PWMs (`Figure S1A`). The CTCF binding site database (CTCFBSDB) [@Ziebarth:2013vj] includes PWMs for the M1+M2 motifs [@Schmidt:2012uk], the Ren_20 motif [@Kim:2007ux], and the LM2, LM7, and LM23 motifs [@Xie:2007wm] (`Figure S1B`). The CIS-BP database [@Weirauch:2014ul] lists 83 human-specific and two mouse-specific PWMs (`Figure S1CD`). These PWMs were used to scan the human and mouse genomes for CTCF binding motifs using the FIMO tool from the MEME suite [@Bailey:2009aa].

The CTCFBSDB database [@Bao:2008vq] is a dedicated resource of predicted CTCF binding sites for hg18 and mm8 genome assemblies<!-- for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes. database has not been updated since 2013 and the predictions are available for hg18 and mm8 genome assemblies only--> (`Table S2`). As coordinate conversion (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts. Expectedly, genome assembly was the primary driver of similarity (`Figure 1A`).<!--among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies. liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table S2`). The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Figure S2A`). These results --> These results were observed using mm9/mm10/mm39 genome assemblies (`Figure S2A`), indicating that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly. Using liftOver, we provide hg19/hg38 and mm9/mm10 versions of predicted CTCF sites from the CTCFBSDB database.

The FIMO tool <!--as one of the oldest and most well-known tool for motif scanning. The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj]. -->uses the default p-value cutoff 1E-4 for CTCF motif detection. However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 1B`, `Figure S2B`), suggesting most CTCF motifs have a relatively poor sequence match. Using the ENCODE SCREEN database of CTCF-bound cis-regulatory elements (cCREs) [@ENCODE-Project-Consortium:2020uq] as ground truth (`Table S2`, included in the package), we quantified the proportion of true and false positives depending on the FIMO p-value cutoff. We found 1E-6 threshold as the most optimal, providing $\sim80\%$ of true positive CTCF motifs. Additionally, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S3`). Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported. Although we provide FIMO results using the default settings, we recommend filtering and reducing them.

# Discussion

Our goal was to provide easy programmatic access to CTCF binding data applicable to any cell or tissue type. The CTCF package contains XX GRanges objects of strand-specific CTCF motifs predicted using 98 PWMs from six databases on human and mouse genome assemblies, including the T2T and mm39 assemblies (`Table S3`). Given the popularity of JASPAR database, we recommend using predictions made with the MA0139.1 PWM. If experimentally detected CTCF motifs are of interest, we suggest using CTCF-bound cCREs from the ENCODE SCREEN database. We hope the CTCF package will enable standardization and reproducibility of studies employing CTCF binding data.

# Funding

This work was supported by the George and Lavinia Blick Research Scholarship to MD.

_Conflict of Interest._ None declared.

# References

### 2022-07-03 Logos.Rmd
Adjust code

--- 2022-05-29 Logos.Rmd
@@
 ---
-title: "Analysis"
+title: "PWM processing and plotting"
 author: "Mikhail Dozmorov"
@@
 dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/"
+dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev/"
 ```
+
+# MEME database
 
@@
 motifs_all <- list()
-```
 
 # JASPAR 2022
-
-```{r fig.width = 6, fig.height = 4}
 # Input files
 fileNameIn1 <- file.path(dir_data, "motif_databases/JASPAR2022_CORE_vertebrates_non-redundant_v2.meme")
-# Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/JASPAR2022_CORE_vertebrates_non-redundant_v2.svg")
 # Read in data
 motifs <- importMatrix(fileNameIn1)
-# plot(motifs[[1]])
-motifStack(motifs)
-dev.off()
-svg(fileNameOut1, width = 6, height = 4)
-motifStack(motifs)
-dev.off()
+# plot(motifs)
 # Standardize names
@@
 motifs_all <- c(motifs_all, motifs)
-```
 
-# HOCOMOCO v11
-
-```{r fig.width = 6, fig.height = 2}
+# HOCOMOCO v11 human
 # Input files
 fileNameIn1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_HUMAN_mono_meme_format.meme")
-# Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_HUMAN_mono_meme_format.svg")
 # Read in data
 motifs <- importMatrix(fileNameIn1)
-plot(motifs[[1]])
-dev.off()
-svg(fileNameOut1, width = 6, height = 2)
-plot(motifs[[1]])
-dev.off()
+# plot(motifs)
 # Standardize names
@@
 motifs_all <- c(motifs_all, motifs)
-```
 
-```{r fig.width = 6, fig.height = 2}
+# HOCOMOCO v11 mouse
 # Input files
 fileNameIn1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_MOUSE_mono_meme_format.meme")
-# Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/HOCOMOCOv11_core_MOUSE_mono_meme_format.svg")
 # Read in data
 motifs <- importMatrix(fileNameIn1)
-plot(motifs[[1]])
-dev.off()
-svg(fileNameOut1, width = 6, height = 2)
-plot(motifs[[1]])
-dev.off()
+# plot(motifs)
 # Standardize names
@@
 motifs_all <- c(motifs_all, motifs)
-```
 
 # SwissRegulon
-
-```{r fig.width = 6, fig.height = 2}
 # Input files
 fileNameIn1 <- file.path(dir_data, "motif_databases/SwissRegulon_human_and_mouse.meme")
-# Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/SwissRegulon_human_and_mouse.svg")
 # Read in data
 motifs <- importMatrix(fileNameIn1)
-plot(motifs[[1]])
-dev.off()
-svg(fileNameOut1, width = 6, height = 2)
-plot(motifs[[1]])
-dev.off()
+# plot(motifs)
 # Standardize names
@@
 motifs_all <- c(motifs_all, motifs)
-```
 
 # Jomla 2013
-
-Needs to be reverse complement
-
-```{r fig.width = 6, fig.height = 2}
 # Input files
 fileNameIn1 <- file.path(dir_data, "motif_databases/jolma2013.meme")
-# Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/jolma2013.svg")
 # Reverse complement
@@
 motifs <- importMatrix(fileNameIn1)
-# Reverse complement
+# Needs to be reverse complement
 motifs_RC <- motif_rc(motifs[[1]])
-plot(motifs_RC)
-dev.off()
-svg(fileNameOut1, width = 6, height = 2)
-plot(motifs[[1]])
-dev.off()
+# plot(motifs_RC)
 # Save Reverse complement
+if (!file.exists(fileNameOut2)) {
   write_meme(motifs_RC, fileNameOut2)
+}
 # Standardize names
@@
 
-# Clustering
+## Clustering
 
@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/Figure_clustered_main_PWMs.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_clustered_main_PWMs.svg")
+if (!file.exists(fileNameOut1)) {
   dev.off()
@@
   dev.off()
+}
 ```
@@
 fileNameOut1 <- file.path(dir_data1, "CIS-BP_2.00_Homo_sapiens.meme")
-fileNameOut2 <- file.path(dir_data1, "CIS-BP_2.00_Homo_sapiens.svg")
+fileNameOut2 <- file.path(dir_project, "results", "Figure_CIS-BP_2.00_Homo_sapiens.svg")
 # Standard header
@@
 fileNameOut1 <- file.path(dir_data1, "CIS-BP_2.00_Mus_musculus.meme")
-fileNameOut2 <- file.path(dir_data1, "CIS-BP_2.00_Mus_musculus.svg")
+fileNameOut2 <- file.path(dir_project, "results", "Figure_CIS-BP_2.00_Mus_musculus.svg")
 # Standard header
@@
 
-EMBL_M1 needs to be reverse complement
-
 ```{r}
@@
 motifs <- importMatrix(fileNameIn1)
-# Correct EMBL_M1
+# EMBL_M1 needs to be reverse complement
 motifs[[1]] <- motif_rc(motifs[[1]])
 # Save the corrected matrix
+if (!file.exists(fileNameOut2)) {
   write_meme(motifs, fileNameOut1)
+}
 # Plot
@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "motif_databases/Figure_clustered_CTCFBSDB_PWMs.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_clustered_CTCFBSDB_PWMs.svg")
 dev.off()
@@
 ```
+
+The end
+
+```{r eval=FALSE}
+# ENCODE
+# Modify input dir
+dir_data1 <- file.path(dir_data, "ENCODE_motifs_CTCF")
+files <- list.files(dir_data1, pattern = "*.txt", full.names = TRUE)
+# Output file
+fileNameOut1 <- file.path(dir_data1, "motifs_CTCF.meme")
+fileNameOut2 <- file.path(dir_data1, "motifs_CTCF.svg")
+# Standard header
+write_lines("MEME version 5.4.1 (Tue Mar 1 19:18:48 2022 -0800)", file = fileNameOut1)
+write_lines("ALPHABET= ACGT", file = fileNameOut1, append = TRUE)
+write_lines("strands: + -", file = fileNameOut1, append = TRUE)
+write_lines("Background letter frequencies (from uniform background):", file = fileNameOut1, append = TRUE)
+write_lines("A 0.25000 C 0.25000 G 0.25000 T 0.25000 ", file = fileNameOut1, append = TRUE)
+# Store number of bases
+nsites <- c()
+for(i in 1:length(files)) {
+  fileNameIn=files[i]
+  mtx <- read.table(fileNameIn, sep = " ", fill = TRUE)
+  # Make motif name
+  motif_name <- strsplit(mtx[1, 1], "\t")[[1]]
+  motif_id <- motif_name[1] %>% str_remove(">")
+  motif_alt <- motif_name[length(motif_name)]
+  # Drop header row
+  mtx <- mtx[2:nrow(mtx), ]
+  # If non-empty, proceed
+  if (nrow(mtx) > 0) {
+    # Collect number of bases
+    nsites <- c(nsites, nrow(mtx))
+    # Motif name
+    write_lines(paste0("MOTIF ", paste(motif_id, motif_alt)), file = fileNameOut1, append = TRUE)
+    # Header line
+    header_line <- paste0("letter-probability matrix: alength= 4 w= ", nrow(mtx), " nsites= 20 E= 0")
+    write_lines(header_line, file = fileNameOut1, append = TRUE)
+    # Make first numerical column empty
+    mtx[, 1] <- ""
+    # Write actual data
+    write_tsv(mtx, file = fileNameOut1, col_names = FALSE, append = TRUE)
+    # Separate from the next motif
+    write_lines("", file = fileNameOut1, append = TRUE)
+  }
+}
+print("Number of motifs")
+print(length(nsites))
+print("Summary of motif widths")
+summary(nsites)
+# Test if can be imported
+motifs_all <- importMatrix(fileNameOut1)
+if (!file.exists(fileNameOut2)) {
+  dev.off()
+  svg(fileNameOut2, width = 10, height = 14)
+  motifStack(motifs_all, layout="tree") 
+  dev.off()
+  }
+```
+
+```{r eval=FALSE}
+# Modify input file
+fileNameIn1 <- file.path(dir_data, "motifs_CTCF.txt")
+# Output file
+fileNameOut1 <- file.path(dir_data1, "motifs_CTCF.meme")
+fileNameOut2 <- file.path(dir_data1, "motifs_CTCF.svg")
+# Standard header
+write_lines("MEME version 5.4.1 (Tue Mar 1 19:18:48 2022 -0800)", file = fileNameOut1)
+write_lines("ALPHABET= ACGT", file = fileNameOut1, append = TRUE)
+write_lines("strands: + -", file = fileNameOut1, append = TRUE)
+write_lines("Background letter frequencies (from uniform background):", file = fileNameOut1, append = TRUE)
+write_lines("A 0.25000 C 0.25000 G 0.25000 T 0.25000 ", file = fileNameOut1, append = TRUE)
+# Store number of bases
+nsites <- c()
+for(i in 1:length(files)) {
+  file=files[i]
+  mtx <- read.table(fileNameIn1, sep = " ", fill = TRUE)
+  # Make motif name
+  motif_name <- strsplit(mtx[1, 1], "\t")[[1]]
+  motif_id <- motif_name[1] %>% str_remove(">")
+  motif_alt <- motif_name[length(motif_name)]
+  # If non-empty, proceed
+  if (nrow(mtx) > 0) {
+    # Collect number of bases
+    nsites <- c(nsites, nrow(mtx))
+    # Motif name
+    write_lines(paste0("MOTIF ", motif_name), file = fileNameOut1, append = TRUE)
+    # Header line
+    header_line <- paste0("letter-probability matrix: alength= 4 w= ", nrow(mtx), " nsites= 20 E= 0")
+    write_lines(header_line, file = fileNameOut1, append = TRUE)
+    # Make first numerical column empty
+    mtx$Pos <- ""
+    # Write actual data
+    write_tsv(mtx, file = fileNameOut1, col_names = FALSE, append = TRUE)
+    # Separate from the next motif
+    write_lines("", file = fileNameOut1, append = TRUE)
+  }
+}
+print("Number of motifs")
+print(length(nsites))
+print("Summary of motif widths")
+summary(nsites)
+# Test if can be imported
+motifs_all <- importMatrix(fileNameOut1)
+if (!file.exists(fileNameOut2)) {
+  dev.off()
+  svg(fileNameOut2, width = 8, height = 2.5)
+  motifStack(motifs_all, layout="tree") 
+  dev.off()
+}
+```

### 2022-07-04 CTCF_Threshold_mm.Rmd
Latest changes

--- 2022-07-02 CTCF_Threshold.Rmd
@@
 ```{r settings}
+dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev"
 # Project folder path
-dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
+dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
 
 # Input files
-fileNameIn1 <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot/hg38.JASPAR2022_MA0139.1.bed"
-fileNameIn2 <- file.path(dir_data, "data/GRCh38-CTCF.bed")
+fileNameIn1 <- file.path(dir_project, "RData", "mm10.MA0139.1.bed")
+fileNameIn2 <- file.path(dir_project, "data", "mm10-CTCF.bed")
 # Output files
-fileNameOut1 <- file.path(dir_data, "")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_mouse_pvalues_threshold.svg")
 ```
@@
 
-fileNameOut1 <- file.path(dir_data, "results", "Figure_human_pvalues_threshold.svg")
 ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)

### 2022-07-04 EDA_SCREEN.Rmd
Latest changes

--- 2022-05-02 EDA_SCREEN.Rmd
@@
 # Project folder path
-dir_data <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
+dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
 
@@
 # Human CTCF-bound cCRE (hg38)
-fileNameIn1 <- file.path(dir_data, "data/GRCh38-CTCF.bed")
-# download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/GRCh38-CTCF.bed", fileNameIn1)
+fileNameIn1 <- file.path(dir_project, "data/GRCh38-CTCF.bed")
+if (!file.exists(fileNameIn1)) {
+  download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/GRCh38-CTCF.bed", fileNameIn1)
+}
 # Mouse CTCF-bound cCRE (mm10)
-fileNameIn2 <- file.path(dir_data, "data/mm10-CTCF.bed")
-# download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/mm10-CTCF.bed", fileNameIn2)
+fileNameIn2 <- file.path(dir_project, "data/mm10-CTCF.bed")
+if (!file.exists(fileNameIn2)) {
+  download.file("https://api.wenglab.org/screen_v13/fdownloads/cCREs/mm10-CTCF.bed", fileNameIn2)
+}
 # Output files
-fileNameOut1 <- file.path(dir_data, "results/CTCF_hg38_SCREEN.RData")
-fileNameOut2 <- file.path(dir_data, "results/CTCF_mm10_SCREEN.RData")
+fileNameOut1 <- file.path(dir_project, "RData", "hg38.SCREEN.GRCh38_CTCF.RData")
+fileNameOut1.1 <- file.path(dir_project, "RData", "hg38.SCREEN.GRCh38_CTCF.bed")
+fileNameOut2 <- file.path(dir_project, "RData", "mm10.SCREEN.mm10_CTCF.RData")
+fileNameOut2.1 <- file.path(dir_project, "RData", "mm10.SCREEN.mm10_CTCF.bed")
 ```
@@
 }
+```
+
+```{r}
+hg38.SCREEN.GRCh38_CTCF <- mtx_to_gr(mtx = mtx_human, genome_id = "hg38")
+length(hg38.SCREEN.GRCh38_CTCF)
+summary(width(hg38.SCREEN.GRCh38_CTCF))
+mm10.SCREEN.mm10_CTCF <- mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10")
+length(mm10.SCREEN.mm10_CTCF)
+summary(width(mm10.SCREEN.mm10_CTCF))
 
 # Save as RData object. subdir is the character name of the CTCF GRanges variable
-save(list = mtx_to_gr(mtx = mtx_human, genome_id = "hg38"), file = fileNameOut1)
-save(list = mtx_to_gr(mtx = mtx_mouse, genome_id = "mm10"), file = fileNameOut2)
+save(list = "hg38.SCREEN.GRCh38_CTCF", file = fileNameOut1)
+save(list = "mm10.SCREEN.mm10_CTCF", file = fileNameOut2)
 # load(file = paste0(subdir, ".RData"))
+# Save BED files
+export.bed(hg38.SCREEN.GRCh38_CTCF, fileNameOut1.1)
+export.bed(mm10.SCREEN.mm10_CTCF, fileNameOut2.1)
 ```

### 2022-07-04 EDA_liftOver.Rmd
Latest changes

@@
 dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/CTCF_liftover/"
-
+dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
 # Input files
@@
 # Output files
-fileNameOut1 <- file.path(dir_data, "Figure_liftOverJaccard.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_liftOverJaccard.svg")
 ```
@@
   # Read data
-  bed_file <- import(file.path(dir_data, "data", files[i]), format = "bed") # , genome = genome_assembly[i], colnames = c("chrom", "start", "end", "name", "score", "strand", "motif", "pval", "qval"), extraCols = extraCols_names
+  # bed_file <- import(file.path(dir_data, "data", files[i]), format = "bed", genome = genome_assembly[i], colnames = c("chrom", "start", "end", "name", "score", "strand", "motif", "pval", "qval"), extraCols = extraCols_names)
+  mtx <- read_tsv(file.path(dir_data, "data", files[i]), col_names = FALSE)
+  bed_file <- GRanges(seqnames = mtx$X1, ranges = IRanges(start = mtx$X2, end = mtx$X3), strand = mtx$X6)
   # Save all data

### 2022-07-04 FIMO_processing.Rmd
Latest changes

--- 2022-06-03 FIMO_processing.Rmd
@@
 dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot"
+dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev/"
+# Rename motif folders to replace "-" by "_"
+# for file in `find . -type d -name "*-*"`; do newfile=`echo $file | sed 's/-/_/g'`; mv $file $newfile; done
+# Capitalize Jolma
+# for file in `find . -type d -name "jolma*"`; do newfile=`echo $file | sed 's/j/J/'`; mv $file $newfile; done
+# Remove "_corrected"
+# for file in `find . -type d -name "*_corrected*"`; do newfile=`echo $file | sed 's/_corrected//'`; mv $file $newfile; done
 # Motifs
-motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
-motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
+motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "Jolma2013", "SwissRegulon_human_and_mouse")
+motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "Jolma2013", "SwissRegulon_human_and_mouse")
 # Genomes
 genome_human <- c("T2T", "hg38", "hg19")
-genome_mouse <- c("mm10", "mm9")
+genome_mouse <- c("mm10", "mm9", "mm39")
 # Chromosomes
@@
 ```{r data}
+# Function to combine database- and chromosome-specific FIMO results for a given genome assembly
 combine_fimo <- function(motif = "SwissRegulon_human_and_mouse", genome_id = "T2T", chromosomes = chromosome_human) {
@@
     # print(chrom)
-    fileNameIn <- file.path(dir_data, paste(motif, genome_id, chrom, sep = "_"), "fimo.txt.gz")
+    fileNameIn <- file.path(dir_data, "results", paste(motif, genome_id, chrom, sep = "_"), "fimo.txt.gz")
     # Read "fimo.bed" created by "fimo.qsub"
@@
 
-  ctcfGR_to_save <- paste(genome_id, motif, "FIMO", sep = ".")
-  fileNameOut1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".RData"))
-  fileNameOut1.1 <- file.path(dir_data, paste0(ctcfGR_to_save, ".bed"))
+  ctcfGR_to_save <- paste(genome_id, motif, sep = ".")
+  fileNameOut1 <- file.path(dir_project, "RData", paste0(ctcfGR_to_save, ".RData"))
+  fileNameOut1.1 <- file.path(dir_project, "RData", paste0(ctcfGR_to_save, ".bed"))
   # Assign this object to the subfolder-specific variable name
@@
 
-# Human data hg18
+# Human data for liftOver
 
-```{r hg18}
+```{r hg}
 for (gen in c("hg18", genome_human)) {
-  for (mot in "JASPAR2022_MA0139.1") {
+  for (mot in "MA0139.1") {
     print(paste(gen, mot, sep = "."))
@@
     combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_human)
+  }
+}
+```
+
+# Mouse data for liftOver
+
+```{r mm}
+for (gen in c(genome_mouse)) {
+  for (mot in "MA0139.1") {
+    print(paste(gen, mot, sep = "."))
+    combine_fimo(motif = mot, genome_id = gen, chromosomes = chromosome_mouse)
   }
@@
 # 
-files <- list.files(dir_data, pattern = "*.bed", full.names = TRUE)
-for (file in files) {
-  print(basename(file) %>% str_remove(., ".bed"))
-  mtx <- read_tsv(file, col_names = FALSE)
+# Output files
+fileNameOut1 <- file.path(dir_project, "results", "log_PWMs.csv")
+if (file.exists(fileNameOut1)) { file.remove(fileNameOut1) }
+# Defined from FIMO_EDA
+pval_threshold <- 1e-6
+files <- list.files(file.path(dir_project, "RData"), pattern = "bed", full.names = TRUE)
+for (fil in files) {
+  print(basename(fil) %>% str_remove(., ".bed"))
+  mtx <- read_tsv(fil, col_names = FALSE, col_types = c("ciicdciddc"))
   colnames(mtx) <- c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")
-  gr <- makeGRangesFromDataFrame(mtx)
+  gr <- makeGRangesFromDataFrame(mtx, keep.extra.columns = TRUE)
   # Summary
-  summary_to_save <- data.frame(File = basename(file) %>% str_remove(., ".bed"),
-                                Assembly = strsplit(basename(file), ".", fixed = TRUE)[[1]][1],
-                                All = length(gr),
-                                Reduced = length(reduce(gr)))
+  summary_to_save <- data.frame(File = basename(fil) %>% str_remove(., ".bed"),
+                                Assembly = strsplit(basename(fil), ".", fixed = TRUE)[[1]][1],
+                                `All (p-value threshold 1e-4)` = length(gr),
+                                `Reduced (p-value threshold 1e-4)` = length(reduce(gr)),
+                                `All (p-value threshold 1e-6)` = length(gr[gr$pvalue < pval_threshold]),
+                                `Reduced (p-value threshold 1e-6)` = length(reduce(gr[gr$pvalue < pval_threshold])) )
   write_csv(summary_to_save, fileNameOut1, append = TRUE)
+  gc()
 }

### 2022-07-04 manuscript/manuscript_bioinformatics.Rmd
Latest changes

@@
 
-Position Weight Matrices (PWMs) [@Stormo:2000uw] of CTCF sequence motifs from Jaspar 2022 [@Castro-Mondragon:2022uv], HOCOMOCO v11 [@Kulakovskiy:2018vw], CIS-BP [@Weirauch:2014ul], Jolma 2013 [@Jolma:2013vd], and SwissRegulon [@Pachkov:2013uk] databases were downloaded from the MEME Motif database (`Table S1`).
+Position Weight Matrices (PWMs) [@Stormo:2000uw] of CTCF sequence motifs from Jaspar 2022 [@Castro-Mondragon:2022uv], HOCOMOCO v11 [@Kulakovskiy:2018vw], Jolma 2013 [@Jolma:2013vd], and SwissRegulon [@Pachkov:2013uk] databases were downloaded from the MEME Motif database (`Table S1`).
 They were largely similar with the exception of Jolma 2013 and two long (34bp and 35bp) Jaspar PWMs (`Figure S1A`).

### 2022-07-04 03_EDA_CTCFBSDB.Rmd
Rename

--- 2022-06-06 EDA_CTCFBSDB.Rmd
@@
 # Predicted data
-fileNameIn1 <- file.path(dir_data, "data/allcomp.txt.gz")
-# download.file("https://insulatordb.uthsc.edu/download/allcomp.txt.gz", fileNameIn1)
+fileNameIn1 <- file.path(dir_project, "data/allcomp.txt.gz")
+if (!file.exists(fileNameIn1)) {
+  download.file("https://insulatordb.uthsc.edu/download/allcomp.txt.gz", fileNameIn1)
+}
 # Experimental data
-fileNameIn2 <- file.path(dir_data, "data/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz")
+fileNameIn2 <- file.path(dir_project, "data/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz")
 # download.file("https://insulatordb.uthsc.edu/download/CTCFBSDB_all_exp_sites_Sept12_2012.txt.gz", fileNameIn2)
 # Output files
-fileNameOut1 <- file.path(dir_data, "results/hg18.CTCFBSDB.CTCF_predicted_human.RData")
-fileNameOut1.1 <- file.path(dir_data, "results/hg18.CTCFBSDB.CTCF_predicted_human.bed")
-fileNameOut2 <- file.path(dir_data, "results/mm8.CTCFBSDB.CTCF_predicted_mouse.RData")
-fileNameOut2.1 <- file.path(dir_data, "results/mm8.CTCFBSDB.CTCF_predicted_mouse.bed")
-# fileNameOut3 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_experimental_human.RData")
-# fileNameOut3.1 <- file.path(dir_data, "results/hg19.CTCFBSDB.CTCF_experimental_human.bed")
-# fileNameOut4 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.RData")
-# fileNameOut4.1 <- file.path(dir_data, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.bed")
+fileNameOut1 <- file.path(dir_project, "RData", "hg18.CTCFBSDB.CTCF_predicted_human.RData")
+fileNameOut1.1 <- file.path(dir_project, "RData", "hg18.CTCFBSDB.CTCF_predicted_human.bed")
+fileNameOut2 <- file.path(dir_project, "RData", "mm8.CTCFBSDB.CTCF_predicted_mouse.RData")
+fileNameOut2.1 <- file.path(dir_project, "RData", "mm8.CTCFBSDB.CTCF_predicted_mouse.bed")
+# fileNameOut3 <- file.path(dir_project, "results/hg19.CTCFBSDB.CTCF_experimental_human.RData")
+# fileNameOut3.1 <- file.path(dir_project, "results/hg19.CTCFBSDB.CTCF_experimental_human.bed")
+# fileNameOut4 <- file.path(dir_project, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.RData")
+# fileNameOut4.1 <- file.path(dir_project, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.bed")
 ```
@@
 
+```{r eval=FALSE}
 # Load EXPERIMENTAL data
 
-```{r eval=FALSE}
 # Read in data
@@
 
+```{r eval=FALSE}
 ## EDA
 
-```{r eval=FALSE}
 # head(mtx, n = 100) %>% View()
@@
 
+```{r eval=FALSE}
 ## GRanges
 
-```{r eval=FALSE}
 mtx_to_gr <- function(mtx = mtx, org = "Rhesus") {
@@
 #   # Output files
-#   fileNameOut1 <- file.path(dir_data, paste0("results/CTCF_", i, "_CTCFBSDB_experimental.RData"))
+#   fileNameOut1 <- file.path(dir_project, paste0("results/CTCF_", i, "_CTCFBSDB_experimental.RData"))
 #   # Save as RData object. 

### 2022-07-04 05_EDA_liftOver_mm.Rmd
Rename

--- 2022-07-04 EDA_liftOver.Rmd
@@
 # Input files
-fileNameIn1 <- file.path(dir_data, "sample_annotation.csv")
+fileNameIn1 <- file.path(dir_data, "sample_annotation_mm.csv")
 # Output files
-fileNameOut1 <- file.path(dir_project, "results", "Figure_liftOverJaccard.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_liftOverJaccard_mm.svg")
 ```
@@
          clustering_distance_cols = clust_method, 
-         annotation_row = mydf, annotation_colors = list(Assembly = c(hg18 = mycols[2], hg19 = mycols[4], hg38 = mycols[5], T2T = mycols[1])),
+         annotation_row = mydf, annotation_colors = list(Assembly = c(mm9 = mycols[2], mm10 = mycols[4], mm39 = mycols[5])),
          treeheight_row = 40,

### 2022-07-04 06_FIMO_EDA.Rmd
Rename

--- 2022-07-02 FIMO_EDA.Rmd
@@
 dir_data <- "/Users/mdozmorov/Documents/Data/GoogleDrive/CTCF.dev/merlot"
+dir_project <- "/Users/mdozmorov/Documents/Work/GitHub/CTCF.dev"
 # Motifs
-motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
-motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "jolma2013", "SwissRegulon_human_and_mouse")
+motif_human  <- c("CIS_BP_2.00_Homo_sapiens", "CTCFBSDB_PWM", "HOCOMOCOv11_core_HUMAN_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "Jolma2013", "SwissRegulon_human_and_mouse")
+motif_mouse  <- c("CIS_BP_2.00_Mus_musculus", "CTCFBSDB_PWM", "HOCOMOCOv11_core_MOUSE_mono_meme_format", "JASPAR2022_CORE_vertebrates_non_redundant_v2", "Jolma2013", "SwissRegulon_human_and_mouse")
 # Genomes
 genome_human <- c("T2T", "hg38", "hg19")
-genome_mouse <- c("mm10", "mm9")
+genome_mouse <- c("mm10", "mm9", "mm39")
 ```
@@
     print(paste(gen, mot, sep = "."))
+    fileNameIn <- file.path(dir_project, paste(gen, mot, "bed", sep = "."))
     mtx <- read_tsv(fileNameIn, col_names = FALSE, col_types = c("ciicdciddc"))
-    fileNameIn <- file.path(dir_data, paste(gen, mot, "bed", sep = "."))
     print(paste("Number of regions more significant than the p-value threshold:", round(sum(mtx$X8 < pval_threshold), digits = 5) ))
@@
     print(paste(gen, mot, sep = "."))
-    fileNameIn <- file.path(dir_data, paste(gen, mot, "bed", sep = "."))
+    fileNameIn <- file.path(dir_project, "RData", paste(gen, mot, "bed", sep = "."))
     mtx <- read_tsv(fileNameIn, col_names = FALSE, col_types = c("ciicdciddc"))
@@
 
-fileNameOut1 <- file.path(dir_data, "Figure_human_pvalues.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_human_pvalues.svg")
 ggsave(filename = fileNameOut1, width = 6.4, height = 1.6)
@@
 
-fileNameOut1 <- file.path(dir_data, "Figure_mouse_pvalues.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_mouse_pvalues.svg")
 # ggsave(filename = fileNameOut1, width = 7, height = 3)
 dev.off()
-svg(filename = fileNameOut1, width = 7, height = 3)
+svg(filename = fileNameOut1, width = 6.4, height = 1.6)
 print(p)

### 2022-07-04 07_CTCF_Threshold.Rmd
Rename

--- 2022-07-04 CTCF_Threshold_mm.Rmd
@@
 # Input files
-fileNameIn1 <- file.path(dir_project, "RData", "mm10.MA0139.1.bed")
-fileNameIn2 <- file.path(dir_project, "data", "mm10-CTCF.bed")
+fileNameIn1 <- file.path(dir_project, "RData", "hg38.MA0139.1.bed")
+fileNameIn2 <- file.path(dir_project, "data", "GRCh38-CTCF.bed")
 # Output files
-fileNameOut1 <- file.path(dir_project, "results", "Figure_mouse_pvalues_threshold.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_human_pvalues_threshold.svg")
 ```

### 2022-07-04 07_CTCF_Threshold_mm.Rmd
Fix plotting

--- 2022-07-04 CTCF_Threshold_mm.Rmd
@@
 mtx_to_plot <- pivot_longer(gr_summary, cols = c("TP", "FP"), names_to = "Type")
-ggplot(mtx_to_plot, aes(x = Threshold, y = value, color = Type)) +
+p <- ggplot(mtx_to_plot, aes(x = Threshold, y = value, color = Type)) +
   geom_line() +
@@
 
-ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
-# dev.off()
-# svg(filename = fileNameOut1, width = 7, height = 3)
-# print(p)
-# dev.off()
+# ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
+dev.off()
+svg(filename = fileNameOut1, width = 4.5, height = 1.5)
+print(p)
+dev.off()
 

### 2022-07-05 03_EDA_CTCFBSDB.Rmd
Add liftOver

@@
 fileNameOut2.1 <- file.path(dir_project, "RData", "mm8.CTCFBSDB.CTCF_predicted_mouse.bed")
-# fileNameOut3 <- file.path(dir_project, "results/hg19.CTCFBSDB.CTCF_experimental_human.RData")
-# fileNameOut3.1 <- file.path(dir_project, "results/hg19.CTCFBSDB.CTCF_experimental_human.bed")
-# fileNameOut4 <- file.path(dir_project, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.RData")
-# fileNameOut4.1 <- file.path(dir_project, "results/mm9.CTCFBSDB.CTCF_experimental_mouse.bed")
+# liftOver human
+fileNameOut3 <- file.path(dir_project, "RData", "hg19.CTCFBSDB.CTCF_predicted_human.RData")
+fileNameOut3.1 <- file.path(dir_project, "RData", "hg19.CTCFBSDB.CTCF_predicted_human.bed")
+fileNameOut4 <- file.path(dir_project, "RData", "hg38.CTCFBSDB.CTCF_predicted_human.RData")
+fileNameOut4.1 <- file.path(dir_project, "RData", "hg38.CTCFBSDB.CTCF_predicted_human.bed")
+# liftOver mouse
+fileNameOut5 <- file.path(dir_project, "RData", "mm9.CTCFBSDB.CTCF_predicted_mouse.RData")
+fileNameOut5.1 <- file.path(dir_project, "RData", "mm9.CTCFBSDB.CTCF_predicted_mouse.bed")
+fileNameOut6 <- file.path(dir_project, "RData", "mm10.CTCFBSDB.CTCF_predicted_mouse.RData")
+fileNameOut6.1 <- file.path(dir_project, "RData", "mm10.CTCFBSDB.CTCF_predicted_mouse.bed")
 ```
@@
 export.bed(mm8.CTCFBSDB.CTCF_predicted_mouse, fileNameOut2.1)
+```
+
+# LiftOver
+
+```{r}
+# Function to download a chain for the gf18_to_XX conversion and create a GRanges object
+liftOver_custom <- function(URL = "https://hgdownload.cse.ucsc.edu/goldenpath/hg18/liftOver/hg18ToHg19.over.chain.gz", gr_in = hg18.CTCFBSDB.CTCF_predicted_human, genome_id = "hg19"){
+  fileChain <- file.path(dir_project, "data", basename(URL))
+  if (!file.exists(str_remove(fileChain, ".gz"))) {
+    download.file(URL, destfile = fileChain)
+    R.utils::gunzip(fileChain)
+  }
+
+  ch <- import.chain(str_remove(fileChain, ".gz"))
+  gr_converted <- liftOver(gr_in, ch) %>% unlist()
+  # Get chromosome info and match it to the chromosome order in ctcfBED
+  chrom_data <- GenomeInfoDb::getChromInfoFromUCSC(genome = genome_id)
+  # Subset to common chromosomes
+  common_chromosomes <- intersect(chrom_data$chrom, seqlevels(gr_converted))
+  chrom_data <- chrom_data[chrom_data$chrom %in% common_chromosomes, ]
+  gr_converted <- keepSeqlevels(gr_converted, common_chromosomes, pruning.mode = "tidy")
+  # Match order
+  chrom_data <- chrom_data[match(seqlevels(gr_converted), chrom_data$chrom), ]
+  # Check if chromosome order is the same
+  if (!all.equal(seqlevels(gr_converted), chrom_data$chrom)) {
+    print(paste("Chromosome order does not match for", genome_id, "genome."))
+    break
+  }
+  # Assign seqinfo data
+  seqlengths(gr_converted) <- chrom_data$size
+  isCircular(gr_converted) <- chrom_data$circular
+  genome(gr_converted)     <- genome_id
+  return(gr_converted)  
+}
+
+# hg18.CTCFBSDB.CTCF_predicted_human to hg19 conversion
+gr_out <- "hg19.CTCFBSDB.CTCF_predicted_human"
+# Assign the converted GRanges to the variable name
+assign(gr_out, liftOver_custom(URL = "https://hgdownload.cse.ucsc.edu/goldenpath/hg18/liftOver/hg18ToHg19.over.chain.gz", gr_in = hg18.CTCFBSDB.CTCF_predicted_human, genome_id = "hg19"))
+# Save as RData object. subdir is the character name of the CTCF GRanges variable
+save(list = gr_out, file = fileNameOut3)
+# load(fileNameOut4)
+# Save BED files
+export.bed(hg19.CTCFBSDB.CTCF_predicted_human, fileNameOut3.1)
+
+# hg18.CTCFBSDB.CTCF_predicted_human to hg38 conversion
+gr_out <- "hg38.CTCFBSDB.CTCF_predicted_human"
+# Assign the converted GRanges to the variable name
+assign(gr_out, liftOver_custom(URL = "https://hgdownload.cse.ucsc.edu/goldenpath/hg18/liftOver/hg18ToHg38.over.chain.gz", gr_in = hg18.CTCFBSDB.CTCF_predicted_human, genome_id = "hg38"))
+# Save as RData object. subdir is the character name of the CTCF GRanges variable
+save(list = gr_out, file = fileNameOut4)
+# load(fileNameOut4)
+# Save BED files
+export.bed(hg38.CTCFBSDB.CTCF_predicted_human, fileNameOut4.1)
+
+# mm8.CTCFBSDB.CTCF_predicted_mouse to mm9 conversion
+gr_out <- "mm9.CTCFBSDB.CTCF_predicted_mouse"
+# Assign the converted GRanges to the variable name
+assign(gr_out, liftOver_custom(URL = "https://hgdownload.cse.ucsc.edu/goldenpath/mm8/liftOver/mm8ToMm9.over.chain.gz", gr_in = mm8.CTCFBSDB.CTCF_predicted_mouse, genome_id = "mm9"))
+# Save as RData object. subdir is the character name of the CTCF GRanges variable
+save(list = gr_out, file = fileNameOut5)
+# load(fileNameOut4)
+# Save BED files
+export.bed(mm9.CTCFBSDB.CTCF_predicted_mouse, fileNameOut5.1)
+
+# mm8.CTCFBSDB.CTCF_predicted_mouse to mm10 conversion
+gr_out <- "mm10.CTCFBSDB.CTCF_predicted_mouse"
+# Assign the converted GRanges to the variable name
+assign(gr_out, liftOver_custom(URL = "https://hgdownload.cse.ucsc.edu/goldenpath/mm8/liftOver/mm8ToMm10.over.chain.gz", gr_in = mm8.CTCFBSDB.CTCF_predicted_mouse, genome_id = "mm10"))
+# Save as RData object. subdir is the character name of the CTCF GRanges variable
+save(list = gr_out, file = fileNameOut6)
+# load(fileNameOut4)
+# Save BED files
+export.bed(mm9.CTCFBSDB.CTCF_predicted_mouse, fileNameOut6.1)
 ```

### 2022-07-05 manuscript/manuscript_bioinformatics.Rmd
Version sent to authors

@@
 # csl: /Volumes/GoogleDrive/My Drive/manuscript_template/styles.ref/cell-reports.csl
-csl: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.ref/cell-reports.csl
+csl: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/styles.ref/bioinformatics.csl
 output: 
@@
 
-# CTCF: A Bioconductor data package of human and mouse CTCF binding sites
+# CTCF: An R/Bioconductor data package of human and mouse CTCF binding sites
 
-Mikhail Dozmorov*, Wancen Mu, Eric Davis, Stuart Lee, Tim Triche, Douglas Phanstiel, Michael Love
+Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric Davis^5^, Stuart Lee^13,14^, Tim Triche^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael Love^3,4^
+
+^1^Department of Biostatistics, Virginia Commonwealth University, Richmond, VA, 23298, USA.
+
+^2^Department of Pathology, Virginia Commonwealth University, Richmond, VA, 23284, USA.
+
+^3^Department of Biostatistics, University of North Carolina-Chapel Hill, Chapel Hill, NC 27514, USA
+
+^4^Department of Genetics, University of North Carolina-Chapel Hill, Chapel Hill, NC 27514, USA
+
+^5^Curriculum in Bioinformatics and Computational Biology, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA
+
+^6^Thurston Arthritis Research Center, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA
+
+^7^Department of Cell Biology and Physiology, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA
+
+^8^Lineberger Comprehensive Cancer Center, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA
+
+^9^Curriculum in Genetics and Molecular Biology, University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA
+
+^10^Center for Epigenetics, Van Andel Research Institute, Grand Rapids, Michigan, United States of America
+
+^11^Department of Pediatrics, College of Human Medicine, Michigan State University, East Lansing, Michigan, United States of America
+
+^12^Department of Translational Genomics, Keck School of Medicine, University of Southern California, Los Angeles, California, United States of America
+
+^13^Department of Econometrics and Business Statistics, Monash University, Clayton, Australia
+
+^14^Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia
+
+^\*^To whom correspondence should be addressed: mdozmorov@vcu.edu
 
@@
 
-### Keywords
-
 # Introduction
 
-The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator <!--evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux].
-Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including -->involved in transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, <!--intra- and inter-chromosomal contacts -->chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy].
+The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux].
+<!--Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv],--> Tt has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy].
 <!--CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs].
@@
 binding motif length varying between 20 and 40bp -->
-CTCF binds DNA through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Ohlsson:2001wu; @Jolma:2013vd]<!--@Filippova:1996tv-->.
+CTCF binds DNA through the combinatorial use of its 11-zinc finger domains to target sites with remarkable sequence variation [@Ohlsson:2001wu; @Jolma:2013vd]<!--@Filippova:1996tv-->.
 A 15-20bp CTCF consensus motif from the ChIP-seq data analysis has been defined [@Kim:2007ux], referred to hereafter as M1.
@@
 
-CTCF binding profiles between different tissues exhibit substantial concordance [@Kim:2007ux].<!--For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].--> We aimed to provide uniformly detected strand-specific CTCF binding sites for the most studied human and mouse genomes.
+CTCF binding is generally conserved between different tissues [@Kim:2007ux].<!--For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].--> We aimed to provide uniformly detected strand-specific CTCF binding sites for human and mouse genomes, generally applicable for any cell type.
 Given the differences in genome assemblies, we defined CTCF binding sites directly for each assembly, including the Telomere to Telomere (T2T) human genome assembly [@Nurk:2022up] and the GRCm39/mm39 mouse genome assembly.
 We demonstrate that coordinate conversion<!--of CTCF binding sites between genome assemblies--> (liftOver) is a viable way to obtain assembly-specific CTCF binding sites.
-We further demonstrate the need for filtering less significant CTCF binding sites.<!--Given the importance of CTCF binding in convergent orientation for forming chromatin loops [@Rao:2014aa; @Nanni:2020aa; @Vietri-Rudan:2015tu], we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding.--> We also include selected experimental CTCF binding sites<!--to provide a comprehensive general-purpose collection of CTCF binding profiles.--> and assemble them in the CTCF R/Bioconductor package.
+We further demonstrate the need to filter less significant CTCF binding sites and merge overlapping sites.<!--Given the importance of CTCF binding in convergent orientation for forming chromatin loops [@Rao:2014aa; @Nanni:2020aa; @Vietri-Rudan:2015tu], we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding.--> We also include selected experimental CTCF binding sites<!--to provide a comprehensive general-purpose collection of CTCF binding profiles.--> and assemble them in the CTCF R/Bioconductor package.
 
@@
 Position Weight Matrices (PWMs) [@Stormo:2000uw] of CTCF sequence motifs from Jaspar 2022 [@Castro-Mondragon:2022uv], HOCOMOCO v11 [@Kulakovskiy:2018vw], Jolma 2013 [@Jolma:2013vd], and SwissRegulon [@Pachkov:2013uk] databases were downloaded from the MEME Motif database (`Table S1`).
-They were largely similar with the exception of Jolma 2013 and two long (34bp and 35bp) Jaspar PWMs (`Figure S1A`).
+They were largely similar except for Jolma 2013 and two long (34bp and 35bp) Jaspar PWMs (`Figure S1A`).
 The CTCF binding site database (CTCFBSDB) [@Ziebarth:2013vj] includes PWMs for the M1+M2 motifs [@Schmidt:2012uk], the Ren_20 motif [@Kim:2007ux], and the LM2, LM7, and LM23 motifs [@Xie:2007wm] (`Figure S1B`).
@@
 database has not been updated since 2013 and the predictions are available for hg18 and mm8 genome assemblies only--> (`Table S2`).
-As coordinate conversion (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts.
-Expectedly, genome assembly was the primary driver of similarity (`Figure 1A`).<!--among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies.
+As coordinate conversion (liftOver) is a well-established practice, we investigated how CTCF sites detected in one genome assembly compare with those converted from another.
+Jaccard overlap among CTCF sites aligned or lifted over to hg18, hg19, hg38, and T2T genomes showed that genome assembly was the primary driver of similarity (`Figure 1A`).<!--among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies.
 liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table S2`).
 The mm10/mm9-detected CTCF binding sites also demonstrated good overlap between the originally detected and liftOver versions (`Figure S2A`).
-These results --> These results were observed using mm9/mm10/mm39 genome assemblies (`Figure S2A`), indicating that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly.
+These results --> These results were also observed using mm9/mm10/mm39 genome assemblies (`Figure S2A`), indicating that liftOver is a viable option to obtain CTCF sites in the required genome assembly.
 Using liftOver, we provide hg19/hg38 and mm9/mm10 versions of predicted CTCF sites from the CTCFBSDB database.
 
-The FIMO tool <!--as one of the oldest and most well-known tool for motif scanning.
-The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj].
--->uses the default p-value cutoff 1E-4 for CTCF motif detection.
-However, we observed that up to 98% of motifs have a detection p-value more than 1E-6 (`Figure 1B`, `Figure S2B`), suggesting most CTCF motifs have a relatively poor sequence match.
+The FIMO tool is one of the oldest and most well-known tool for motif scanning.
+It uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values [@Cuellar-Partida:2012wj], with the default p-value cutoff 1E-4.
+However, we observed that up to 98% of CTCF sites are detected within the 1E-4 - 1E-6 p-value range (`Figure 1B`, `Figure S2B`), suggesting most CTCF sites have a relatively poor sequence match.
 Using the ENCODE SCREEN database of CTCF-bound cis-regulatory elements (cCREs) [@ENCODE-Project-Consortium:2020uq] as ground truth (`Table S2`, included in the package), we quantified the proportion of true and false positives depending on the FIMO p-value cutoff.
-We found 1E-6 threshold as the most optimal, providing $\sim80\%$ of true positive CTCF motifs.
+We found the 1E-6 threshold the most optimal, providing $\sim80\%$ of true positive CTCF motifs (`Figure 1C`, `Figure S2C`).
 Additionally, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S3`).
-Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering yields the number of CTCF sites comparable to previously reported.
+Reducing them (merging overlapping CTCF sites), combined with 1E-6 cutoff filtering, yields the number of CTCF sites comparable to previously reported ones.
 Although we provide FIMO results using the default settings, we recommend filtering and reducing them.
@@
 Our goal was to provide easy programmatic access to CTCF binding data applicable to any cell or tissue type.
-The CTCF package contains XX GRanges objects of strand-specific CTCF motifs predicted using 98 PWMs from six databases on human and mouse genome assemblies, including the T2T and mm39 assemblies (`Table S3`).
-Given the popularity of JASPAR database, we recommend using predictions made with the MA0139.1 PWM.
-If experimentally detected CTCF motifs are of interest, we suggest using CTCF-bound cCREs from the ENCODE SCREEN database.
+The CTCF package contains 43 GRanges objects of strand-specific CTCF motifs predicted using 98 PWMs from six databases on human and mouse genome assemblies, including the T2T and mm39 assemblies (`Table S3`).
+Given the popularity of the JASPAR database, we recommend using predictions made with the MA0139.1 PWM, selecting sites detected at 1E-6 p-value cutoff, and merging overlapping sites.
+If experimentally detected CTCF sites are of interest, we suggest using CTCF-bound cCREs from the ENCODE SCREEN database.
 We hope the CTCF package will enable standardization and reproducibility of studies employing CTCF binding data.

### 2022-07-05 manuscript/manuscript_bioinformatics.Rmd
Minor tweaks

@@
 Our goal was to provide easy programmatic access to CTCF binding data applicable to any cell or tissue type.
-The CTCF package contains 43 GRanges objects of strand-specific CTCF motifs predicted using 98 PWMs from six databases on human and mouse genome assemblies, including the T2T and mm39 assemblies (`Table S3`).
+The CTCF package contains 51 GRanges objects of strand-specific CTCF motifs predicted using 98 PWMs from six databases on human and mouse genome assemblies, including the T2T and mm39 assemblies (`Table S3`).
 Given the popularity of the JASPAR database, we recommend using predictions made with the MA0139.1 PWM, selecting sites detected at 1E-6 p-value cutoff, and merging overlapping sites.

### 2022-07-19 manuscript/manuscript_bioinformatics.Rmd
Add Eric's initial
Add Mike's comments

@@
 
-Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric Davis^5^, Stuart Lee^13,14^, Tim Triche^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael Love^3,4^
+Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric S. Davis^5^, Stuart Lee^13,14^, Tim Triche^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael I. Love^3,4^
 
@@
 The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux].
-<!--Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv],--> Tt has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy].
+<!--Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv],--> It has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy].
 <!--CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs].
@@
 However, we observed that up to 98% of CTCF sites are detected within the 1E-4 - 1E-6 p-value range (`Figure 1B`, `Figure S2B`), suggesting most CTCF sites have a relatively poor sequence match.
-Using the ENCODE SCREEN database of CTCF-bound cis-regulatory elements (cCREs) [@ENCODE-Project-Consortium:2020uq] as ground truth (`Table S2`, included in the package), we quantified the proportion of true and false positives depending on the FIMO p-value cutoff.
+Using the ENCODE SCREEN V3 database of CTCF-bound cis-regulatory elements (cCREs) [@ENCODE-Project-Consortium:2020uq] as ground truth (`Table S2`, included in the package), we quantified the proportion of true and false positives depending on the FIMO p-value cutoff.
 We found the 1E-6 threshold the most optimal, providing $\sim80\%$ of true positive CTCF motifs (`Figure 1C`, `Figure S2C`).

### 2022-07-19 scripts/downloads.sh
Add fimo download and processing scripts

@@ -1,0 +1,40 @@
+# Downloaded May 20, 2022
+
+# MEME Motif Databases (updated 20 Mar 2022)
+# https://meme-suite.org/meme/doc/download.html
+wget https://meme-suite.org/meme/meme-software/Databases/motifs/motif_databases.12.23.tgz
+# Extract
+tar zxvfh motif_databases.12.23.tgz
+# Go there
+cd motif_databases
+# File to store found CTCF motifs
+FILEOUT=CTCF_motifs.txt
+# Find CTCF motif occurrences, output in the file
+for file in `find . -type f -name "*.meme"`; do 
+	# Output only if CTCF, not CTCFL is found
+	if grep -iv CTCFL $file | grep -iq CTCF; then
+		echo $file >> ${FILEOUT}; 
+		grep -iv CTCFL $file | grep -i CTCF >> ${FILEOUT};
+	fi;
+done
+
+# Find CTCF motif occurrences, output on screen
+for file in `find . -type f -name "*.meme"`; do 
+	# Output only if CTCF, not CTCFL is found
+	if grep -iv CTCFL $file | grep -iq CTCF; then
+		echo $file; 
+		grep -iv CTCFL $file | grep -i CTCF;
+	fi;
+done
+
+# CTCFBSDB PWM matrix
+wget https://insulatordb.uthsc.edu/download/CTCFBSDB_PWM.mat
+transfac2meme -use_acc CTCFBSDB_PWM.mat > CTCFBSDB_PWM.meme
+
+
+# T2T, https://github.com/marbl/CHM13
+# Downloaded May 20, 2022
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/T2T/CHM13/assemblies/analysis_set/chm13v2.0.fa.gz
+mv chm13v2.0.fa.gz /home/sequencing/data/ExtData/UCSC/T2T/
+gzip -d chm13v2.0.fa.gz
+

### 2022-07-20 manuscript/manuscript_bioinformatics.Rmd
Add Tim's notes

@@
 
-Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric S. Davis^5^, Stuart Lee^13,14^, Tim Triche^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael I. Love^3,4^
+Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric S. Davis^5^, Stuart Lee^13,14^, Timothy J. Triche, Jr.^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael I. Love^3,4^
 
@@
 
-**Summary:** CTCF (CCCTC-binding factor) is considered the master regulator of the genome's structure and function.
-The diversity of CTCF binding motifs leads to the lack of a unified source of CTCF binding data.
-We collected position weight matrices of CTCF binding motifs and defined strand-oriented CTCF binding sites in the human and mouse genome assemblies, including the Telomere to Telomere human and mm39 mouse genomes.
-We also included selected experimentally determined and predicted CTCF binding sites, such as CTCF-bound cis-regulatory element from SCREEN ENCODE.
-We recommend filtering less significant CTCF binding motifs and demonstrate that liftOver is a viable option to convert CTCF coordinates between genome assemblies.
-Our comprehensive data resource and usage recommendations will strengthen reproducibility of genomic studies utilizing CTCF binding data.
+**Summary:** CTCF (CCCTC-binding factor) is an 11-zinc-finger DNA binding protein which regulates much of the eukaryotic genome’s 3-dimensional structure and function.
+The diversity of CTCF binding motifs has led to a fragmented landscape of CTCF binding data.
+We collected position weight matrices of CTCF binding motifs and defined strand-oriented CTCF binding sites in the human and mouse genomes, including the recent Telomere to Telomere (T2T) and mm39 assemblies.
+We included selected experimentally determined and predicted CTCF binding sites, such as CTCF-bound cis-regulatory elements from SCREEN ENCODE.
+We recommend filtering strategies for CTCF binding motifs and demonstrate that liftOver is a viable alternative to convert CTCF coordinates between assemblies.
+Our comprehensive data resource and usage recommendations can serve to harmonize and strengthen reproducibility of genomic studies utilizing CTCF binding data.
 

### 2022-07-20 02_EDA_SCREEN.Rmd
More comments

--- 2022-07-04 EDA_SCREEN.Rmd
@@
 ```{r}
+# Function to convert SCREEN data to GRanges
 mtx_to_gr <- function(mtx = mtx_human, genome_id = "hg38") {
@@
 ```{r}
+# Convert SCREEN data to GRanges
 hg38.SCREEN.GRCh38_CTCF <- mtx_to_gr(mtx = mtx_human, genome_id = "hg38")

### 2022-07-20 03_EDA_CTCFBSDB.Rmd
More comments

@@
 ```{r}
+# Function to convert CTCFBSDB data to GRanges
 mtx_to_gr <- function(mtx = mtx_human, genome_id = "hg19") {
@@
 }
-
+# Convert CTCFBSDB data to GRanges
 hg18.CTCFBSDB.CTCF_predicted_human <- mtx_to_gr(mtx = mtx_human, genome_id = "hg18")
@@
 ```{r}
-# Function to download a chain for the gf18_to_XX conversion and create a GRanges object
+# Function to LiftOver hg18-hg19-hg38, mm8-mm9-mm10
+# Downloads a chain for the gf18_to_XX conversion and create a GRanges object
 liftOver_custom <- function(URL = "https://hgdownload.cse.ucsc.edu/goldenpath/hg18/liftOver/hg18ToHg19.over.chain.gz", gr_in = hg18.CTCFBSDB.CTCF_predicted_human, genome_id = "hg19"){

### 2022-11-14 04_FIMO_processing.Rmd
CTCF Overlap analysis

--- 2022-07-04 FIMO_processing.Rmd
@@
 library(rtracklayer)
+library(GenomicDistributions)
 ```
@@
 pval_threshold <- 1e-6
-files <- list.files(file.path(dir_project, "RData"), pattern = "bed", full.names = TRUE)
+files <- list.files(file.path(dir_project, "RData", "BEDfiles"), pattern = "bed", full.names = TRUE) %>% sort
 for (fil in files) {
@@
   gr <- makeGRangesFromDataFrame(mtx, keep.extra.columns = TRUE)
+  # Width distribution
+  gr_width <- width(gr)
+  width_distribution <- paste(formatC(min(gr_width), format = "f", digits = 0), 
+                              formatC(mean(gr_width), format = "f", digits = 2), 
+                              formatC(max(gr_width), format = "f", digits = 0), sep = "/")
+  # Distance to nearby
+  gr_nearest <- calcNearestNeighbors(gr)
+  nearest_distribution <- paste(formatC(min(gr_nearest), format = "f", digits = 0), 
+                              formatC(mean(gr_nearest), format = "f", digits = 2), 
+                              formatC(max(gr_nearest), format = "f", digits = 0), sep = "/")
+  # # Proportion of overlapping enrties
+  # tmp <- findOverlaps(gr, gr)
+  # prop_overlapping <- sort(table(queryHits(tmp)) %>% table(), decreasing = TRUE)
+  # prop_overlapping <- sum(prop_overlapping[which(names(prop_overlapping) != "1")]) / length(gr)
   # Summary
@@
                                 Assembly = strsplit(basename(fil), ".", fixed = TRUE)[[1]][1],
+                                `Width, bp (min/mean/max)` = width_distribution,
+                                `Distance to nearest, bp (min/mean/max)` = nearest_distribution,
+#                                 `Proportion of overlapping` = formatC(prop_overlapping, format = "f", digits = 3),
                                 `All (p-value threshold 1e-4)` = length(gr),
@@
 ```
+
+# Overlap
+
+```{r}
+files <- list.files(file.path(dir_project, "RData", "BEDfiles"), pattern = "bed", full.names = TRUE) %>% sort
+# Vector to store proportion of overlapping sites
+prop_overlapping <- vector(mode = "numeric", length = length(files))
+for (i in 1:length(files)) {
+  fil <- files[i]
+  print(basename(fil) %>% str_remove(., ".bed"))
+  # Process all but CIS_BP data
+  if (!grepl("CIS_BP", fil)) {
+    mtx <- read_tsv(fil, col_names = FALSE, col_types = c("ciicdciddc"))
+    colnames(mtx) <- c("seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence")
+    gr <- makeGRangesFromDataFrame(mtx, keep.extra.columns = TRUE)
+    # Proportion of overlapping enrties
+    tmp <- findOverlaps(gr, gr)
+    prop_overlap <- sort(table(queryHits(tmp)) %>% table(), decreasing = TRUE)
+    prop_overlapping[i] <- sum(prop_overlap[which(names(prop_overlap) != "1")]) / length(gr)
+  } else {
+    prop_overlapping[i] <- NA
+  }
+}
+prop_overlap_df <- data.frame(Data = basename(files) %>% str_remove(., ".bed"),
+                              Prop_overlap = prop_overlapping)
+```

### 2022-11-14 07_CTCF_Threshold.Rmd
Overlap with cell specific CTCF sites analysis

--- 2022-07-04 CTCF_Threshold_mm.Rmd
@@
 library(GenomicRanges)
+library(rtracklayer)
+library(plyranges)
 ```
@@
 # Input files
-fileNameIn1 <- file.path(dir_project, "RData", "mm10.MA0139.1.bed")
-fileNameIn2 <- file.path(dir_project, "data", "mm10-CTCF.bed")
+fileNameIn1 <- file.path(dir_project, "RData", "BEDfiles", "hg38.MA0139.1.bed")
+fileNameIn2 <- file.path(dir_project, "data", "GRCh38-CTCF.bed")
+fileNameIn3 <- file.path(dir_project, "data", "UCSC_CTCF.tsv")
 # Output files
-fileNameOut1 <- file.path(dir_project, "results", "Figure_mouse_pvalues_threshold.svg")
+fileNameOut1 <- file.path(dir_project, "results", "Figure_human_pvalues_threshold.svg")
+fileNameOut2 <- file.path(dir_project, "results", "overlap_meme_below.csv")
+fileNameOut3 <- file.path(dir_project, "results", "overlap_meme_above.csv")
 ```
@@
 # dev.off()
+```
 
+## Overlap with cell-specific sites
+
+```{r}
+# Separate FIMO CTCF peaks into above and below the threshold
+pval_threshold <- 1E-6
+gr_meme_above <- gr_meme %>% filter(pval <= pval_threshold) %>% keepStandardChromosomes(., pruning.mode = "tidy") %>% sort()
+gr_meme_below <- gr_meme %>% filter(pval >  pval_threshold) %>% keepStandardChromosomes(., pruning.mode = "tidy") %>% sort()
+print(paste("Number of CTCFs above", pval_threshold, "p-value threshold", length(gr_meme_above)))
+print(paste("Number of CTCFs below", pval_threshold, "p-value threshold", length(gr_meme_below)))
+# Get all cell type-specific CTCF sites
+# IDs of all CTCF experiments
+mtx <- read_tsv(fileNameIn3, col_names = FALSE)
+# Session for UCSC broser
+mySession <- browserSession()
+genome_id <- "hg38"
+genome(mySession) <- genome_id
+# Overlap counts
+overlap_meme_above <- cbind()
+overlap_meme_below <- cbind()
+for (i in 158:nrow(mtx)) {
+  print(paste(i, "out of", nrow(mtx)))
+  table_name <- paste0("encTfChipPk", mtx[i, "X4"])
+  query <- ucscTableQuery(mySession, table = table_name)
+  gr_ctcf <- getTable(query) %>% makeGRangesFromDataFrame(.) %>% keepStandardChromosomes(., pruning.mode = "tidy") %>% sort()
+  overlap_meme_below <- cbind(overlap_meme_below, countOverlaps(gr_meme_below, gr_ctcf))
+  overlap_meme_above <- cbind(overlap_meme_above, countOverlaps(gr_meme_above, gr_ctcf))
+  gc()
+  write_csv(as.data.frame(overlap_meme_below), fileNameOut2)
+  write_csv(as.data.frame(overlap_meme_above), fileNameOut3)
+}
+
+overlap_meme_below <- read_csv(paste0(fileNameOut2, ".gz"))
+dim(overlap_meme_below)
+overlap_meme_above <- read_csv(paste0(fileNameOut3, ".gz"))
+dim(overlap_meme_above)
+
+# The proportion of CTCF sites detected at p-value > 1E-6 overlapping at least one experimentally detected cell type-specific CTCF sites was XX +/-YY.
+# In contrast, this number for CTCF sites detected at p-value <= 1E-6 was XX+/-YY, suggesting higher proportion of false positives among less significant CTCF sites.
+overlap_meme_above_summary <- rowSums(overlap_meme_above)
+overlap_meme_below_summary <- rowSums(overlap_meme_below)
+print(paste("Average number of times less significant CTCF sites overlap with experimental CTCF sites", mean(overlap_meme_below_summary), "+/-", sd(overlap_meme_below_summary)))
+# [1] "Average number of times less significant CTCF sites overlap with experimental CTCF sites 7.43024371211658 +/- 29.2354917718536"
+print(paste("Average number of times more significant CTCF sites overlap with experimental CTCF sites", mean(overlap_meme_above_summary), "+/-", sd(overlap_meme_above_summary)))
+# [1] "Average number of times more significant CTCF sites overlap with experimental CTCF sites 120.151631212219 +/- 81.188988520227"
+
+# Among less significant CTCF sites, XX% did not overlap any cell type-specific CTCF sites.
+# In contrast, only YY% more significant sites overlap
+print(paste("The proportion of less significant CTCF sites overlapping at least one experimental CTCF site", 1 - (sum(overlap_meme_below_summary == 0) / length(overlap_meme_below_summary)) ))
+# [1] "The proportion of less significant CTCF sites overlapping at least one experimental CTCF site 0.218351650508075"
+print(paste("The proportion of more significant CTCF sites overlapping at least one experimental CTCF site", 1 - (sum(overlap_meme_above_summary == 0) / length(overlap_meme_above_summary)) ))
+# [1] "The proportion of more significant CTCF sites overlapping at least one experimental CTCF site 0.888007013981819"
+
+# P-value of the differences
+res <- wilcox.test(overlap_meme_above_summary, overlap_meme_below_summary)
+res$p.value
 ```
